---
title: "Modelling Diabetes Health Indicators"
subtitle: "MA7442 Modelling Data — Assessment 1 (Task 2)"
author: "Shahmeer Shahzad"
date: "2026-03-02"
format:
  pdf:
    geometry:
      - top=2cm
      - bottom=2cm
      - left=1.5cm
      - right=1.5cm
    documentclass: article
    df-print: kable
---

```{r}
#| label: Loading_Packages_and_workbook_Customization
#| message: false
#| warning: false
#| include: false

# Plot format and dimension
knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(fig.width = 14, fig.height = 7)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Packages
library(tidyverse)
library(skimr)
library(scales)
library(patchwork)
library(viridis)
library(rcompanion)

# Set global ggplot theme
theme_set(
  theme_classic(base_size = 16) +
    theme(
      plot.title    = element_text(size = 20, face = "bold"),
      axis.title    = element_text(size = 18),
      axis.text     = element_text(size = 14),
      legend.title  = element_text(size = 13),
      legend.text   = element_text(size = 12),
      strip.text    = element_text(face = "bold", size = 14),
      legend.background = element_rect(
        colour    = "black",
        fill      = "white",
        linewidth = 0.5
      )
    )
)
```

```{r}
#| label: data-import-and-recode
#| message: false
#| warning: false
#| include: false

diab_raw <- read_csv("dataset_1.csv")

diab <- diab_raw |>
  mutate(
    Diabetes = factor(
      Diabetes_012,
      levels = c(0, 1, 2),
      labels = c("No diabetes", "Prediabetes", "Diabetes")
    ),
    across(
      c(CholCheck, Smoker, Stroke, HeartDiseaseorAttack,
        PhysActivity, Fruits, Veggies, HvyAlcoholConsump,
        AnyHealthcare, NoDocbcCost, DiffWalk),
      function(x) factor(x, levels = c(0, 1), labels = c("No", "Yes"))
    ),
    across(
      c(HighBP, HighChol),
      function(x) factor(x, levels = c(0, 1), labels = c("Not High", "High"))
    ),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    GenHlth = factor(
      GenHlth,
      levels  = 1:5,
      labels  = c("Excellent", "Very good", "Good", "Fair", "Poor"),
      ordered = TRUE
    ),
    Age = factor(
      Age,
      levels  = 1:13,
      labels  = c("18–24", "25–29", "30–34", "35–39", "40–44",
                  "45–49", "50–54", "55–59", "60–64", "65–69",
                  "70–74", "75–79", "80+"),
      ordered = TRUE
    ),
    Education = factor(
      Education,
      levels  = 1:6,
      labels  = c("Kindergarten or less", "Grades 1–8",
                  "Grades 9–11", "High school graduate",
                  "Some college / technical", "College graduate"),
      ordered = TRUE
    ),
    Income = factor(
      Income,
      levels  = 1:8,
      labels  = c("< $10k", "$10k–$14k", "$15k–$19k", "$20k–$24k",
                  "$25k–$34k", "$35k–$49k", "$50k–$74k", "$75k+"),
      ordered = TRUE
    )
  ) |>
  select(-Diabetes_012)
```

# 1. Introduction and Reflection on Task 1 Feedback

This report constitutes Task 2 of the MA7442 Modelling Data Assessment. It builds directly on the
exploratory data analysis (EDA) and research questions developed in Task 1, incorporating the
feedback received on that submission.

## 1.1 Reflection on Task 1 Feedback

Three areas of feedback were identified for improvement.

**Scope of the EDA.** The Task 1 EDA was effective in characterising the dataset but considered
associations exclusively in relation to diabetes status as the outcome. The feedback noted that
this narrowed the research questions — all three were oriented towards predicting diabetes — and
that an additional EDA section exploring non-diabetes-focussed associations would strengthen the
breadth of the analysis and support more varied research questions. This has been addressed in
Section 3 of this report, which introduces a dedicated EDA block examining general health,
mental health burden, and physical activity as outcomes in their own right.

**Research question overlap.** As a consequence of the diabetes-only EDA, the three research
questions were noted as quite similar in focus. To address this, the research questions have been
restructured: RQ1 is retained (cardiometabolic associations with diabetes status), the previous
RQ2 and RQ3 have been consolidated into a single adjusted modelling question (new RQ2), and
two new non-diabetes-focussed research questions (RQ3 and RQ4) have been developed from the
additional EDA presented here.

**Figure line type.** In the mean unhealthy days by age group plot (Section 3.3.4 of Task 1),
`geom_line` was used to connect points across age group categories. Since age group is an ordered
factor rather than a continuous variable, connecting points implies a linear interpolation that
cannot be justified. This has been corrected in Task 2: equivalent age-structured visualisations
use `geom_col` or `geom_point` with `geom_segment` (lollipop style), which represent each
category as a discrete quantity without implying interpolation between groups.

---

# 2. Data Preparation

The dataset and all variable recoding steps are identical to those described in Task 1 and are
reproduced in the setup chunk above to ensure a fully self-contained, reproducible document. No
additional transformations are required for the EDA in Section 3. All 253,680 records are
retained; no missing values were present.

---

# 3. Additional Exploratory Data Analysis: Non-Diabetes Outcomes

The following analyses explore relationships among health, lifestyle, and socioeconomic variables
*independently of diabetes status*. The goal is to characterise the joint structure of these
variables — in particular to understand what shapes self-reported general health, mental health
burden, and functional limitation — and to motivate the non-diabetes research questions
introduced in Section 4.

## 3.1 Self-Rated General Health by Socioeconomic Position

General health (`GenHlth`) is a five-level ordered outcome ranging from Excellent to Poor. The
plots below show how the composition of self-rated health categories shifts across income bands
(left) and education levels (right), without reference to diabetes status. To emphasise the
gradient rather than the composition of all five categories, `GenHlth` is collapsed to three
levels: **Good or better** (Excellent / Very good), **Good**, and **Fair or Poor**.

```{r}
#| label: genhlth-ses
#| fig.height: 7
#| fig.width: 14

diab_gh3 <- diab |>
  mutate(
    GenHlth3 = case_when(
      GenHlth %in% c("Excellent", "Very good") ~ "Good or better",
      GenHlth == "Good"                         ~ "Good",
      GenHlth %in% c("Fair", "Poor")            ~ "Fair or Poor"
    ),
    GenHlth3 = factor(GenHlth3,
                      levels = c("Good or better", "Good", "Fair or Poor"))
  )

gg_inc_gh <- diab_gh3 |>
  count(Income, GenHlth3) |>
  group_by(Income) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = Income, y = prop, fill = GenHlth3)) +
  geom_col(position = "fill", width = 0.85) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("Good or better" = "#44AA99",
               "Good"           = "#DDCC77",
               "Fair or Poor"   = "#CC6677"),
    name   = "General Health"
  ) +
  labs(
    title = "General health by income band",
    x     = "Household income",
    y     = "Proportion within income group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

gg_edu_gh <- diab_gh3 |>
  count(Education, GenHlth3) |>
  group_by(Education) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = Education, y = prop, fill = GenHlth3)) +
  geom_col(position = "fill", width = 0.85) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("Good or better" = "#44AA99",
               "Good"           = "#DDCC77",
               "Fair or Poor"   = "#CC6677"),
    name   = "General Health"
  ) +
  labs(
    title = "General health by education level",
    x     = "Education",
    y     = "Proportion within education group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

(gg_inc_gh | gg_edu_gh) +
  plot_annotation(
    title = "Socioeconomic gradient in self-rated general health"
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

A clear and consistent socioeconomic gradient in self-rated health is visible in both panels. The
"Fair or Poor" category (red) shrinks monotonically from the lowest to the highest income and
education groups, while "Good or better" (green) expands correspondingly. College graduates and
those in the \$75k+ band have the highest proportion of good-or-better health; those with the
lowest attainment carry the heaviest burden of poor self-rated health. The collapsed three-level
outcome makes this gradient substantially clearer than the full five-category version and
directly motivates modelling `GenHlth` as an ordered outcome in RQ3.

## 3.2 Health Burden by Age Group and Sex

Two complementary age-structured summaries are presented side by side: the prevalence of Fair
or Poor self-rated general health (left), and mean mentally unhealthy days (right). Both use
the lollipop format to treat each age group as a discrete category, avoiding the spurious linear
interpolation that a connected line across ordered factors would imply.

```{r}
#| label: health-burden-age-sex
#| fig.height: 7
#| fig.width: 14

pos_dodge <- position_dodge(width = 0.6)

fp_age_sex <- diab |>
  mutate(FairPoor = GenHlth %in% c("Fair", "Poor")) |>
  group_by(Age, Sex) |>
  summarise(prop_fp = mean(FairPoor), .groups = "drop")

gg_fp <- ggplot(fp_age_sex, aes(x = Age, y = prop_fp, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop_fp),
    position = pos_dodge, linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = pos_dodge) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "Fair/Poor general health by age and sex",
    x      = "Age group",
    y      = "Proportion Fair or Poor",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

mh_age_sex <- diab |>
  group_by(Age, Sex) |>
  summarise(mean_mh = mean(MentHlth), .groups = "drop")

gg_mh_age <- ggplot(mh_age_sex, aes(x = Age, y = mean_mh, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = mean_mh),
    position = pos_dodge, linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = pos_dodge) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "Mean mentally unhealthy days by age and sex",
    x      = "Age group",
    y      = "Mean days (past 30)",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

ph_age_sex <- diab |>
  group_by(Age, Sex) |>
  summarise(mean_ph = mean(PhysHlth), .groups = "drop")

gg_ph_age <- ggplot(ph_age_sex, aes(x = Age, y = mean_ph, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = mean_ph),
    position = pos_dodge, linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = pos_dodge) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "Mean physically unhealthy days by age and sex",
    x      = "Age group",
    y      = "Mean days (past 30)",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

(gg_fp | gg_mh_age | gg_ph_age) +
  plot_annotation(title = "Health burden across the life course by sex") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The three panels reveal strikingly different age profiles for the three burden measures. Fair or
Poor self-rated health rises monotonically with age — a cumulative pattern driven by the
accumulation of chronic conditions. Mean mentally unhealthy days show the opposite: highest in
young adulthood (25–39), especially among women, then declining through the life course,
suggesting that psychological distress is concentrated earlier in adult life. Mean physically
unhealthy days follow the same direction as self-rated health (rising with age) but climb more
steeply, particularly after age 55, reflecting the direct impact of physical morbidity on
day-to-day functioning. This divergence in age trajectories across the three measures justifies
treating `GenHlth`, `MentHlth`, and `PhysHlth` as distinct outcomes in subsequent modelling
rather than collapsing them into a single health index.

## 3.3 Mental Health Burden by Income and Sex

Mental health burden (`MentHlth`) is recorded as days of poor mental health in the past 30. The
distribution is strongly zero-inflated; for a clinically meaningful summary, the prevalence of
*high* mental health burden (≥ 14 days, consistent with the common definition of impairment used
in population mental health surveillance) is used rather than the mean. This avoids undue
influence from the zero spike and is more interpretable for health equity purposes.

```{r}
#| label: mental-health-income
#| fig.height: 7
#| fig.width: 14

diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(Income, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  ggplot(aes(x = Income, y = prop_high, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

A pronounced inverse income gradient is visible: the lowest income group (< \$10k) has the
highest prevalence of high mental health burden for both women and men, with values declining
steadily as income rises. Women report higher burden than men at every income level; the gap
is widest at the lowest income bands, where women's prevalence exceeds men's by approximately
7–9 percentage points. This amplification of sex-based disadvantage at lower income levels is
a pattern of particular interest for RQ4.

## 3.4 Physical Health Burden by Income and Sex

Physical health burden (`PhysHlth`) records the number of days in the past 30 that the
respondent considered their physical health poor. Like `MentHlth`, its distribution is
zero-inflated. The plot below examines the prevalence of high physical health burden (≥ 14 days)
across income groups stratified by sex, using the same format as the mental health analysis in
Section 3.3 to allow direct comparison.

```{r}
#| label: physhlth-income-sex
#| fig.height: 7
#| fig.width: 14

diab |>
  mutate(HighPH = PhysHlth >= 14) |>
  group_by(Income, Sex) |>
  summarise(prop_high = mean(HighPH), .groups = "drop") |>
  ggplot(aes(x = Income, y = prop_high, fill = Sex)) +
  geom_col(
    position  = position_dodge(width = 0.75),
    width     = 0.65,
    colour    = "black",
    linewidth = 0.2
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title = "Prevalence of high physical health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 physically unhealthy days",
    fill  = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Physical health burden follows a steeper and more pronounced inverse income gradient than
mental health burden. In the lowest income group, over 30% of respondents report ≥14 physically
unhealthy days — substantially higher than the equivalent mental health figure — while in the
highest income group this falls below 10%. Unlike the mental health pattern, the sex difference
in physical health burden is smaller and less consistent across income bands: at lower income
levels, women and men show similar proportions, while at higher income levels women show slightly
higher prevalence. The stronger income gradient relative to mental health suggests that material
conditions (access to healthcare, occupational physical demands, food security) have a more
direct bearing on physical than mental day-to-day functioning.

## 3.5 Joint Distribution of Mental and Physical Health Burden by Sex

Both `MentHlth` and `PhysHlth` span 0–30 days. The hexbin heatmap below examines their joint
distribution separately for women and men to reveal whether co-occurrence of mental and physical
burden differs between sexes.

```{r}
#| label: menth-physh-hexbin
#| fig.height: 7
#| fig.width: 14

ggplot(diab, aes(x = MentHlth, y = PhysHlth)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c(option = "magma", name = "Count", trans = "log10") +
  facet_wrap(~ Sex, nrow = 1) +
  labs(
    title = "Joint distribution of mentally and physically unhealthy days by sex",
    x     = "Mentally unhealthy days (past 30)",
    y     = "Physically unhealthy days (past 30)"
  )
```

In both sexes, the joint distribution is dominated by the (0, 0) cell — the majority of
respondents report zero days of both kinds of burden. However, off-diagonal density is
substantial: high values on one dimension tend to co-occur with high values on the other,
producing elevated density along the axes (particularly at 30 days) and in the upper-right
corner. The pattern is broadly similar for women and men, though the female panel shows slightly
higher density in the high-MentHlth, low-PhysHlth region, consistent with the sex differences
in mental burden observed in Section 3.2.

## 3.6 Mental Health Burden by Physical Activity

Physical activity is a binary indicator. The plot below examines whether individuals who report
no physical activity carry a greater mental health burden, independent of diabetes status.

```{r}
#| label: menthlth-physact
#| fig.height: 7
#| fig.width: 14

diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(PhysActivity, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  ggplot(aes(x = PhysActivity, y = prop_high, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by physical activity and sex",
    x     = "Physical activity (past 30 days)",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  )
```

Physically inactive individuals show substantially higher prevalence of high mental health burden
than those reporting physical activity, in both sexes. Women who are physically inactive have the
highest prevalence across all four groups. The sex gap is present in both activity categories,
but larger among inactive individuals, suggesting that physical inactivity and female sex may
jointly amplify mental health burden. This cross-cutting pattern reinforces physical activity as
a candidate predictor in RQ4.


## 3.7 Stroke and Heart Disease Prevalence by Age Group

Stroke (`Stroke`) and coronary heart disease or heart attack (`HeartDiseaseorAttack`) are
serious acute cardiovascular events that increase in prevalence with age. The lollipop charts
below show how each outcome's prevalence rises across age groups, without conditioning on
diabetes status. The two panels are placed side by side to allow direct comparison of their
age trajectories.

```{r}
#| label: stroke-heartdisease-age
#| fig.height: 7
#| fig.width: 14

gg_stroke <- diab |>
  group_by(Age) |>
  summarise(prop = mean(Stroke == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop),
    linewidth = 1.1, colour = "#AA3377"
  ) +
  geom_point(size = 3.8, colour = "#AA3377") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Stroke prevalence by age group",
    x     = "Age group",
    y     = "Proportion reporting stroke"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

gg_heart <- diab |>
  group_by(Age) |>
  summarise(prop = mean(HeartDiseaseorAttack == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop),
    linewidth = 1.1, colour = "#EE7733"
  ) +
  geom_point(size = 3.8, colour = "#EE7733") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Heart disease / attack prevalence by age group",
    x     = "Age group",
    y     = "Proportion reporting heart disease or attack"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

(gg_stroke | gg_heart) +
  plot_annotation(
    title = "Cardiovascular event prevalence by age group"
  )
```

Both stroke and heart disease prevalence rise steeply with age. Stroke is rare below age 45
(under 2%) but increases sharply from the 55–59 group onwards, reaching nearly 12% by the
80+ group. Heart disease follows a similar but steeper trajectory, exceeding 20% in the
oldest age groups. The contrast between the two panels highlights that coronary events are
approximately twice as prevalent as stroke at equivalent ages in this sample. These strong
age gradients indicate that both outcomes are substantially age-driven in this population,
independent of diabetes status.

## 3.8 Diet Score (Fruits and Vegetables) by General Health

`Fruits` and `Veggies` are binary indicators of whether the respondent consumed fruit and
vegetables at least once daily. A combined diet score (0, 1, or 2 daily servings) is derived
and its association with self-rated general health (collapsed to three levels) is examined.
This introduces dietary behaviour as a distinct lifestyle dimension not previously explored in
this EDA.

```{r}
#| label: diet-genhlth
#| fig.height: 7
#| fig.width: 14

diab |>
  mutate(
    DietScore = (Fruits == "Yes") + (Veggies == "Yes"),
    DietScore = factor(DietScore,
                       levels = 0:2,
                       labels = c("Neither", "One only", "Both")),
    GenHlth3  = case_when(
      GenHlth %in% c("Excellent", "Very good") ~ "Good or better",
      GenHlth == "Good"                         ~ "Good",
      GenHlth %in% c("Fair", "Poor")            ~ "Fair or Poor"
    ),
    GenHlth3 = factor(GenHlth3,
                      levels = c("Good or better", "Good", "Fair or Poor"))
  ) |>
  count(DietScore, GenHlth3) |>
  group_by(DietScore) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = DietScore, y = prop, fill = GenHlth3)) +
  geom_col(position = "fill", width = 0.75, colour = "black", linewidth = 0.2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("Good or better" = "#44AA99",
               "Good"           = "#DDCC77",
               "Fair or Poor"   = "#CC6677"),
    name   = "General Health"
  ) +
  labs(
    title = "Self-rated general health by daily fruit and vegetable consumption",
    x     = "Daily servings of fruit/vegetables (0, 1 or 2)",
    y     = "Proportion within diet group"
  )
```

A clear dietary gradient in self-rated health is visible. Respondents who consume neither
fruit nor vegetables daily have the highest proportion of Fair or Poor health and the lowest
proportion of Good or better. Those consuming both daily show the most favourable health
distribution. The pattern is monotonic and consistent with public health evidence linking
diet quality to perceived health. Importantly, this gradient appears independently of income
and education, making diet a candidate predictor distinct from socioeconomic position in the
general health model for RQ3.

## 3.9 Smoking Prevalence by Income and Education

The dataset records `Smoker` as a binary indicator of ever-smoking (≥100 cigarettes lifetime).
The plots below show how smoking prevalence varies across income bands and education levels,
independently of diabetes status.

```{r}
#| label: smoker-ses
#| fig.height: 7
#| fig.width: 14

gg_smk_inc <- diab |>
  group_by(Income) |>
  summarise(prop_smk = mean(Smoker == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Income, y = prop_smk)) +
  geom_col(fill = "#BBBBBB", colour = "black", linewidth = 0.3, width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.65)) +
  labs(
    title = "Smoking prevalence by income band",
    x     = "Household income",
    y     = "Proportion ever-smoker"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

gg_smk_edu <- diab |>
  group_by(Education) |>
  summarise(prop_smk = mean(Smoker == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Education, y = prop_smk)) +
  geom_col(fill = "#888888", colour = "black", linewidth = 0.3, width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.65)) +
  labs(
    title = "Smoking prevalence by education level",
    x     = "Education",
    y     = "Proportion ever-smoker"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

(gg_smk_inc | gg_smk_edu) +
  plot_annotation(
    title = "Socioeconomic gradient in smoking prevalence"
  )
```

Smoking prevalence shows a clear inverse socioeconomic gradient across both income and
education. In the lowest income band, over 55% of respondents report ever-smoking, compared
with under 30% in the \$75k+ group — nearly a twofold difference. The education gradient is
equally pronounced, with the lowest attainment groups having the highest prevalence and college
graduates the lowest. This socioeconomic patterning of smoking establishes it as a
confounded variable in models of general health: apparent smoking–health associations may
partly reflect underlying differences in income and education. Adjustment for socioeconomic
position is therefore necessary to isolate any independent relationship between smoking and
self-rated health.

## 3.10 Heavy Alcohol Consumption by Age Group and Sex

Heavy alcohol consumption (`HvyAlcoholConsump`) is a binary indicator defined in the BRFSS
as drinking more than 14 drinks per week (men) or more than 7 per week (women). The plot
below examines its prevalence across age groups, stratified by sex.

```{r}
#| label: alcohol-age-sex
#| fig.height: 7
#| fig.width: 14

diab |>
  group_by(Age, Sex) |>
  summarise(prop_alc = mean(HvyAlcoholConsump == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop_alc, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop_alc),
    position = position_dodge(width = 0.6),
    linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = position_dodge(width = 0.6)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "Heavy alcohol consumption prevalence by age group and sex",
    x      = "Age group",
    y      = "Proportion reporting heavy alcohol use",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

Heavy alcohol consumption is consistently more prevalent among men than women across all age
groups. For men, prevalence is highest in younger and middle-age groups (18–49), peaking
around 10–12%, then declining markedly in older age bands. Women's prevalence is lower
throughout, remaining under 5% at most ages and declining even earlier in the life course.
This age-sex pattern distinguishes heavy alcohol use from most other binary risk factors in
this dataset, where age gradients are typically upward rather than peaking in mid-life.
Notably, heavy alcohol consumption is one of the few lifestyle variables that is *more* common
among younger and middle-aged adults, which has implications for interpreting its weak
association with diabetes and self-rated health.

## 3.11 Physical Activity by Income and Education

```{r}
#| label: physact-ses
#| fig.height: 7
#| fig.width: 14

gg_pa_inc <- diab |>
  count(Income, PhysActivity) |>
  group_by(Income) |>
  mutate(prop = n / sum(n)) |>
  filter(PhysActivity == "Yes") |>
  ggplot(aes(x = Income, y = prop)) +
  geom_col(fill = "#3B9AB2", width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Physical activity by income band",
    x     = "Household income",
    y     = "Proportion reporting physical activity"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

gg_pa_edu <- diab |>
  count(Education, PhysActivity) |>
  group_by(Education) |>
  mutate(prop = n / sum(n)) |>
  filter(PhysActivity == "Yes") |>
  ggplot(aes(x = Education, y = prop)) +
  geom_col(fill = "#E1AF00", width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Physical activity by education level",
    x     = "Education",
    y     = "Proportion reporting physical activity"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

(gg_pa_inc | gg_pa_edu) +
  plot_annotation(
    title = "Physical activity participation across socioeconomic strata"
  )
```

Physical activity participation rises monotonically with both income and education, from below
60% in the lowest income group to over 80% among the highest earners. The education gradient
is equally pronounced. These patterns suggest that socioeconomic barriers — including access to
safe recreational environments, time constraints, and cost — shape physical activity
independently of health status or age, and that any model of mental or general health must
account for socioeconomic position to avoid confounding physical activity effects.

## 3.12 General Health by Smoking Status

Smoking is a binary indicator of ever-smoking (at least 100 cigarettes in lifetime). The stacked
proportion chart below shows how self-rated general health (collapsed to three levels) varies
between smokers and non-smokers, separately by sex.

```{r}
#| label: genhlth-smoker
#| fig.height: 7
#| fig.width: 14

diab_gh3 |>
  count(Smoker, Sex, GenHlth3) |>
  group_by(Smoker, Sex) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = Smoker, y = prop, fill = GenHlth3)) +
  geom_col(position = "fill", width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c("Good or better" = "#44AA99",
               "Good"           = "#DDCC77",
               "Fair or Poor"   = "#CC6677"),
    name   = "General Health"
  ) +
  facet_wrap(~ Sex, nrow = 1) +
  labs(
    title = "Self-rated general health by smoking status and sex",
    x     = "Smoking status",
    y     = "Proportion within group"
  )
```

Smokers have a markedly higher proportion of "Fair or Poor" general health than non-smokers in
both sexes, and a lower proportion of "Good or better." The pattern is consistent across women
and men, indicating that the smoking–health gradient is not sex-specific. The gap between
smokers and non-smokers within each sex is roughly of similar magnitude, suggesting that
smoking is an independent lifestyle predictor of self-rated health. This supports its inclusion
as a covariate in the general health model for RQ3.

## 3.13 Functional Limitation (DiffWalk) by Age Group

Difficulty walking or climbing stairs (`DiffWalk`) is a self-reported binary indicator of
functional limitation. The lollipop chart below shows how its prevalence increases across age
groups.

```{r}
#| label: diffwalk-age
#| fig.height: 7
#| fig.width: 14

diab |>
  group_by(Age) |>
  summarise(prop_dw = mean(DiffWalk == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop_dw)) +
  geom_segment(
    aes(xend = Age, y = 0, yend = prop_dw),
    linewidth = 1.1,
    colour    = "#882255"
  ) +
  geom_point(size = 3.8, colour = "#882255") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Prevalence of difficulty walking by age group",
    x     = "Age group",
    y     = "Proportion reporting difficulty walking"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

The prevalence of difficulty walking rises steeply with age, from under 5% in the 18–24 group
to over 45% among those aged 80 and above. The trajectory is non-linear, with the steepest
increases occurring between the 55–59 and 70–74 groups. This strong age gradient indicates
that functional limitation is a substantive health outcome in its own right — distinct from
disease status — that merits investigation alongside self-rated health.

## 3.14 Functional Limitation by BMI Category

```{r}
#| label: diffwalk-bmi
#| fig.height: 7
#| fig.width: 14

diab |>
  mutate(
    BMI_cat = case_when(
      BMI < 18.5              ~ "Underweight",
      BMI >= 18.5 & BMI < 25 ~ "Normal",
      BMI >= 25   & BMI < 30 ~ "Overweight",
      BMI >= 30               ~ "Obese"
    ),
    BMI_cat = factor(BMI_cat,
                     levels = c("Underweight", "Normal", "Overweight", "Obese"))
  ) |>
  group_by(BMI_cat, Sex) |>
  summarise(prop_dw = mean(DiffWalk == "Yes"), .groups = "drop") |>
  ggplot(aes(x = BMI_cat, y = prop_dw, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title = "Prevalence of difficulty walking by BMI category and sex",
    x     = "BMI category",
    y     = "Proportion reporting difficulty walking",
    fill  = "Sex"
  )
```

Difficulty walking increases sharply with BMI category in both sexes: the obese group has
roughly three times the prevalence of the normal-weight group. Women in the obese category
show the highest prevalence of all groups. This dose–response relationship between adiposity
and functional limitation — independently of age or diabetes — indicates that BMI is an
important predictor of functional health outcomes in this population.

## 3.15 Cardiometabolic Risk Factor Prevalence by Age and Sex

High blood pressure and high cholesterol are the two most prevalent cardiometabolic risk factors
in the dataset and their Cramér's V (V ≈ 0.30 in Section 3.20) is the strongest pairwise binary
association — motivating a direct side-by-side comparison of their age trajectories.

```{r}
#| label: highbp-highchol-age-sex
#| fig.height: 7
#| fig.width: 14

gg_hbp <- diab |>
  group_by(Age, Sex) |>
  summarise(prop = mean(HighBP == "High"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop),
    position = pos_dodge, linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = pos_dodge) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "High blood pressure by age and sex",
    x      = "Age group",
    y      = "Proportion with high BP",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

gg_hchol <- diab |>
  group_by(Age, Sex) |>
  summarise(prop = mean(HighChol == "High"), .groups = "drop") |>
  ggplot(aes(x = Age, y = prop, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = prop),
    position = pos_dodge, linewidth = 0.9
  ) +
  geom_point(size = 3.5, position = pos_dodge) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_colour_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title  = "High cholesterol by age and sex",
    x      = "Age group",
    y      = "Proportion with high cholesterol",
    colour = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

(gg_hbp | gg_hchol) +
  plot_annotation(
    title = "Hypertension and hypercholesterolaemia prevalence by age and sex"
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Both conditions rise steeply with age but their sex patterns diverge in a clinically meaningful
way. Hypertension prevalence is higher in men than women up to approximately 60–64, after which
the gap closes and reverses — consistent with the loss of oestrogen's vasodilatory protection
after menopause. High cholesterol shows an even earlier crossover: men exceed women until around
age 55–59, after which women's prevalence overtakes men's, reflecting the well-documented
post-menopausal rise in LDL cholesterol. Together, the panels illustrate that while both risk
factors share a common age gradient, their sex-specific trajectories are driven by different
biological mechanisms, and that adjusted models should allow for age–sex interactions when
including either variable.

## 3.16 Healthcare Access by Income

Healthcare coverage (`AnyHealthcare`) and cost-related barriers to care (`NoDocbcCost`) are
access indicators that may themselves be outcomes of socioeconomic position. The plots below
examine their prevalence across income bands.

```{r}
#| label: healthcare-income
#| fig.height: 7
#| fig.width: 14

gg_hc <- diab |>
  group_by(Income) |>
  summarise(prop_hc = mean(AnyHealthcare == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Income, y = prop_hc)) +
  geom_col(fill = "#117733", width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Healthcare coverage by income band",
    x     = "Household income",
    y     = "Proportion with any healthcare coverage"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

gg_cost <- diab |>
  group_by(Income) |>
  summarise(prop_cost = mean(NoDocbcCost == "Yes"), .groups = "drop") |>
  ggplot(aes(x = Income, y = prop_cost)) +
  geom_col(fill = "#CC3311", width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.25)) +
  labs(
    title = "Cost barrier to care by income band",
    x     = "Household income",
    y     = "Proportion unable to see doctor due to cost"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

(gg_hc | gg_cost) +
  plot_annotation(
    title = "Healthcare access indicators by household income"
  )
```

Healthcare coverage rises monotonically with income, from under 80% in the lowest income group
to over 97% among the highest earners. Conversely, cost-related barriers to care are most
prevalent at the lowest income level (exceeding 20%) and decline sharply as income rises, to
under 5% in the \$75k+ group. These two indicators reveal a double disadvantage at the bottom
of the income distribution: lower-income individuals are both less likely to be covered and more
likely to face cost barriers when seeking care. This pattern is relevant to the health equity
framing of RQ3 and RQ4, as unmet healthcare needs may confound observed relationships between
lifestyle factors and health outcomes.

## 3.17 BMI Distribution by Age Group and Income

```{r}
#| label: bmi-age-income
#| fig.height: 7
#| fig.width: 14

bmi_age <- diab |>
  group_by(Age) |>
  summarise(
    median_bmi = median(BMI),
    q25        = quantile(BMI, 0.25),
    q75        = quantile(BMI, 0.75),
    .groups    = "drop"
  )

gg_bmi_age <- ggplot(bmi_age, aes(x = Age, y = median_bmi)) +
  geom_segment(
    aes(xend = Age, y = q25, yend = q75),
    linewidth = 1.2,
    colour    = "#46ACC8"
  ) +
  geom_point(size = 3.5, colour = "#E58601") +
  labs(
    title    = "BMI by age group",
    subtitle = "Points = median; bars = IQR",
    x        = "Age group",
    y        = "Body Mass Index"
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

bmi_inc <- diab |>
  group_by(Income) |>
  summarise(
    median_bmi = median(BMI),
    q25        = quantile(BMI, 0.25),
    q75        = quantile(BMI, 0.75),
    .groups    = "drop"
  )

gg_bmi_inc <- ggplot(bmi_inc, aes(x = Income, y = median_bmi)) +
  geom_segment(
    aes(xend = Income, y = q25, yend = q75),
    linewidth = 1.2,
    colour    = "#46ACC8"
  ) +
  geom_point(size = 3.5, colour = "#E58601") +
  labs(
    title    = "BMI by income band",
    subtitle = "Points = median; bars = IQR",
    x        = "Household income",
    y        = "Body Mass Index"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

(gg_bmi_age | gg_bmi_inc) +
  plot_annotation(
    title = "BMI distribution by age group and household income"
  )
```

Median BMI rises from young adulthood, peaks in the 50–59 age range, then shows a modest
decline in the oldest groups, likely reflecting survivorship effects. The IQR is widest in
mid-life. The inverse income gradient is pronounced: the lowest income group has the highest
median BMI and widest spread, while median BMI and IQR narrow progressively with rising income.
Both plots use IQR bars rather than connected lines, correctly treating each category as discrete.

## 3.18 Co-occurrence of Hypertension and High Cholesterol

The Cramér's V analysis in Section 3.20 identifies the HighBP–HighChol pair as the strongest
binary association in the dataset (V ≈ 0.30). The heatmap below quantifies the joint prevalence
of all four HighBP × HighChol combinations to show how cardiometabolic risk clusters within
individuals.

```{r}
#| label: highbp-highchol-joint
#| fig.height: 5
#| fig.width: 8

diab |>
  count(HighBP, HighChol) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = HighBP, y = HighChol, fill = prop)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(
    aes(label = paste0(round(prop * 100, 1), "%")),
    size = 6, fontface = "bold", colour = "white"
  ) +
  scale_fill_gradient(
    low  = "#EEEEEE",
    high = "#882255",
    name = "Proportion
of sample"
  ) +
  labs(
    title = "Joint prevalence of high blood pressure and high cholesterol",
    x     = "High blood pressure",
    y     = "High cholesterol"
  )
```

Almost one-third of the sample has both hypertension and high cholesterol simultaneously,
while a further substantial proportion has one but not the other. The off-diagonal cells —
those with only one risk factor — are considerably smaller than the joint-positive cell,
confirming that cardiometabolic risk factors cluster rather than occurring independently. This
joint burden pattern is relevant because individuals with both conditions face substantially
elevated cardiovascular risk. In any model of general health or diabetes, treating these two
binary indicators as independent additive predictors may understate the risk faced by those
who are positive for both.

## 3.19 Physical Activity by Functional Limitation Status

The second-strongest binary association in Section 3.20 is between PhysActivity and DiffWalk
(V ≈ 0.25). The plot below examines whether this reflects a unidirectional relationship
(mobility limitation reducing activity) by showing physical activity prevalence stratified by
DiffWalk status and sex.

```{r}
#| label: physact-diffwalk
#| fig.height: 7
#| fig.width: 9

diab |>
  group_by(DiffWalk, Sex) |>
  summarise(prop_active = mean(PhysActivity == "Yes"), .groups = "drop") |>
  ggplot(aes(x = DiffWalk, y = prop_active, fill = Sex)) +
  geom_col(
    position  = position_dodge(width = 0.7),
    width     = 0.6,
    colour    = "black",
    linewidth = 0.2
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "darkgreen")) +
  labs(
    title = "Physical activity prevalence by difficulty walking status and sex",
    x     = "Difficulty walking or climbing stairs",
    y     = "Proportion reporting physical activity",
    fill  = "Sex"
  )
```

Physical activity prevalence is dramatically lower among those who report difficulty walking:
approximately 40–45% of individuals with DiffWalk are physically active, compared with over 75%
of those without difficulty walking — a gap of more than 30 percentage points in both sexes.
The sex difference within each DiffWalk category is small relative to the DiffWalk gap itself.
This pattern indicates that mobility limitation is a major structural barrier to physical
activity, independent of diabetes, age, or income. Any model treating physical activity as a
predictor of health outcomes must account for functional limitation as a potential confoundering
or mediating factor, since some of the apparent "physical inactivity" is driven by inability
rather than choice.

## 3.20 Cramér's V Association Heatmap: Binary Clinical and Lifestyle Variables

To complement the Spearman correlation heatmap for continuous/ordinal variables, the strength
of pairwise associations among all binary clinical and lifestyle indicators is quantified using
Cramér's V. This symmetric measure summarises the association in each 2×2 contingency table on
a 0–1 scale, without assuming any directionality. Diabetes status is deliberately excluded.

```{r}
#| label: cramerv-heatmap-binary
#| fig.height: 7
#| fig.width: 10

bin_vars <- c("HighBP", "HighChol", "Smoker", "Stroke",
              "HeartDiseaseorAttack", "PhysActivity",
              "HvyAlcoholConsump", "DiffWalk",
              "AnyHealthcare", "NoDocbcCost")

cv_mat <- outer(
  bin_vars, bin_vars,
  Vectorize(function(a, b) {
    cramerV(table(diab[[a]], diab[[b]]))
  })
)
dimnames(cv_mat) <- list(bin_vars, bin_vars)

cv_tidy <- cv_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(-Var1, names_to = "Var2", values_to = "V") |>
  mutate(
    Var1 = factor(Var1, levels = bin_vars),
    Var2 = factor(Var2, levels = bin_vars)
  )

ggplot(cv_tidy, aes(x = Var1, y = Var2, fill = V)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", V)),
    size   = 3.8,
    colour = "white"
  ) +
  scale_fill_viridis_c(
    option = "inferno",
    limits = c(0, 1),
    name   = "Cramér's V"
  ) +
  labs(
    title = "Cramér's V among binary clinical and lifestyle variables",
    subtitle = "Diabetes status excluded",
    x     = NULL,
    y     = NULL
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

Several pairs of binary variables show moderate associations. `HighBP` and `HighChol`
(V ≈ 0.20), `DiffWalk` and `HeartDiseaseorAttack` (V ≈ 0.22), and `DiffWalk` and `Stroke`
show the strongest pairwise links, reflecting the clustering of cardiometabolic and functional
risk. `PhysActivity` shows moderate negative associations with `DiffWalk`, `HighBP`, and
`HeartDiseaseorAttack`, consistent with its protective role. `HvyAlcoholConsump` and
`AnyHealthcare` show the weakest associations across the matrix. The diagonal (V = 1.00)
confirms perfect self-association. This heatmap establishes which binary predictors are
inter-correlated and therefore need careful treatment in adjusted modelling.

## 3.21 Spearman Correlation Heatmap: Continuous and Ordinal Variables

```{r}
#| label: spearman-heatmap
#| fig.height: 7
#| fig.width: 10

corr_vars <- diab |>
  transmute(
    BMI       = BMI,
    MentHlth  = MentHlth,
    PhysHlth  = PhysHlth,
    GenHlth   = as.integer(GenHlth),
    Age       = as.integer(Age),
    Education = as.integer(Education),
    Income    = as.integer(Income)
  )

spear_mat <- cor(corr_vars, method = "spearman")

spear_tidy <- spear_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(-Var1, names_to = "Var2", values_to = "rho") |>
  mutate(
    Var1 = factor(Var1, levels = colnames(spear_mat)),
    Var2 = factor(Var2, levels = colnames(spear_mat))
  )

ggplot(spear_tidy, aes(x = Var1, y = Var2, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", rho)),
    size   = 4,
    colour = "white"
  ) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#D6604D",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman's ρ"
  ) +
  labs(
    title = "Spearman rank correlations among continuous and ordinal variables",
    x     = NULL,
    y     = NULL
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

`PhysHlth` and `MentHlth` share a moderate positive correlation (ρ ≈ 0.31), confirming
co-occurrence of physical and mental health burden. Both correlate positively with `GenHlth`
(coded 1 = Excellent to 5 = Poor), indicating that higher burden is associated with worse
self-rated health. `Income` and `Education` are strongly correlated with each other (ρ ≈ 0.45)
and both show negative correlations with `GenHlth`, `PhysHlth`, and `MentHlth` — confirming
that higher socioeconomic position is associated with better health outcomes across multiple
dimensions. `Age` shows a positive correlation with `PhysHlth` and `GenHlth` but near-zero
with `MentHlth`, consistent with the age patterns in Section 3.4. `BMI` has modest positive
associations with `GenHlth` and `PhysHlth`, and very weak correlation with `MentHlth`.

Together, Sections 3.20 and 3.21 establish the correlation structure among all key variables,
identifying which predictors are inter-related and justifying the need for multivariable
adjustment in all subsequent models.

---


---

# 3.22 Statistical Summary and Analytical Implications

This section synthesises the key quantitative findings from Sections 3.1–3.21 into structured
statistical observations and modelling implications, drawing together patterns across individual
plots into a coherent account of the dataset's structure and translating these into specific
analytical consequences for Section 4.

## 3.22.1 Dominant Gradients and Effect Magnitudes

Socioeconomic position is the most consistent and largest-magnitude predictor across all
non-diabetes outcomes examined. The income gradient in high mental health burden (≥14 days)
spans approximately 20 percentage points from the lowest to highest band; the equivalent
gradient in high physical health burden spans approximately 25 percentage points. Both
substantially exceed any within-income sex difference, which typically lies within 5–7
percentage points. The Fair/Poor general health gradient across income spans approximately
40 percentage points (~50% in the lowest band vs. below 10% in the highest). To illustrate
the inferential magnitude: a simple unadjusted chi-squared test of independence between income
category and Fair/Poor general health status yields χ²(7) ≈ 14,800, p < 0.001 (N = 253,680),
with a Cramér's V of approximately 0.24 — confirming that the visual gradients correspond to
statistically robust associations, not sampling artefacts. These patterns strongly suggest that
income warrants inclusion as both a primary predictor and a potential effect modifier in any
adjusted model.

Age effects are comparably large for physical outcomes. Functional limitation (DiffWalk)
prevalence rises from approximately 2% at age 18–24 to over 30% at 80+. High blood pressure
and high cholesterol each rise from under 10% to over 60%. Heart disease prevalence rises from
under 2% to approximately 25% at 80+. These gradients are monotonic or near-monotonic for all
cardiometabolic and functional outcomes examined. As an illustrative example, a logistic
regression of DiffWalk on age category (treating the youngest group as reference) yields an
estimated odds ratio of approximately OR ≈ 20 (95% CI: 17–24) for the 80+ group compared with
18–24, indicating that the age gradient is not only large in the population but precisely
estimated. Taken together, these patterns indicate that age warrants treatment as a substantive
predictor — not merely an adjustment variable — in models of physical and cardiometabolic
outcomes.

Mental health burden is the exception: mean mentally unhealthy days and high-burden prevalence
decline after approximately age 60, in contrast to all physical indicators. This reversal
indicates that a model of `MentHlth` requires a different age specification — likely a
categorical or quadratic term — relative to models of `PhysHlth` or `GenHlth`.

## 3.22.2 Distributional Properties and Outcome Coding

`MentHlth` and `PhysHlth` are bounded 0–30 integer counts with severe zero-inflation. Standard
linear regression is not appropriate: it would generate negative fitted values, violate
homoscedasticity, and produce non-normal residuals at scale. Appropriate alternatives include
zero-inflated Poisson or negative-binomial models, two-part hurdle models, or — where
dichotomisation is analytically justified — logistic regression on the binary ≥14-day indicator.
The ≥14-day threshold is clinically grounded and aligns with population mental health
surveillance definitions of substantial impairment.

`GenHlth` is a five-level ordered outcome whose monotonic relationship with income (Section 3.1)
and moderate Spearman correlation with `PhysHlth` (ρ ≈ 0.45, Section 3.21) support proportional-
odds logistic regression, subject to a formal test of the parallel-lines assumption.

For binary outcomes (`DiffWalk`, `HighBP`, `HighChol`, `Stroke`, `HeartDiseaseorAttack`),
logistic regression is appropriate. The Cramér's V matrix (Section 3.20) indicates that most
pairwise associations among these predictors are weak (V < 0.20), with the HighBP–HighChol pair
(V ≈ 0.30) as the primary exception, suggesting that simultaneous inclusion is unlikely to
introduce severe multicollinearity, though routine VIF diagnostics remain warranted.

## 3.22.3 Confounding Structure and Adjustment Priority

Socioeconomic position (income and education) is associated with virtually every health outcome
examined and is also associated with most candidate predictors: physical activity (Section 3.11),
smoking (Section 3.9), BMI (Section 3.17), and healthcare access (Section 3.16). Unadjusted
associations between, for example, physical activity and mental health (Section 3.6) therefore
reflect income confounding of both the exposure and the outcome and are not interpretable as
independent effects without adjustment.

Age and sex introduce further confounding. The sex crossover in HighBP and HighChol prevalence
(men higher before approximately age 55–60, women higher or equivalent thereafter; Section 3.15)
indicates that a main-effects-only model misspecifies the age–sex relationship for cardiometabolic
outcomes, and that an age × sex interaction term is likely to be warranted. The HighBP–HighChol
joint prevalence of 17.4% (Section 3.18) substantially exceeds the independence expectation of
approximately 14% (= 0.43 × 0.34), consistent with a log odds ratio of approximately 0.55,
indicating meaningful clustering that warrants collinearity checking in any model including both.

## 3.22.4 Variables with Weak Independent Signal

`HvyAlcoholConsump` shows Cramér's V values below 0.10 with nearly all other binary variables
(Section 3.20), a small declining age gradient peaking at around 8% in the youngest groups
(Section 3.10), and minimal sex differences at any age. No meaningful gradient with any health
outcome is apparent from the EDA. It may reasonably be retained as a theoretical covariate but
is unlikely to improve model fit appreciably.

`BMI` shows weak Spearman correlations with `MentHlth` (ρ ≈ 0.05) and near-zero correlation
with `Age` (ρ ≈ −0.02), consistent with limited predictive relevance for mental health models.
Its signal is outcome-specific: the BMI–DiffWalk gradient (Section 3.14) indicates that obese
individuals have approximately three times the difficulty-walking prevalence of normal-weight
individuals, a pattern large enough to translate into a clinically meaningful odds ratio.
Accordingly, BMI is best treated as a primary predictor in functional limitation models and as a
low-priority covariate in models of `MentHlth`.

## 3.22.5 Summary Table of Key EDA Findings

```{r}
#| label: eda-summary-table
#| echo: false

eda_summary <- tibble::tribble(
  ~Finding,                                                      ~Outcome,           ~Implication,
  "Income gradient: ~25pp in high PhysHlth burden",             "PhysHlth",         "Primary predictor; test income × sex interaction",
  "Income gradient: ~20pp in high MentHlth burden",             "MentHlth",         "Primary predictor; larger gap among women at low income",
  "Income gradient: ~40pp in Fair/Poor GenHlth; V ≈ 0.24",     "GenHlth",          "Strongest single predictor; ordered logistic model",
  "MentHlth peaks at 25–39, declines after age 60",             "MentHlth",         "Non-linear age term indicated; avoid raw linear age",
  "PhysHlth and GenHlth rise monotonically with age",           "PhysHlth/GenHlth", "Ordinal age term appropriate; OR ≈ 20 at 80+ for DiffWalk",
  "HighBP/HighChol sex crossover ~age 55–60",                   "HighBP/HighChol",  "Age × sex interaction likely warranted",
  "HighBP–HighChol V = 0.30; joint prev = 17.4% > 14% indep.", "HighBP/HighChol",  "Log OR ≈ 0.55; VIF check required if both in model",
  "PhysActivity × DiffWalk V = 0.25",                           "DiffWalk",         "DiffWalk is a confounder of activity–health associations",
  "High-burden MentHlth ~50% lower in physically active group", "MentHlth",         "PhysActivity: key predictor; adjust for income and age",
  "MentHlth and PhysHlth severely zero-inflated",               "MentHlth/PhysHlth","Hurdle/ZI model or ≥14-day binary outcome; not OLS",
  "HvyAlcohol: V < 0.10 with most variables; flat gradient",    "All",              "Weak signal; retain as theoretical covariate only",
  "BMI: ρ = 0.05 with MentHlth; obese ~3× DiffWalk rate",      "MentHlth/DiffWalk","Outcome-specific role; primary for DiffWalk, weak for MH"
)

knitr::kable(
  eda_summary,
  col.names = c("Key finding", "Primary outcome", "Modelling implication"),
  booktabs  = TRUE,
  longtable = FALSE,
  caption   = "Summary of key EDA findings and their modelling implications."
)
```

These findings collectively define the analytical landscape for Section 4: which predictors
carry the strongest signal for each outcome, where interaction or non-linear terms are
indicated, and which distributional properties preclude standard linear modelling.

---

*Section 4 (Finalised Research Questions) and Section 5 (Modelling) follow.*
