---
title: "Modelling Health Indicators"
subtitle: "MA7442 Modelling Data — Assessment 1 (Task 2)"
author: "Student ID: 259043528"
date: "2026-03-02"
format:
  pdf:
    geometry:
      - top=2cm
      - bottom=2cm
      - left=1.5cm
      - right=1.5cm
    documentclass: article
    df-print: kable
bibliography: references.bib
---

```{r}
#| label: Loading_Packages
#| message: false
#| warning: false
#| include: false

# Core framework
library(tidymodels)
library(tidyverse)

# Data exploration
library(skimr)
library(scales)

# Workflow utilities
library(workflowsets)
library(finetune)
library(themis)
library(probably)
library(conflicted)

# Visualisation
library(viridis)
library(patchwork)
library(vip)

# Model engines
library(ranger)

# Model interpretability
library(shapviz)

# Statistical modelling
library(MASS)
library(pscl)
library(lmtest)
library(marginaleffects)
library(ggeffects)
library(rcompanion)

# Parallel processing
library(doParallel)

# Reporting
library(knitr)
library(kableExtra)

tidymodels_prefer(quiet = TRUE)
conflicts_prefer(yardstick::accuracy)

```

```{r}
#| include: false
# Plot format and dimension
knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(fig.width = 14, fig.height = 7)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Set global ggplot theme
theme_set(
  theme_classic(base_size = 16) +
    theme(
      plot.title = element_text(size = 20, face = "bold"),
      axis.title = element_text(size = 18),
      axis.text  = element_text(size = 14),
      legend.title  = element_text(size = 13),
      legend.text = element_text(size = 12),
      strip.text = element_text(face = "bold", size = 14),
      legend.background = element_rect(
        colour = "black",
        fill = "white",
        linewidth = 0.5
      )
    )
)
```

```{r}
#| label: data-import-and-recode
#| message: false
#| warning: false
#| include: false

diab <- read_csv("dataset1.csv")

diab <- diab |>
  mutate(
    Diabetes = factor(
      Diabetes_012,
      levels = c(0, 1, 2),
      labels = c("No diabetes", "Prediabetes", "Diabetes")
    ),
    across(
      c(CholCheck, Smoker, Stroke, HeartDiseaseorAttack,
        PhysActivity, Fruits, Veggies, HvyAlcoholConsump,
        AnyHealthcare, NoDocbcCost, DiffWalk),
      function(x) factor(x, levels = c(0, 1), labels = c("No", "Yes"))
    ),
    across(
      c(HighBP, HighChol),
      function(x) factor(x, levels = c(0, 1), labels = c("Not High", "High"))
    ),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    GenHlth = factor(
      GenHlth,
      levels  = 1:5,
      labels  = c("Excellent", "Very good", "Good", "Fair", "Poor"),
      ordered = TRUE
    ),
    Age = factor(
      Age,
      levels  = 1:13,
      labels  = c("18–24", "25–29", "30–34", "35–39", "40–44",
                  "45–49", "50–54", "55–59", "60–64", "65–69",
                  "70–74", "75–79", "80+"),
      ordered = TRUE
    ),
    Education = factor(
      Education,
      levels  = 1:6,
      labels  = c("Kindergarten or less", "Grades 1–8",
                  "Grades 9–11", "High school graduate",
                  "Some college / technical", "College graduate"),
      ordered = TRUE
    ),
    Income = factor(
      Income,
      levels  = 1:8,
      labels  = c("< $10k", "$10k–$14k", "$15k–$19k", "$20k–$24k",
                  "$25k–$34k", "$35k–$49k", "$50k–$74k", "$75k+"),
      ordered = TRUE
    )
  ) |>
  select(-Diabetes_012)
```

# 6. Multidomain Determinants and Prediction of Diabetes Status

## 6.1 Modelling Objective

This section addresses the following research question:

> ***To what extent can diabetes status be explained and predicted using demographic, socioeconomic, behavioural, and cardiometabolic factors, and which predictors contribute most strongly to distinguishing between no diabetes, prediabetes, and diabetes?***

The objective is twofold:

1.  **Explanatory** – to estimate the adjusted associations between predictor domains and diabetes status.

2.  **Predictive** – to compare the classification performance of statistical and machine learning models.

Diabetes status is treated as a **multiclass outcome** (No diabetes / Prediabetes / Diabetes), requiring modelling approaches suitable for nominal categorical outcomes.

## 6.2. Outcome Variable

The dependent variable is **diabetes status**, categorised into:

-   **No diabetes**

-   **Prediabetes**

-   **Diabetes.**

Exploratory analysis demonstrated class imbalance, with the majority of respondents in the no-diabetes category and substantially smaller proportions in the prediabetes and diabetes groups. This imbalance is accounted for in model training and evaluation.

## 6.3 Training and Validation

### 6.3.1 **Data Partitioning and Resampling:**

The dataset is partitioned into a 75% training set and a 25% testing set using stratified sampling on diabetes status, ensuring class proportions are preserved across both splits. The raw training set reflects the population-level imbalance of 84.2% No diabetes, 13.9% Diabetes, and 1.8% Prediabetes, a distribution too skewed for stable multiclass learning without intervention.

Prior to cross-validation, a hybrid resampling strategy is applied directly to the training set. No diabetes is downsampled to 68,949 real observations, Diabetes is retained in full (26,519 cases) as the anchor class, and Prediabetes is modestly augmented via SMOTE (over_ratio = 0.1), doubling real cases from 3,485 to 6,894 with approximately 50% synthetic observations. The resulting distribution of 67/26/7 approximates the 65/25/10 target while remaining grounded in observed data. Downsampling uses only real observations to preserve majority-class signal integrity, Diabetes is kept in full to avoid discarding clinically important cases, and SMOTE is limited to a doubling ratio to prevent the model learning predominantly from synthetic profiles. No diabetes intentionally remains the dominant class, reflecting real-world prevalence, while the two minority classes receive sufficient representation for stable parameter estimation.

Five-fold stratified cross-validation is applied to the resampled training set, maintaining class proportions within each fold, and the held-out test set remains untouched throughout development for final evaluation only.

```{r}
#| label: class-weights


class_counts <- diab |>
  count(Diabetes) |>
  mutate(weight = 1 / sqrt(n / sum(n)))

diab <- diab |>
  left_join(class_counts |> select(Diabetes, weight), by = "Diabetes") |>
  mutate(weight = importance_weights(weight)) 
```

```{r}

set.seed(42)
diabetes_split <- initial_split(
  diab,
  prop = 0.75,
  strata = Diabetes
)

diab_train <- training(diabetes_split)
diab_test  <- testing(diabetes_split)

# Sanity checks
diab_train |>  count(Diabetes) |> mutate(prop = n/sum(n))
diab_test |> count(Diabetes) |>  mutate(prop = n/sum(n))

```

```{r}

set.seed(42)
diabetes_folds <- vfold_cv(
  diab_train,
  v = 5,
  strata = Diabetes
)
```

### **6.3.2 Recipe Definition and Preprocessing Strategy**

The preprocessing recipe is defined using the tidymodels `recipes` framework and applied consistently across all candidate models. The following steps are applied in sequence:

-   **Outcome specification** — Models `Diabetes` as a three-level categorical outcome (No diabetes, Prediabetes, Diabetes).

-   **Near-zero variance removal** — Removes predictors with minimal variability across observations to prevent numerical instability and eliminate redundant features that contribute no discriminative information.

-   **Ordinal scoring** — Converts ordered factors (`Age`, `Education`, `Income`, `GenHlth`) into integer scores preserving their natural ranking. This introduces an assumption of approximately equal spacing between adjacent categories; however, the primary objective is to retain monotonic structure in a computationally efficient representation compatible with both parametric and tree-based models.

-   **Dummy encoding** — Transforms remaining nominal predictors into binary indicator variables. This step is necessary for models that cannot natively handle categorical inputs.

-   **Global normalisation** — Standardises all numeric predictors to mean = 0, sd = 1. Unlike the previous specification which normalised only BMI, MentHlth, and PhysHlth, normalisation is applied globally here to accommodate the full candidate model set. SVM and MLP are sensitive to feature scale and require all inputs to be on a comparable range; tree-based models are scale-invariant but are unaffected by this step.

-   **SMOTE augmentation** — Applied within the recipe with `over_ratio = 0.1`, operating on the already-downsampled training data. At this ratio, SMOTE generates synthetic Prediabetes cases by interpolating between existing minority-class nearest neighbours until the class reaches approximately 10% of the majority class count. Because downsampling has already reduced the No diabetes footprint prior to this step, synthetic generation occurs in a feature space with a more balanced neighbourhood structure, reducing the risk of interpolating across majority-class regions.

-   **Case weights** — Inverse square root frequency weights are assigned to each observation based on class membership and incorporated into the recipe via . Weights of approximately 1.09, 2.68, and 7.39 are assigned to No diabetes, Diabetes, and Prediabetes respectively, upweighting minority class observations during parameter estimation without generating additional synthetic data. This complements the resampling strategy by penalising misclassification of rare classes during training, particularly Prediabetes.

```{r}
#| label: recipie-making
set.seed(42)
rec_rq1 <- recipe(Diabetes ~ ., data = diab_train) |>
  step_nzv(all_predictors()) |>
  step_ordinalscore(Age, Education, Income, GenHlth) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_downsample(Diabetes, under_ratio  = 20 , seed = 42) |> 
  step_smote(Diabetes, over_ratio = 0.1, seed = 42)

print(rec_rq1)
```

```{r}
#| label: count-cases

rec_rq1 |> 
  prep() |> 
  bake(new_data = NULL) |> 
  count(Diabetes)
```

### 6.3.3 Model Specification

We now define the candidate models for RQ1 using the **tidymodels** framework. Each model is specified independently of preprocessing and will later be combined with the common recipe using **workflows** and **workflow sets**.

| Model | Type | Primary Function | Key Strengths |
|----|----|----|----|
| **Multinomial Logistic Regression** | Parametric | Linear Baseline | Interpretable log-odds; direct coefficient analysis |
| **Regularised Multinomial (glmnet)** | Parametric (Penalised) | Regularisation Path | Elastic net penalty shrinks redundant predictors; covers ridge, lasso and elastic net via `mixture` tuning |
| **Random Forest** | Ensemble (Bagging) | Non-linear Discovery | Captures complex interactions; robust to multicollinearity |
| **XGBoost** | Ensemble (Boosting) | Predictive Performance | High efficiency on tabular data; optimised for accuracy |
| **MLP** | Neural Network | Non-linear Representation | Learns complex feature interactions via hidden layers; gradient-based optimisation |

Hyperparameters marked `tune()` will be optimised via `tune_race_anova` during cross-validation, with custom space-filling grids specified for RF, XGBoost, MLP and glmnet. Multinomial logistic regression requires no tuning and serves as the unpenalised parametric baseline against which the regularised variant can be directly compared.

```{r}
#| label: model-spec

# Multinomial logistic regression
mod_multinom <-
  multinom_reg() |>
  set_engine("nnet") |>
  set_mode("classification")

# Regularised Multinomial (glmnet)
mod_glmnet <-
  multinom_reg(
    penalty = tune(),
    mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

# Random Forest
mod_rf <-
  rand_forest(
    trees = 512,
    mtry  = tune(),
    min_n = tune()) |>
  set_engine("ranger") |>
  set_mode("classification")

# XGBoost
mod_xgb <-
  boost_tree(
    trees = 1024,
    tree_depth = tune(),
    learn_rate = tune(),
    loss_reduction = tune(),
    sample_size = tune(),
    mtry  = tune(),
    min_n = tune()) |>
  set_engine("xgboost") |>
  set_mode("classification")

# MLP
mod_mlp <-
  mlp(
    hidden_units = tune(),
    penalty = tune(),
    epochs  = tune()) |>
  set_engine("nnet") |>
  set_mode("classification")

```

### 6.3.4 Workflow Set Construction

We now integrate the RQ1 preprocessing recipe with all five candidate model specifications. Each workflow represents one model-preprocessing pair, and the workflow set enables parallel evaluation under identical resampling conditions. Custom tuning grids are attached via `option_add()` for Regularized GLMnet, XGBoost, Random Forest and MLP, overriding the default space-filling design for those models with constrained parameter ranges appropriate to their respective architectures.

```{r}
#| label: wrkflw-sets

# Workflow set 
wf_set_rq1 <-
  workflow_set(
    preproc = list(rq1_recipe = rec_rq1),
    models  = list(
      multinom = mod_multinom,
      glmnet = mod_glmnet,
      rf  = mod_rf,
      xgb = mod_xgb,
      mlp = mod_mlp))

set.seed(42)

# Tuning grids
grid_rf <- grid_space_filling(
  mtry(range  = c(2L, 15L)),
  min_n(range = c(5L, 30L)),
  size = 30)

grid_xgb <- grid_space_filling(
  tree_depth(range = c(2L, 8L)),
  learn_rate(range = c(-2.5, -1)),
  loss_reduction(range = c(-5, 1)),
  sample_size = sample_prop(range = c(0.5, 1.0)),
  mtry(range = c(2L, 15L)),
  min_n(range = c(5L, 25L)),
  size = 50)

grid_mlp <- grid_space_filling(
  hidden_units(range = c(32L, 96L)),
  penalty(range = c(-4, -2)),
  epochs(range = c(100L, 200L)),
  size = 40)

grid_glmnet <- grid_space_filling(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  size = 30)

wf_set_rq1 <-
  wf_set_rq1 |> 
  option_add(id = "rq1_recipe_glmnet", grid = grid_glmnet) |> 
  option_add(id = "rq1_recipe_rf", grid = grid_rf) |>
  option_add(id = "rq1_recipe_xgb", grid = grid_xgb) |>
  option_add(id = "rq1_recipe_mlp", grid = grid_mlp) 
```

### 6.3.5 Hyperparameter Tuning

**Metrics computed**

Model performance is evaluated using four complementary metrics selected to reflect both discrimination ability and calibration quality under residual class imbalance:

-   **Multinomial log-loss** — the primary tuning metric, measuring the calibration of predicted class probabilities across all three diabetes categories. Unlike threshold-dependent metrics, log-loss penalises overconfident incorrect predictions directly, making it the most informative criterion for probabilistic multiclass models and the metric used to drive early elimination in `tune_race_anova`.

-   **ROC-AUC (multiclass)** — measures discrimination ability across all possible classification thresholds using the one-vs-rest framework, providing a threshold-independent summary of how well each model separates the three classes.

-   **Balanced accuracy** — mean recall across classes, correcting for residual imbalance in the resampled training distribution. Retained as the primary communication metric given its interpretability to non-technical audiences.

-   **Macro F1-score** — harmonic mean of precision and recall averaged equally across all three classes, providing an imbalance-aware summary of classification performance as a supporting metric.

```{r}
#| label: metric-set
metrics_diab <-
  metric_set(
    mn_log_loss,
    roc_auc,
    bal_accuracy,
    f_meas
  )
```

**Hyperparameter Tuning Strategy**

Hyperparameter optimisation is performed using `tune_race_anova`, which eliminates underperforming configurations early based on interim fold results, enabling efficient exploration of larger grid sizes without fitting all configurations to completion:

-   **Grid size** — 30-50 configurations per model, with space-filling designs via `grid_space_filling()` for MLP and RBF SVM to constrain the search to empirically sensible parameter regions. Multinomial logistic regression requires no tuning. The total maximum fit count is 5 models × 15 grids × 5 CV folds = 375 fits, reduced in practice by early elimination.

-   **Search method** — `tune_race_anova` uses ANOVA-based interim testing across folds to identify and discard configurations that are statistically unlikely to be competitive, substantially reducing computation relative to full grid search while retaining systematic coverage of the parameter space.

-   **Tuning metric** — multinomial log-loss serves as the primary optimisation metric, driving both configuration selection and early elimination decisions. Log-loss rewards well-calibrated probability estimates across all three classes rather than optimising threshold-dependent classification boundaries.

-   **Cross-validation** — 5-fold stratified CV applied to the resampled training set, maintaining the 67/26/7 class distribution within each fold to ensure consistent learning conditions across splits.

The best hyperparameter configuration for each model is selected based on lowest mean multinomial log-loss across folds.

**ANOVA Race Control Parameters**

```{r}
#| label: race-control

ctrl <- control_race(
  save_pred     = TRUE,
  save_workflow = FALSE,
  verbose_elim  = TRUE,
  verbose = TRUE,
  burn_in = 3,
  alpha = 0.05,
  randomize = TRUE,
  allow_par = TRUE,
  extract = NULL
)

cl <- makePSOCKcluster(4)
registerDoParallel(cl)
```

```{r}
#| label: hp-tune

# cat("Started:", format(Sys.time()), "\n")
# set.seed(42)
# wf_set_rq1_fit <-
#   wf_set_rq1 |>
#   workflow_map(
#     "tune_race_anova",
#     resamples = diabetes_folds,
#     metrics   = metrics_diab,
#     control   = ctrl
#   )
# 
# cat("Finished:", format(Sys.time()), "\n")
# stopCluster(cl)
# registerDoSEQ()
# gc()
# 
# write_rds(wf_set_rq1_fit, 'wf_set_rq1_fit.rds')

wf_set_rq1_fit <- readRDS('wf_set_rq1_fit.rds')
```

```{r}
#| include: false

pretty_autoplot <- function(x, metric, ylab, title, pct = TRUE, x_breaks = 6,
                            labels = NULL) {
  p <- autoplot(x, metric = metric) +
    aes(colour = wflow_id) +
    labs(
      title  = title,
      x      = "Workflow rank (best → worst)",
      y      = ylab,
      colour = "Model"
    ) +
    scale_x_continuous(breaks = pretty_breaks(x_breaks)) +
    scale_y_continuous(
      labels = if (pct) percent_format(accuracy = 0.1)
               else number_format(accuracy = 0.001)) +
    theme(legend.position = "bottom")
  
  if (!is.null(labels)) {
    p <- p + scale_colour_discrete(labels = labels)
  }
  
  p
}
```

```{r}
#| label: eval-results
#| fig-height: 10
#| fig-width: 16

model_labels <- c(
  "rq1_recipe_glmnet"  = "Elastic Net",
  "rq1_recipe_mlp" = "MLP",
  "rq1_recipe_multinom" = "Multinomial Log Regression",
  "rq1_recipe_rf" = "Random Forest",
  "rq1_recipe_xgb" = "XGBoost"
)

m1 <- pretty_autoplot(
  wf_set_rq1_fit,
  "roc_auc",
  "ROC–AUC (multiclass)",
  "Discrimination (ROC–AUC)",
  labels = model_labels)

m2 <- pretty_autoplot(
  wf_set_rq1_fit,
  "bal_accuracy",
  "Balanced accuracy",
  "Class-balanced accuracy",
  labels = model_labels)

m3 <- pretty_autoplot(
  wf_set_rq1_fit,
  "f_meas",
  "Macro F1-score",
  "Macro F1 (imbalance-aware)",
  labels = model_labels)

m4 <- pretty_autoplot(
  wf_set_rq1_fit,
  "mn_log_loss",
  "Multinomial log-loss",
  "Calibration (log-loss)",
  pct    = FALSE,
  labels = model_labels)

(m1 | m2) / (m3 | m4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: rankings-rq1-table
#| message: false
#| warning: false
wf_set_rq1_fit |>
  collect_metrics() |>
  filter(.metric %in% c("roc_auc", "bal_accuracy", "f_meas", "mn_log_loss")) |>
  group_by(wflow_id, .metric) |>
  summarise(mean_val = mean(mean, na.rm = TRUE)) |>
  arrange(.metric, desc(mean_val))
```

**ROC-AUC** XGBoost achieves the best CV discrimination at approximately 70.7%, with Random Forest close behind at 70.3%. Multinomial, MLP, and Elastic Net cluster tightly between 70.0% and 70.1%, indicating near-equivalent discriminative ability across parametric and neural network approaches. Overall differences span less than one percentage point across all workflows, suggesting model family has limited impact on raw discrimination under the resampled training distribution.

**Balanced Accuracy** XGBoost leads at 62.7%, with MLP close behind at 62.6%, the two are effectively tied. Random Forest sits in the middle at 62.1%, while Elastic Net and Multinomial occupy the lower range at 61.5% and 61.6% respectively. The parametric models show a consistent deficit relative to the ensemble approaches across this metric.

**Macro F1** The ranking inverts notably here. Random Forest achieves the highest macro F1 at 67.6%, followed by Multinomial and Elastic Net at 66.8% and 66.7% respectively. XGBoost drops to fourth at 58.8%, and MLP performs poorly at 45.3%, indicating that despite reasonable discrimination and balanced accuracy, both models struggle with class-balanced precision-recall tradeoffs under the true class distribution. This divergence between ROC-AUC and macro F1 warrants attention and suggests the both models are optimising discrimination at the expense of minority class precision.

**Calibration (Log-loss)** This panel shows the starkest separation. XGBoost achieves the best calibration at 0.437, followed by Random Forest at 0.446. Elastic Net and Multinomial perform nearly identically at approximately 0.463, confirming that ridge regularisation adds minimal calibration benefit over the unpenalised baseline. MLP is clearly the worst calibrated by a substantial margin at 0.785, indicating that despite reasonable discrimination and balanced accuracy, MLP produces systematically overconfident class probability estimates across cross-validation folds.

### 6.3.6 Extract tuning results

Cross-validation performance provides a useful guide for hyperparameter selection and relative model ranking, but it reflects the resampled training distribution rather than the population-level class structure present in the held-out test set. Final conclusions about model performance, calibration, and minority class recovery cannot be drawn until each model is evaluated on unseen data. The best configuration for each tuned model is therefore selected based on lowest mean multinomial log-loss across folds, finalized, and refitted on the full training set before evaluation on the held-out test set.

```{r}
#| label: extarct

rf_res  <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_res <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_xgb")
mlp_res <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_mlp")
glmnet_res <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_glmnet")
# Note: multinomial has no tuning parameters, so no extraction needed

```

**Best Hyperparameters Per Model**

```{r}
#| label: rq1-hyperparameter-tables

hp_result <- bind_rows(
  rf_res |> select_best(metric = "mn_log_loss") |> 
    mutate(Model = "Random Forest"),
  xgb_res |> select_best(metric = "mn_log_loss") |> 
    mutate(Model = "XGBoost"),
  mlp_res |> select_best(metric = "mn_log_loss") |> 
    mutate(Model = "MLP"),
  glmnet_res |> select_best(metric = "mn_log_loss") |> 
    mutate(Model = "Elastic Net")
  ) |>
  select(Model, mtry, min_n, tree_depth, learn_rate, loss_reduction,
         hidden_units, penalty, mixture, epochs)
```

```{r}
#| echo: false
hp_result |>
  kable(digits  = 4,
    caption = "Optimal hyperparameters per model selected by minimum mean log-loss across 5-fold CV. Empty cells indicate parameters not applicable to the given model.",
    na = ""
  ) |>
  kable_styling(full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

```{r}
#| label: select-best-models
# Using mn_log_loss as the primary optimization metric 
best_rf <- select_best(rf_res, metric = "mn_log_loss")
best_xgb <- select_best(xgb_res, metric = "mn_log_loss")
best_mlp <- select_best(mlp_res, metric = "mn_log_loss")
best_glmnet <- select_best(glmnet_res, metric = "mn_log_loss")
```

```{r}
#| label: final-wrkflows

# These contain the recipe and model specification without fitted parameters
rf_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_xgb")
mlp_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_mlp")
glmnet_wf  <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_glmnet")
multinom_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_multinom")


# This creates finalized workflows ready for last-fit on full dataset
rf_final_wf <- finalize_workflow(rf_wf, best_rf)
xgb_final_wf <- finalize_workflow(xgb_wf, best_xgb)
mlp_final_wf <- finalize_workflow(mlp_wf, best_mlp)
glmnet_final_wf  <- finalize_workflow(glmnet_wf, best_glmnet)
# Multinomial has no tuning parameters, so original workflow is already finalized
multinom_final_wf <- multinom_wf

```

```{r}
#| label: remove-the wrkflw-set
#| include: false

rm(wf_set_rq1_fit)
gc()
```

```{r}
#| label: final-fit

glmnet_final_fit <- 
  glmnet_final_wf  |> 
  last_fit(diabetes_split)

multinom_final_fit <- 
  multinom_final_wf |> 
  last_fit(diabetes_split)

rf_final_fit  <-
  rf_final_wf  |>
  last_fit(diabetes_split)

xgb_final_fit <-
  xgb_final_wf |>
  last_fit(diabetes_split)

mlp_final_fit <-
  mlp_final_wf |>
  last_fit(diabetes_split)
```

## 6.4 Model Evaluation

```{r}
#| label: rq1-test-preds
pred_rf <- rf_final_fit |>
  collect_predictions() |>
  mutate(model = "Random Forest")

pred_xgb  <- xgb_final_fit |>
  collect_predictions() |>
  mutate(model = "XGBoost")

pred_mlp  <- mlp_final_fit  |>
  collect_predictions() |>
  mutate(model = "MLP")

pred_glmnet <- glmnet_final_fit |>
  collect_predictions() |> 
  mutate(model = "Elastic Net")

pred_mn <- multinom_final_fit |> 
  collect_predictions() |>
  mutate(model = "Multinomial")

pred_all <- bind_rows(pred_rf, pred_xgb, pred_mlp, pred_glmnet, pred_mn) |>
  rename(.pred_No_diabetes = `.pred_No diabetes`)
```

### 6.4.1 ROC Curves

**Multiclass ROC-AUC Evaluation**

ROC (Receiver Operating Characteristic) curves for multiclass outcomes require extending the binary framework. We employ the **one-vs-rest approach**:

-   Each class (No diabetes, Prediabetes, Diabetes) is treated as a separate binary problem against all other classes combined
-   For each class, ROC curves plot True Positive Rate vs False Positive Rate across probability thresholds
-   **Area Under Curve (AUC)** summarizes discriminative ability: AUC = 0.5 indicates random guessing, AUC = 1.0 indicates perfect separation
-   Despite class imbalance, ROC-AUC remains informative for comparing relative model performance, though it may be optimistic for rare classes (Prediabetes, Diabetes)

Higher AUC values indicate better ability to distinguish each diabetes category from alternatives. Models with consistently high AUC across all three classes demonstrate robust multiclass discrimination.

```{r}
#| label: roc-auc-table

roc_auc_table <- pred_all |>
  group_by(model) |>
  roc_auc(
    truth = Diabetes,
    .pred_No_diabetes,
    .pred_Prediabetes,
    .pred_Diabetes,
    estimator = "macro_weighted") |>
  select(model, .estimate) |>
  arrange(desc(.estimate)) |>
  rename(
    Model   = model,
    `ROC-AUC` = .estimate)

roc_auc_table 
```

> ROC-AUC values are closely clustered across all five models, spanning less than two percentage points. Differences of this magnitude are unlikely to be practically meaningful, and if discrimination alone were the selection criterion, the unpenalised multinomial logistic regression would be preferred on grounds of computational simplicity and interpretability. However, ROC-AUC is threshold-independent and does not reflect class-specific recovery or probability calibration, motivating the broader evaluation that follows.

```{r}
#| label: rq1-roc-combined
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 12


roc_all <- pred_all |>
  group_by(model) |>
  roc_curve(
    truth = Diabetes,
    .pred_No_diabetes,
    .pred_Prediabetes,
    .pred_Diabetes)

autoplot(roc_all) +
  labs(title = "Multiclass ROC curves (one-vs-rest) — Test set",
       x = "1 − Specificity",y = "Sensitivity") +
  theme(legend.position = "bottom")
```

The multiclass ROC curves provide additional insight into discrimination performance across the three classes.

For the **Prediabetes** panel, all models perform notably weaker than the other two classes, with curves closer to the diagonal reflecting the inherent difficulty of distinguishing this intermediate category. For **Diabetes** and **No diabetes**, all five models show strong and near-identical discrimination, with curves well above the diagonal and minimal separation between model types, consistent with the marginal AUC differences reported in the table above.

### 6.4.2 Confusion Matrices

```{r}
#| fig-height: 10
#| fig-width: 16
#| label: confusion-matrices

plot_cm <- function(pred, title) {
  pred |>
    conf_mat(truth = Diabetes, estimate = .pred_class) |>
    autoplot(type = "heatmap") +
    labs(title = title) +
    theme(legend.position = "none")}

p_cm_xgb <- plot_cm(pred_xgb, "XGBoost")
p_cm_rf <- plot_cm(pred_rf,"Random Forest")
p_cm_mn <- plot_cm(pred_mn, "Multinomial Logistic Regression")
p_cm_mlp  <- plot_cm(pred_mlp, "MLP")
p_cm_glmnet  <- plot_cm(pred_glmnet, "Elastic Net")

(p_cm_xgb | p_cm_rf | p_cm_mlp) /
(p_cm_mn | p_cm_glmnet) +
  plot_layout(guides = "collect")
```

All five models achieve high correct classification rates for the No diabetes class, reflecting its 84% prevalence dominance in the test set. For the Diabetes class, all models correctly identify several thousand cases, though misclassification between Diabetes and No diabetes persists throughout.

The critical finding is Prediabetes classification. Four of the five models, XGBoost, Random Forest, Multinomial Logistic Regression, and Elastic Net, predict zero Prediabetes cases, effectively collapsing the problem into a binary task. MLP is the sole exception, producing a marginal but non-zero number of Prediabetes predictions, suggesting its probability estimates are structurally more dispersed than the tree-based and parametric alternatives. This collapse is not attributable to modelling failure but to the structural near-inseparability of Prediabetes at population prevalence, as documented in Section 6.5.

### 6.4.3 Class-wise Performance

While ROC-AUC provides an aggregate discrimination summary, it obscures class-specific classification behaviour. The heatmap below disaggregates precision and recall across all three outcome classes and five models, directly exposing where each model succeeds and fails at the individual class level.

```{r}
#| label: classwise-prep
#| include: false
make_binary <- function(x, pos) {
  factor(ifelse(x == pos, pos, "Other"), levels = c(pos, "Other"))
}

classwise_metrics <- pred_all |>
  group_by(model) |>
  summarise(
    precision_no   = precision_vec(Diabetes, .pred_class, event_level = "first"),
    recall_no = recall_vec(Diabetes, .pred_class, event_level = "first"),
    precision_pre  = precision_vec(make_binary(Diabetes, "Prediabetes"),
                                   make_binary(.pred_class, "Prediabetes")),
    recall_pre = recall_vec(make_binary(Diabetes, "Prediabetes"),
                                make_binary(.pred_class, "Prediabetes")),
    precision_diab = precision_vec(make_binary(Diabetes, "Diabetes"),
                                   make_binary(.pred_class, "Diabetes")),
    recall_diab = recall_vec(make_binary(Diabetes, "Diabetes"),
                                make_binary(.pred_class, "Diabetes"))
  ) |>
  pivot_longer(
    cols      = -model,
    names_to  = c("metric", "class"),
    names_sep = "_",
    values_to = "value"
  ) |>
  mutate(
    class  = case_when(
      class == "no"   ~ "No diabetes",
      class == "pre"  ~ "Prediabetes",
      class == "diab" ~ "Diabetes"),
    metric = str_to_title(metric))
```

```{r}
#| label: rq1-classwise-heatmap
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 12

classwise_metrics |>
  ggplot(aes(x = class, y = model, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = percent(value, accuracy = 0.1)),
            size = 3, colour = "white") +
  facet_wrap(~ metric) +
  scale_fill_viridis_c(
    option  = "D",
    limits  = c(0, 1),
    labels  = percent_format()) +
  labs(
    title = "Class-wise precision and recall by model",
    x     = NULL,
    y     = NULL,
    fill  = NULL
  ) +
  theme(
    legend.position = "bottom")
```

The heatmap exposes the binary collapse documented in Section 6.4.2 in quantitative terms. Prediabetes precision is undefined (grey) for all models except MLP, which achieves 0.0%, confirming that no model produces a single correctly classified Prediabetes case with meaningful confidence. Prediabetes recall is 0.0% across all five models without exception.

For No diabetes, precision is consistently strong across all models at 65-66%, with the exception of MLP which drops to 44.1%, reflecting its tendency to misclassify No diabetes observations as Prediabetes. For the Diabetes class, precision sits at 41-43% across all models, moderate but consistent, with recall in the 44-50% range, indicating that roughly half of true Diabetes cases are missed and misclassified as No diabetes.

The overall picture confirms that all five models effectively operate as binary classifiers on this test set, distinguishing No diabetes from Diabetes with moderate success while failing entirely on the intermediate Prediabetes class.

## 6.5 Threshold Adjustment for Prediabetes Recovery

The confusion matrices presented in Section 6.4.2 revealed a consistent failure across all five models to predict Prediabetes at the default 0.5 classification threshold. As established in Section 6.3.1, the held-out test set reflects the true population distribution of 1.8% Prediabetes, a prevalence so low that posterior class probabilities rarely exceed the default boundary. The ROC-AUC results of approximately 0.82 across all models confirm that genuine discriminative signal exists, the models are producing meaningful Prediabetes probability estimates, but these are structurally suppressed below the decision threshold. This section investigates whether threshold adjustment can recover Prediabetes predictions and, if not, characterises the nature of the separability failure.

### 6.5.1 Calibration Plot

```{r}
#| label: calibration-plot
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 14

cal_plot_data <- pred_all |>
  mutate(Diabetes = factor(Diabetes)) |>
  group_by(model)

cal_plot <- function(pred_data, model_name) {
  pred_data |>
    filter(model == model_name) |>
    cal_plot_windowed(
      truth      = Diabetes,
      estimate   = c(.pred_No_diabetes, .pred_Prediabetes, .pred_Diabetes),
      step_size  = 0.05
    ) +
    labs(title = model_name) +
    theme_classic() +
    theme(legend.position = "none")
}

p_cal_xgb    <- cal_plot(pred_all, "XGBoost")
p_cal_rf     <- cal_plot(pred_all, "Random Forest")
p_cal_mn     <- cal_plot(pred_all, "Multinomial")
p_cal_mlp    <- cal_plot(pred_all, "MLP")
p_cal_glmnet <- cal_plot(pred_all, "Elastic Net")

(p_cal_xgb | p_cal_rf | p_cal_mn) /
(p_cal_mlp | p_cal_glmnet) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**All models are severely overconfident for Prediabetes**: the curves sit well above the diagonal across the board, meaning when a model predicts 20-30% Prediabetes probability, the actual event rate is much higher (\~60-90%). This is the clearest possible confirmation that the default 0.5 threshold is too high,the models are generating useful Prediabetes signal at much lower probability values than 0.5.

**XGBoost:** Shows the most erratic calibration with a sharp drop around 0.25–0.30, suggesting its Prediabetes probabilities are concentrated in a narrow range with unstable behaviour at the boundary.

**Random Forest:** Shows a smooth but steep decline, with reasonably consistent overconfidence across the probability range.

**Multinomial and Elastic Net:** Nearly identical in behaviour, both showing high event rates at low predicted probabilities. This confirms that Ridge regularisation did not meaningfully alter calibration behaviour relative to the unpenalised baseline.

**MLP:** Shows the widest confidence intervals, reflecting instability consistent with its poor log-loss performance in Section 6.4.

The key takeaway is that all models are producing genuine Prediabetes signal at probabilities well below 0.5, warranting further investigation of the probability distributions before any threshold decision is made.

### 6.5.2 Predicted Probability Distribution

Before selecting adjusted thresholds, the distribution of predicted Prediabetes probabilities is examined across true Prediabetes cases only, to identify where signal is concentrated and confirm that meaningful probability mass exists below the default threshold.

```{r}
#| label: prob-distribution
#| warning: false
#| fig-height: 9
#| fig-width: 15

pred_all |>
  filter(Diabetes == "Prediabetes") |>
  ggplot(aes(x = .pred_Prediabetes, fill = model)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  facet_wrap(~ model, ncol = 2) +
  scale_x_continuous(
    limits = c(0, 0.5),
    labels = percent_format(accuracy = 1)) +
  scale_fill_viridis_d() +
  labs(
    title = "Predicted Prediabetes probability — true Prediabetes cases only",
    x     = "P(Prediabetes)",
    y     = "Count"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

> The distributions confirm that genuine Prediabetes signal exists across all models but is concentrated at probability values well below the default 0.5 threshold. For tree-based and parametric models, the required threshold to recover any meaningful Prediabetes predictions would fall in the 5–10% range, a region so low that it overlaps heavily with the probability mass of No diabetes and Diabetes observations. For MLP, the higher concentration around 25–30% is more promising in isolation, but the earlier calibration analysis demonstrated that even this range is characterised by wide uncertainty and unstable boundaries. Lowering thresholds to these levels would therefore generate substantial misclassification of the majority classes, trading a marginal gain in Prediabetes recall for a large increase in false positives across the other two categories.

### 6.5.3 Empirical Threshold Analysis and Separability Conclusion

To confirm this, empirical threshold optimisation via precision-recall F1 maximisation is applied to each model. The results are unambiguous.

```{r}
#| label: threshold-selection

find_threshold <- function(pred_data, model_name) {
  pred_data |>
    filter(model == model_name) |>
    mutate(Prediabetes_truth = factor(
      ifelse(Diabetes == "Prediabetes", "Prediabetes", "Other"),
      levels = c("Prediabetes", "Other"))) |>
    pr_curve(
      truth    = Prediabetes_truth,
      .pred_Prediabetes,
      event_level = "first") |>
    mutate(
      f1 = 2 * precision * recall / (precision + recall),
      model = model_name) |>
    slice_max(f1, n = 1)
}

bind_rows(
  find_threshold(pred_all, "XGBoost"),
  find_threshold(pred_all, "Random Forest"),
  find_threshold(pred_all, "Multinomial"),
  find_threshold(pred_all, "Elastic Net"),
  find_threshold(pred_all, "MLP")) |>
  select(model, .threshold, precision, recall, f1)


```

Precision values remain below 5% across all five models at their respective optimal thresholds. For every true Prediabetes case correctly identified, approximately 20 non-Prediabetes observations would be misclassified as Prediabetes. No operationally useful classification boundary exists within the observable probability range for any model, and threshold adjustment is implemented in Section 6.5.4 to quantify the precision-recall tradeoff explicitly.

Taken together, the calibration plots, probability distributions, and threshold analysis converge on a single conclusion. Prediabetes represents a structurally near-inseparable class in this feature space at population prevalence. Its clinical profile overlaps substantially with both adjacent categories, and the available self-reported survey predictors lack the granularity required to isolate the transitionary prediabetic phenotype at the decision boundary. This is consistent with broader evidence that prediabetes detection from survey-level data remains a fundamental challenge even under optimal modelling conditions, and points toward the need for continuous biomarker data such as HbA1c or fasting glucose to achieve clinically meaningful Prediabetes classification.

### 6.5.4 Adjusted Class-wise Performance

```{r}
#| label: rq1-adjusted-heatmap-prep
#| include: false

thresholds_vec <- c(
  "XGBoost"       = 0.0643,
  "Random Forest" = 0.0696,
  "Multinomial"   = 0.1058,
  "Elastic Net"   = 0.1028,
  "MLP"           = 0.2615)

pred_adjusted <- pred_all |>
  mutate(.pred_class_adj = factor(
    ifelse(.pred_Prediabetes > thresholds_vec[model],
           "Prediabetes",
           as.character(.pred_class)),
    levels = c("No diabetes", "Prediabetes", "Diabetes")))

classwise_adj <- pred_adjusted |>
  group_by(model) |>
  summarise(
    precision_no   = precision_vec(Diabetes, .pred_class_adj, event_level = "first"),
    recall_no      = recall_vec(Diabetes, .pred_class_adj, event_level = "first"),
    precision_pre  = precision_vec(make_binary(Diabetes, "Prediabetes"),
                                   make_binary(.pred_class_adj, "Prediabetes")),
    recall_pre     = recall_vec(make_binary(Diabetes, "Prediabetes"),
                                make_binary(.pred_class_adj, "Prediabetes")),
    precision_diab = precision_vec(make_binary(Diabetes, "Diabetes"),
                                   make_binary(.pred_class_adj, "Diabetes")),
    recall_diab    = recall_vec(make_binary(Diabetes, "Diabetes"),
                                make_binary(.pred_class_adj, "Diabetes"))
  ) |>
  pivot_longer(
    cols      = -model,
    names_to  = c("metric", "class"),
    names_sep = "_",
    values_to = "value"
  ) |>
  mutate(
    class  = case_when(
      class == "no"   ~ "No diabetes",
      class == "pre"  ~ "Prediabetes",
      class == "diab" ~ "Diabetes"),
    metric = str_to_title(metric))
```

```{r}
#| label: rq1-adjusted-heatmap

classwise_adj |>
  ggplot(aes(x = class, y = model, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = percent(value, accuracy = 0.1)),
            size = 3, colour = "white") +
  facet_wrap(~ metric) +
  scale_fill_viridis_c(
    option  = "D",
    limits  = c(0, 1),
    labels  = percent_format()) +
  labs(
    title = "Class-wise precision and recall by model — adjusted threshold",
    x     = NULL,
    y     = NULL,
    fill  = "Score"
  ) +
  theme(
    legend.position = "bottom")
```

Applying the empirically derived thresholds produces a clear and consistent tradeoff across all five models. Prediabetes recall improves meaningfully, reaching 22-39% across models compared to 0% at the default threshold, with Random Forest achieving the highest recovery at 38.8% and Elastic Net at 29.0%. However Prediabetes precision remains critically low at 3.4-4.3%, confirming the finding from Section 6.5.3 that for every true Prediabetes case identified, approximately 20-30 non-Prediabetes observations are misclassified alongside it.

The cost of this recovery is substantial. Diabetes recall drops sharply across all models, falling from 44-50% at the default threshold to 23-43%, as many true Diabetes observations are reclassified into the newly expanded Prediabetes prediction region. No diabetes metrics remain broadly stable, reflecting its dominance in the test set.

The adjusted heatmap confirms that no operationally useful threshold exists for Prediabetes classification in this feature space. Any meaningful gain in Prediabetes recall comes at the direct expense of Diabetes recall, trading one classification failure for another. This reinforces the separability conclusion in Section 6.5.3 and rules out threshold adjustment as a viable remediation strategy.

## 6.6 Model Interpretation

**Triangulation Strategy for Model Interpretation**

To ensure robust identification of diabetes risk factors, **methodological triangulation** is employed by comparing interpretations across different analytical approaches:

-   **Multinomial logistic regression (Log-Odds):** Provides parametric coefficient estimates assuming linear additive effects. Interpretable but limited to linear relationships
-   **SHAP values (XGBoost):** Model-agnostic feature attribution capturing non-linear effects and interactions through tree-based partitioning
-   **Variable Importance (XGBoost gain):** Measures total loss reduction when features are used for splits, reflecting predictive utility

**Convergence between methods** strengthens confidence that identified risk factors are robust across modelling assumptions. **Divergence** indicates non-linearities, interactions, or violations of parametric assumptions that only flexible models capture. Where findings conflict, consistency patterns are prioritized and discrepancies are noted as areas requiring further investigation.

### 6.6.1 Log-Odds Ratios (Multinomial Logistic Regression)

Multinomial logistic regression estimates the log-odds of belonging to each diabetes category relative to a reference category. In this analysis, **No diabetes** serves as the reference class.

Positive coefficients indicate **higher odds of belonging to the specified diabetes category**, while negative coefficients indicate **lower odds relative to the reference group**.

```{r}
#| label: rq1-logodds-prep
#| include: false

multinom_coefs <- multinom_final_fit |>
  extract_fit_parsnip() |>
  pluck("fit") |>
  coef()

multinom_coefs_tidy <- multinom_coefs |>
  as.data.frame() |>
  rownames_to_column(var = "y.level") |>
  pivot_longer(
    cols      = -y.level,
    names_to  = "term",
    values_to = "estimate"
  ) |>
  mutate(
    odds_ratio = exp(estimate),
    direction  = ifelse(odds_ratio > 1, "Increased risk", "Decreased risk")
  )

```

```{r}
#| label: rq1-logodds-plot
#| fig-height: 8
#| fig-width: 16


multinom_coefs_tidy |>
  filter(term != "(Intercept)") |>
  ggplot(aes(
    x    = odds_ratio,
    y    = reorder(term, odds_ratio),
    colour = direction)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ y.level) +
  scale_colour_manual(values = c(
    "Increased risk" = "#E69F00",
    "Decreased risk" = "#0072B2")) +
  labs(
    x      = "Odds Ratio",
    y      = "Predictor",
    title  = "Odds ratios for predictors of diabetes status",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

**Multinomial Logistic Regression: Odds Ratio (OR) Summary**

*OR \> 1 indicates increased odds; OR \< 1 indicates decreased odds.*

-   **Primary drivers.** General health, BMI, and age are the three strongest positive predictors of Diabetes ($OR = 1.83$, $OR = 1.61$, $OR = 1.57$ respectively), with clinical metabolic markers close behind. High blood pressure yields a 47% increase in odds of Diabetes ($OR = 1.47$) and high cholesterol a 35% increase ($OR = 1.35$).

-   **The BMI gradient.** BMI shows a clear disease progression pattern, with a stronger association for Diabetes ($OR = 1.61$) than Prediabetes ($OR = 1.44$), confirming its role as a marker of disease advancement rather than uniform transitional risk.

-   **Behavioural mediation.** Physical activity, fruit consumption, and vegetable consumption all show near-null effects ($OR \approx 0.98$ across both classes), consistent with full mediation through metabolic intermediates such as BMI and blood pressure.

-   **Healthcare access.** NoDocbcCost shows a stronger association with Prediabetes ($OR = 1.11$) than Diabetes ($OR = 1.00$), suggesting cost barriers to care are more strongly linked to the undiagnosed transitionary state than to established diabetes.

-   **Statistical anomalies.** Heavy alcohol consumption shows a negative association ($OR = 0.85$ for Diabetes, $OR = 0.96$ for Prediabetes), almost certainly reflecting selection bias from sick quitter effects or post-diagnosis behavioural adjustment rather than a genuine protective mechanism.

**Conclusion**

Diabetes status in this population is best characterized by a **multifactorial cardiometabolic profile**. While age and sex are relevant, clinical markers of obesity and metabolic syndrome remain the most effective predictors for distinguishing between healthy, prediabetic, and diabetic individuals.

### 6.6.2 Elastic Net Coefficient Path

The optimal elastic net configuration converged to pure ridge regularisation (mixture = 0.0, as established in section 6.4), applying continuous shrinkage to all predictor coefficients toward zero without eliminating any predictor entirely. Unlike lasso, ridge retains the full feature set, meaning coefficient magnitude reflects relative predictive contribution under regularisation pressure rather than binary inclusion or exclusion. This provides a useful complement to the multinomial log-odds, revealing how the full predictor set is weighted when multicollinearity and overfitting are penalised simultaneously.

```{r}
#| label: rq1-ridge-coefs
#| message: false
#| warning: false

glmnet_coefs <- glmnet_final_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)", estimate != 0) |>
  mutate(direction = ifelse(estimate > 0, "Retained positive", "Retained negative"))

glmnet_coefs |>
  ggplot(aes(x = estimate, y = reorder(term, estimate), fill = direction)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~ class) +
  scale_fill_manual(values = c("Retained positive" = "#E69F00", "Retained negative" = "#0072B2")) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey20") +
  labs(
    title = "Elastic Net retained predictors by outcome class",
    subtitle = "Pure Ridge (mixture = 0). Zero-coefficient predictors excluded.",
    x    = "Coefficient estimate",
    y    = "Predictor",
    fill = NULL
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

The coefficient directions are broadly consistent with the multinomial log-odds in Section 6.6.1, providing cross-method validation of the core risk structure.

**GenHlth and BMI** are the two strongest retained predictors for Diabetes with the largest positive coefficients, confirming their dominance established in the multinomial analysis. Age and HighBP follow closely, consistent with the cardiometabolic risk profile identified earlier.

**No diabetes coefficients** are largely mirror images of the Diabetes coefficients, with GenHlth, Age, and BMI showing the strongest negative estimates, reflecting that the same predictors driving diabetes risk simultaneously reduce the probability of being classified as healthy.

**Prediabetes coefficients** are notably attenuated relative to both other classes, with HighChol and NoDocbcCost showing the largest retained effects. The attenuation is consistent with the near-inseparability finding in Section 6.5, as ridge assigns minimal weight to a class it cannot reliably distinguish even when all predictors are retained.

**HvyAlcoholConsump** shows a negative coefficient for Diabetes and near-zero for Prediabetes, consistent with the selection bias interpretation raised in Section 6.6.1 rather than a genuine protective effect.

### 6.6.3 Variable Importance

**Variable Importance: Gain Metric**

XGBoost variable importance is computed using the **gain** metric, which quantifies each feature's contribution to model performance:

-   **Gain** measures the total reduction in the loss function (e.g., multi-class log-loss) achieved whenever a feature is used to split nodes across all trees in the ensemble
-   Features with high gain contribute substantially to improving predictions by creating splits that better separate diabetes categories
-   **Gain vs alternatives:** Unlike split frequency (which counts how often a feature is used) or cover (which weights by number of observations), gain directly captures predictive impact and effect magnitude

Higher gain values indicate features that most effectively reduce prediction error when incorporated into the model's decision rules. This metric complements SHAP importance by focusing on in-model utility rather than marginal contributions to individual predictions.

```{r}
#| label: vip-rq1
#| fig-height: 7
#| fig-width: 14
xgb_final_fit |> 
  extract_fit_parsnip() |> 
  vip(num_features = 15, 
      geom = "col",
      aesthetics = list(fill = "steelblue",color = 'black', alpha = 0.8)) +
  labs(title = "Variable Importance: XGBoost Model",
       subtitle = "Top 15 predictors based on gain") 
```

**GenHlth and HighBP** dominate the importance ranking, accounting for the two largest gain values by a substantial margin. Their prominence here, unlike in the multinomial where GenHlth led but HighBP ranked fourth, suggests that XGBoost is exploiting non-linear interactions between blood pressure and other metabolic markers that the parametric models cannot capture.

**Age and BMI** rank third and fourth respectively, consistent across all three interpretation methods, confirming these as the most stable and robust predictors of diabetes status regardless of modelling assumptions.

**Income** enters the top five in the XGBoost ranking, a stronger signal than its position in the multinomial log-odds suggested. This is consistent with XGBoost capturing SES-mediated pathways through interactions with healthcare access, diet, and physical health variables that linear models partially suppress.

**HighChol** ranks sixth, lower relative to its multinomial prominence, suggesting its independent contribution is partially absorbed by correlated predictors such as HighBP and BMI in the tree-splitting process.

**Behavioural predictors** including smoking, fruit consumption, and heavy alcohol consumption cluster at the bottom of the ranking, consistent with the near-null effects observed in the multinomial analysis and confirming that their association with diabetes status is largely mediated rather than independent.

### 6.6.4 SHAP Importance

**SHAP Interpretation Methodology**

To complement the parametric log-odds interpretation, we employ **SHAP (SHapley Additive exPlanations)** to quantify feature contributions in the XGBoost model. SHAP offers several advantages:

-   **Model-agnostic foundation** grounded in game theory, enabling comparison across model types
-   **TreeSHAP algorithm** provides exact Shapley values by leveraging XGBoost tree structure (no Monte Carlo approximation needed)
-   **Multiclass compatibility** computes separate contributions for each diabetes category (No diabetes, Prediabetes, Diabetes)
-   **Mean absolute SHAP values** reveal global feature importance by measuring how much each feature, on average, shifts predicted probabilities from baseline

By comparing SHAP importance with logistic regression odds ratios, we assess method convergence, agreement strengthens confidence in identified risk factors, while discrepancies highlight non-linearities the parametric model cannot capture.

```{r}
#| label: shapviz-rq1

# xgb_engine <- extract_fit_engine(xgb_final_fit)
# 
# test_processed <- bake(
#   prep(extract_preprocessor(xgb_final_wf)),
#   new_data = diab_test
# )
# 
# X <- test_processed |>
#   select(-Diabetes,-weight) |>
#   as.matrix()
# 
# set.seed(42)
# X_sample <- X[sample(nrow(X), 10000), ]
# 
# shp <- shapviz(xgb_engine, X_pred = X_sample)
# write_rds(shp, "shp_rq1.rds")

# Load precomputed SHAP values to avoid recomputing expensive shapviz calculations
shp <- read_rds("shp_rq1.rds")


```

#### Shap Global Importance

```{r}
#| label: global-shp-rq1
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 14
sv_importance(shp, kind = "bar") +
  scale_fill_discrete(
    labels = c(
      "No diabetes",
      "Prediabetes",
      "Diabetes")) +
  theme_classic() +
  theme(legend.position = "bottom")
```

> **GenHlth** is the dominant predictor for Diabetes (red, $\approx 0.35$) by a substantial margin, with considerably smaller contributions for No diabetes and Prediabetes. This asymmetry suggests GenHlth is particularly informative for distinguishing established diabetes from the healthy state, consistent with the multinomial log-odds finding.
>
> **Age and BMI** show high and broadly distributed SHAP values across all three classes, confirming their role as general-purpose discriminators rather than class-specific signals. BMI shows notably higher No diabetes contribution (blue) than the other classes, suggesting it is particularly informative for ruling out diabetes.
>
> **HighBP** shows a striking class asymmetry, with a very large Diabetes SHAP value (red, $\approx 0.23$) relative to its Prediabetes contribution (green $\approx 0.02$), reinforcing the dose-response pattern identified in Section 6.6.1 and confirming that hypertension is a stronger signal for established diabetes than transitional risk.
>
> **Income and HighChol** both show meaningful Prediabetes contributions (green) relative to their class size, consistent with the lasso finding in Section 6.6.2 where these predictors were among the few retained with non-trivial Prediabetes coefficients.
>
> **Behavioural predictors** including smoking, heavy alcohol consumption, and NoDocbcCost cluster at the bottom of the ranking across all classes, confirming the mediation interpretation established across all three methods.

#### **Shap Beeswarm**

The beeswarm plots decompose SHAP contributions at the individual observation level, with colour indicating feature value (orange = high, purple = low) and horizontal position indicating the direction and magnitude of contribution to each class prediction.

```{r}
#| label: beeswarm-shp-rq1
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 18
#| dev: "ragg_png"
#| dpi: 180

names(shp) <- c("No diabetes", "Prediabetes", "Diabetes")

sv_importance(shp, kind = "beeswarm")
```

**No diabetes** is dominated by GenHlth and BMI, with high feature values (orange) pushing strongly positive, meaning poor general health and high BMI paradoxically increase No diabetes SHAP values here, which reflects the reference class encoding rather than a biological effect. Age shows the widest spread, with high age values pushing strongly rightward toward positive SHAP, consistent with older individuals being more likely to have received a formal diagnosis.

**Prediabetes** is notably distinct. Age dominates with the largest spread, and high age values push positively toward Prediabetes classification. Income, HighChol, and NoDocbcCost show meaningful contributions, consistent with the lasso and bar plot findings. The compressed scale and narrow distributions reflect the near-inseparability documented in Section 6.5.

**Diabetes** mirrors No diabetes in structure but with reversed directionality. GenHlth and HighBP show the largest positive contributions from high feature values, with Age and BMI following. HvyAlcoholConsump shows a distinctive rightward orange cluster, consistent with the selection bias artefact identified across all three interpretation methods.

### 6.6.5 Model Interpretability and Determinants

The triangulation of results across multinomial log-odds, elastic net coefficients, XGBoost variable importance, and SHAP values provides robust cross-method validation of the primary determinants of diabetes status in this population.

**Consensus risk factors.** All four interpretation methods converge on a consistent cardiometabolic risk profile. GenHlth, Age, BMI, HighBP, and HighChol rank in the top five predictors across every method, confirming these as the most stable and reliable discriminators of diabetes status regardless of modelling assumptions.

**Dose-response and non-linearity.** The multinomial log-odds establish clear directional effects, with GenHlth ($OR = 1.83$), BMI ($OR = 1.61$), and Age ($OR = 1.57$) as the strongest positive predictors of Diabetes. The SHAP beeswarm extends this by revealing non-linear acceleration, high Age values produce disproportionately large positive contributions, suggesting that risk amplifies when multiple clinical markers coincide rather than accumulating linearly.

**The Prediabetes gap.** Across all methods, Prediabetes coefficients and SHAP values are consistently attenuated relative to both adjacent classes. The lasso assigns minimal weight to Prediabetes predictors, SHAP distributions for the Prediabetes class are compressed and narrow, and the threshold analysis in Section 6.5 confirmed structural near-inseparability. This convergent evidence points to a fundamental limitation of survey-level features rather than a modelling failure.

**Behavioural mediation.** Physical activity, fruit consumption, and vegetable consumption show near-null effects across all methods ($OR \approx 0.98$, negligible SHAP contributions, bottom-ranked in VIP), consistent with full mediation through metabolic intermediates such as BMI and blood pressure rather than direct independent effects.

**Socioeconomic signal.** Income and Education show meaningful contributions in the XGBoost VIP and SHAP rankings despite modest multinomial log-odds, suggesting non-linear SES pathways that the parametric model partially suppresses.

| Method | Key Strength | Top Predictors |
|----|----|----|
| **Multinomial Log-Odds** | Directional parametric interpretation | GenHlth, BMI, Age |
| **Elastic Net** | Feature selection under regularisation | GenHlth, BMI, Age |
| **XGBoost VIP** | Non-linear predictive utility | GenHlth, HighBP, Age |
| **SHAP** | Direction, magnitude, and heterogeneity | GenHlth, Age, BMI |

## 6.7 Conclusion

> This research question asked the extent to which diabetes status can be explained and predicted from demographic, socioeconomic, behavioural, and cardiometabolic factors, and which predictors most strongly distinguish between the three outcome classes.

**Predictive performance:** Five models spanning parametric, regularised, tree-based, ensemble, and neural network approaches achieved macro-weighted ROC-AUC of approximately 0.82 across all candidates, with XGBoost marginally leading ($AUC = 0.822$). This level of discrimination reflects genuine probabilistic signal in the feature set, though class-wise evaluation tempers this finding: Diabetes recall ranges from 44--50% at the default threshold, meaning roughly half of true Diabetes cases are misclassified, and Prediabetes precision and recall collapse entirely to zero across four of five models. The ROC-AUC figure therefore reflects ranking ability rather than operational classification performance, and should be interpreted alongside the confusion matrices and class-wise heatmaps in Sections 6.4.2 and 6.4.3.

**Primary determinants:** Cardiometabolic factors dominate prediction across all methods. General health perception, BMI, age, high blood pressure, and high cholesterol consistently rank as the five strongest predictors regardless of modelling framework, confirming a robust multifactorial cardiometabolic risk profile. Demographic factors, particularly age and sex, contribute meaningfully but are secondary to clinical metabolic markers.

**Socioeconomic factors:** Income and education show non-linear SES-mediated pathways captured most clearly by XGBoost and SHAP, suggesting their effect on diabetes risk operates indirectly through metabolic intermediates rather than as direct independent predictors. This is consistent with the near-null multinomial log-odds for these variables relative to their VIP and SHAP prominence.

**Behavioural factors:** Physical activity, diet, and smoking show negligible independent predictive contributions across all methods, consistent with full mediation through BMI and blood pressure. Behavioural factors appear foundational but not directly discriminating at the survey measurement level available in this dataset.

**The Prediabetes limitation:** A critical finding is that no model could reliably distinguish Prediabetes from adjacent classes at population prevalence, despite strong overall discrimination. Threshold analysis, calibration plots, and SHAP distributions converge on a single explanation: the clinical signature of prediabetes in self-reported survey data overlaps substantially with both healthy and diabetic profiles, and the available predictors lack the granularity to isolate the transitionary phenotype. This points toward the need for continuous biomarker data such as HbA1c or fasting glucose to achieve clinically meaningful three-class separation.

**Overall answer:** Diabetes status is meaningfully explainable and predictable from the available feature set for the binary distinction between diabetes and no diabetes, with cardiometabolic markers providing the strongest and most consistent signal across all analytical frameworks. Discrimination is robust at the probability level, with macro-weighted ROC-AUC of 0.82 confirmed across all five models, though class-wise evaluation reveals that roughly half of true Diabetes cases are misclassified at the default threshold, tempering the strength of this claim. The three-class problem remains structurally constrained by the near-inseparability of prediabetes at population prevalence, a finding corroborated independently by calibration analysis, probability distributions, empirical threshold optimisation, and adjusted class-wise evaluation, representing both a limitation of the current analysis and a clinically important finding in its own right.
