---
title: "Modelling Diabetes Health Indicators"
subtitle: "MA7442 Modelling Data — Assessment 1 (Task 2)"
author: "Shahmeer Shahzad"
date: "2026-03-02"
format:
  pdf:
    geometry:
      - top=2cm
      - bottom=2cm
      - left=1.5cm
      - right=1.5cm
    documentclass: article
    df-print: kable
    mainfont: "Arial"
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r}
#| label: Loading_Packages
#| message: false
#| warning: false
#| include: false

# Packages
library(tidymodels)
library(tidyverse)
library(skimr)
library(scales)
library(patchwork)
library(viridis)
library(workflowsets)
library(rcompanion)
library(themis)
library(ranger)
library(vip)
library(shapviz)
library(knitr)
library(probably)
library(conflicted)
library(vip)
library(doParallel)

tidymodels_prefer(quiet = TRUE)
conflicts_prefer(yardstick::accuracy)

```

```{r}
#| label: plot-customise
#| include: false

# Plot format and dimension
knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(fig.width = 14, fig.height = 7)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Set global ggplot theme
theme_set(
  theme_classic(base_size = 16) +
    theme(
      plot.title    = element_text(size = 20, face = "bold"),
      axis.title    = element_text(size = 18),
      axis.text     = element_text(size = 14),
      legend.title  = element_text(size = 13),
      legend.text   = element_text(size = 12),
      strip.text    = element_text(face = "bold", size = 14),
      legend.background = element_rect(
        colour    = "black",
        fill      = "white",
        linewidth = 0.5
      )
    )
)
```

```{r}
#| label: palettes-used
#| include: false
pal_sex <- c(
  "Female" = "orange", 
  "Male"   = "darkgreen"   
)

pal_binary <- c(
  "No"  = "#BBBBBB",  
  "Yes" = "#882255"  
)

pal_genhlth <- c(
  "Excellent" = "#117733",  
  "Very good" = "#44AA99",  
  "Good"      = "#DDCC77",  
  "Fair"      = "#EE8866",  
  "Poor"      = "#CC6677"   
)
```

# 1. Introduction and Reflection on Task 1 Feedback

This report constitutes Task 2 of the MA7442 *Modelling Data*
assessment. It builds directly on the exploratory data analysis (EDA)
and research questions developed in Task 1, incorporating the feedback
received and extending the analytical scope accordingly.

In Task 1, the EDA was intentionally centred on diabetes status as the
primary outcome variable. This focus was adopted to maintain analytical
coherence and prevent the report from becoming overly expansive in
scope. Given the breadth of variables available in the dataset,
conducting a fully comprehensive EDA across multiple outcomes at that
stage would likely have diluted the conceptual focus and resulted in
excessive length. The analysis was therefore deliberately targeted to
support a diabetes-oriented modelling framework.

Feedback indicated, however, that restricting the EDA exclusively to
diabetes limited the diversity of subsequent research questions and led
to conceptual overlap. Task 2 therefore broadens the analytical
framework and introduces dedicated non-diabetes outcome analyses,
substantially expanding the empirical foundation of the study.

## 1.1 Reflection on Task 1 Feedback

Three areas of feedback were identified for improvement.

**Scope of the EDA.**\
The Task 1 EDA effectively characterised the dataset but examined
associations solely in relation to diabetes status. This constrained the
diversity of the research questions. In Task 2, Section 3 introduces a
comprehensive EDA block examining cardiovascular indicators, general
health measures, and behavioural variables as primary outcomes in their
own right.

**Research question overlap.**\
Because the initial EDA centred exclusively on diabetes, the research
questions were conceptually similar. In response, the structure has been
revised. RQ1 is retained in refined form; the previous RQ2 and RQ3 have
been consolidated into a single modelling-focused question (new RQ2);
and two new research questions (RQ3 and RQ4) are derived directly from
the expanded non-diabetes EDA.

**Figure line type.**\
In Task 1, a `geom_line` layer was used to connect mean unhealthy days
across age groups. Since age group is an ordered categorical variable
rather than a continuous measure, this implied unjustified linear
interpolation. In Task 2, age-structured visualisations have been
revised to use `geom_col` or lollipop-style representations, preserving
ordinal structure without implying continuity between categories.

## 1.2 Rationale for Expanded Exploratory Analysis

<div>

The expanded EDA presented in Task 2 is intentionally broader in scope
than in Task 1. This extension responds directly to feedback requesting
analysis of non-diabetes outcomes and greater utilisation of the
dataset. As a result, Section 3 systematically examines cardiovascular
indicators, health status variables, and behavioural correlates using
multiple complementary visualisations.

Given the size and multidimensional nature of the dataset, this
comprehensive exploration necessarily increases the length of the EDA
section. However, the extended analysis is structured purposefully: each
block serves to establish demographic gradients, socioeconomic
patterning, behavioural clustering, and outcome progression. The
additional length reflects analytical depth rather than redundancy, and
it provides the empirical basis for the revised and diversified research
questions developed in Task 2.

</div>

------------------------------------------------------------------------

# 2. Data Preparation

The dataset and all variable recoding steps are identical to those
described in Task 1 and are reproduced in the setup chunk below (Hidden)
to ensure a fully self-contained, reproducible document. No additional
transformations are required for the EDA in Section 3. All 253,680
records are retained; no missing values were present.

```{r}
#| label: data-import-and-recode
#| message: false
#| warning: false
#| include: false

diab <- read_csv("dataset1.csv")

diab <- diab |>
  mutate(
    Diabetes = factor(
      Diabetes_012,
      levels = c(0, 1, 2),
      labels = c("No diabetes", "Prediabetes", "Diabetes")
    ),
    across(
      c(CholCheck, Smoker, Stroke, HeartDiseaseorAttack,
        PhysActivity, Fruits, Veggies, HvyAlcoholConsump,
        AnyHealthcare, NoDocbcCost, DiffWalk),
      function(x) factor(x, levels = c(0, 1), labels = c("No", "Yes"))
    ),
    across(
      c(HighBP, HighChol),
      function(x) factor(x, levels = c(0, 1), labels = c("Not High", "High"))
    ),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    GenHlth = factor(
      GenHlth,
      levels  = 1:5,
      labels  = c("Excellent", "Very good", "Good", "Fair", "Poor"),
      ordered = TRUE
    ),
    Age = factor(
      Age,
      levels  = 1:13,
      labels  = c("18–24", "25–29", "30–34", "35–39", "40–44",
                  "45–49", "50–54", "55–59", "60–64", "65–69",
                  "70–74", "75–79", "80+"),
      ordered = TRUE
    ),
    Education = factor(
      Education,
      levels  = 1:6,
      labels  = c("Kindergarten or less", "Grades 1–8",
                  "Grades 9–11", "High school graduate",
                  "Some college / technical", "College graduate"),
      ordered = TRUE
    ),
    Income = factor(
      Income,
      levels  = 1:8,
      labels  = c("< $10k", "$10k–$14k", "$15k–$19k", "$20k–$24k",
                  "$25k–$34k", "$35k–$49k", "$50k–$74k", "$75k+"),
      ordered = TRUE
    )
  ) |>
  select(-Diabetes_012)
```

------------------------------------------------------------------------

# 3. Additional Exploratory Data Analysis: Non-Diabetes Outcomes

The following analyses explore relationships among health, lifestyle,
and socioeconomic variables *independently of diabetes status*. The goal
is to characterise the joint structure of these variables, in particular
to understand what shapes self-reported general health, mental health
burden, and functional limitation, and to motivate the non-diabetes
research questions introduced in Section 4.

## 3.1 Cardiovascular Risk Indicators: Demographic, Socioeconomic and Behavioural Gradients

This section explores the population structure and correlates of key
cardiovascular risk indicators, namely high blood pressure, high
cholesterol, and stroke. The analysis moves beyond diabetes to examine
how these outcomes are patterned across age, sex, socioeconomic
position, and behavioural factors. Emphasis is placed on identifying
demographic concentration, socioeconomic gradients, and clustering with
related health behaviours. The objective is to establish structural
regularities in the data that may inform subsequent modelling and risk
interpretation.

```{r}
#| label: age-sex-pyramid-prep
#| include: false

bp_pyramid <- diab |>
  count(HighBP, Age, Sex) |>
  group_by(HighBP, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

chol_pyramid <- diab |>
  count(HighChol, Age, Sex) |>
  group_by(HighChol, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_))

stroke_pyramid <- diab |>
  filter(!is.na(Stroke), !is.na(Age), !is.na(Sex)) |>
  count(Stroke, Age, Sex) |>
  group_by(Stroke, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

```

### 3.1.1 Age–Sex Distribution by High Blood Pressure Status

```{r}
#| label: highbp-age-sex-pyramid
#| fig.height: 7
#| fig.width: 14

ggplot(bp_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ HighBP, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by high blood pressure status",
    x     = "Age group",
    y     = "Percent within blood pressure group",
    fill  = "Sex"
  )
```

### 3.1.2 Age–sex distribution by **high cholesterol** status

```{r}
#| label: highchol-age-sex-pyramid
#| fig-height: 7
#| fig-width: 14

ggplot(chol_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ HighChol, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by high cholesterol status",
    x     = "Age group",
    y     = "Percent within cholesterol group",
    fill  = "Sex")
```

### 3.1.3 Age–sex distribution by stroke history

```{r}
#| label: stroke-age-sex-pyramid
#| fig-height: 7
#| fig-width: 14

stroke_pyramid <- diab |>
  count(Stroke, Age, Sex) |>
  group_by(Stroke, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

ggplot(stroke_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ Stroke, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by stroke history",
    x     = "Age group",
    y     = "Percent within stroke-history group",
    fill  = "Sex"
  )
```

> Across all three outcomes, high blood pressure, high cholesterol, and
> stroke, the dominant structural feature is age concentration. Each
> “Yes” panel is heavily skewed toward the 55–69 and older age bands,
> while representation below age 40 is minimal. This confirms that
> cardiovascular burden in this dataset is fundamentally age-structured
> rather than evenly distributed across the life course.
>
> However, the degree of age concentration differs meaningfully between
> outcomes. High blood pressure and high cholesterol show similar
> mid-to-late life clustering, with substantial presence beginning in
> the 50–54 band and peaking around 60–69. Stroke, by contrast, exhibits
> a far sharper right-skew: the “Yes” group is overwhelmingly
> concentrated in older age bands, with negligible representation in
> younger adults. This suggests stroke is not merely age-associated but
> strongly age-dependent relative to the other indicators.
>
> Sex differences are comparatively modest within age strata for high
> blood pressure and high cholesterol. Once age is conditioned upon,
> male–female distributions appear broadly parallel, implying that sex
> operates as a secondary structuring factor relative to age. In the
> stroke group, however, there is a slight female predominance in the
> oldest age bands, which may reflect differential survival into
> advanced age rather than a pure sex-specific risk gradient.
>
> Importantly, these pyramids depict the **composition within outcome
> groups**, not prevalence across the population. They therefore
> highlight demographic concentration rather than risk probability. The
> consistency across high blood pressure and high cholesterol suggests
> shared cardiometabolic ageing processes, whereas stroke displays a
> more extreme demographic compression, indicating it may represent the
> culmination of accumulated vascular risk rather than an early-stage
> cardiovascular condition.
>
> Taken together, the figures establish age as the primary axis of
> cardiovascular risk patterning in this sample. Any socioeconomic or
> behavioural gradients observed later must therefore be interpreted
> within the context of this strong underlying demographic structure.

```{r}
#| label: ses-heatmap
#| include: false

bp_ses <- diab |>
  mutate(
    bp_binary = if_else(HighBP == "High", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(prev_bp = mean(bp_binary),
            n = n(),
    .groups = "drop")

chol_ses <- diab |>
  mutate(chol_bin = if_else(HighChol == "High", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(
    prev_chol = mean(chol_bin),
    n = n(),
    .groups = "drop")

stroke_ses <- diab |>
  mutate(
    stroke_bin = if_else(Stroke == "Yes", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(prev_stroke = mean(stroke_bin),n = n(),
    .groups = "drop")
```

### 3.1.4 Joint SES heatmap of High Blood Presure prevalence.

```{r}
#| label: highbp-ses-heatmap
#| fig.height: 8
#| fig.width: 15

bp_ses |> 
  ggplot(aes(x = Income, y = Education, fill = prev_bp)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_bp, accuracy = 1)),
    colour   = "white",
    fontface = "bold",
    size     = 4.5) +
  scale_fill_gradient(
    low  = "#F1EEF6",
    high = "#882255",
    name = "Prevalence\nHigh BP",
    labels = percent_format(accuracy = 1)) +
  labs(title = "Prevalence of high blood pressure by education and income",
    x     = "Income band",
    y     = "Education level")
```

### 3.1.5 Joint SES heatmap of High Cholestrol prevalence.

```{r}
#| label: highchol-ses-heatmap
#| fig.height: 8
#| fig.width: 15

chol_ses |>
  ggplot(aes(x = Income, y = Education, fill = prev_chol)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_chol, accuracy = 1)),
    colour   = "white",
    fontface = "bold",
    size     = 3.5
  ) +
  scale_fill_gradient(
    low    = "#F1EEF6",      
    high   = "#882255",      
    name   = "Prevalence\nHigh cholesterol",
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Prevalence of high cholesterol by education and income",
    x     = "Income band",
    y     = "Education level"
  )
```

### 3.1.6 Joint SES heatmap of Stroke prevalence.

```{r}
#| label: stroke-ses-heatmap
#| fig.height: 8
#| fig.width: 15

stroke_ses |>
  ggplot(aes(x = Income, y = Education, fill = prev_stroke)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_stroke, accuracy = 1)),
    colour = "white",
    fontface = "bold",
    size = 3.5) +
  scale_fill_gradient(
    low = "#F1EEF6",      
    high = "#882255",
    name = "Prevalence\nStroke history",
    labels = percent_format(accuracy = 1)) +
  labs(title = "Prevalence of stroke history by education and income",
        x = "Income band",
        y = "Education level")
```

> Across all three outcomes, a clear socioeconomic gradient is evident.
> Prevalence is consistently highest in the lowest education and lowest
> income strata and declines with increasing educational attainment and
> income. The lower-left cells (least educated, lowest income) exhibit
> the greatest burden, while college graduates in the highest income
> bands show the lowest prevalence. This pattern indicates cumulative
> socioeconomic disadvantage rather than a single-dimension effect.
>
> High blood pressure and high cholesterol display particularly strong
> and stable two-dimensional gradients. Within nearly every education
> category, prevalence decreases progressively as income rises.
> Likewise, within each income band, higher education corresponds to
> lower prevalence. The monotonic pattern across both axes suggests that
> both income and education independently structure cardiometabolic
> risk, rather than one merely proxying for the other.
>
> Stroke exhibits the same directional gradient but at substantially
> lower absolute prevalence and with more variability across cells. Some
> low-education, low-income cells show markedly elevated stroke
> prevalence (e.g., 19%), while higher SES cells approach very low
> levels. The noisier pattern likely reflects smaller counts and the
> relative rarity of stroke, but the overall concentration in
> socioeconomically disadvantaged groups remains clear.
>
> Importantly, these are crude prevalence estimates. As shown in the
> age–sex pyramids, lower socioeconomic strata contain
> disproportionately older individuals, and age is a dominant
> determinant of all three outcomes. Therefore, part of the observed SES
> gradient likely reflects demographic composition. However, the
> persistence of a consistent gradient across both income and education
> dimensions suggests that socioeconomic position is not merely a proxy
> for age but a meaningful structural correlate of cardiovascular
> burden.

### 3.1.7 Prevalence Cardiovascular Outcomes Accumulated Risk Factors

```{r}
#| label: cvd-obesity-physact-prep
#| include: false
diab_risk <- diab |>
  mutate(
    Obese        = if_else(BMI >= 30, 1L, 0L),
    Inactive     = if_else(PhysActivity == "No", 1L, 0L),
    Smoker_bin   = if_else(Smoker == "Yes", 1L, 0L),
    Diabetes_bin = if_else(Diabetes != "No diabetes", 1L, 0L),
    risk_count   = Obese + Inactive + Smoker_bin + Diabetes_bin,
    risk_group   = case_when(
      risk_count == 0 ~ "0",
      risk_count == 1 ~ "1",
      risk_count == 2 ~ "2",
      risk_count == 3 ~ "3",
      risk_count >= 4 ~ "4+"
    ),
    risk_group = factor(risk_group, levels = c("0","1","2","3","4+"))
  )

cvd_long <- diab_risk |>
  select(risk_group, HighBP, HighChol, Stroke) |>
  pivot_longer(
    cols      = c(HighBP, HighChol, Stroke),
    names_to  = "Outcome",
    values_to = "Status"
  ) |>
  mutate(
    Outcome = recode(
      Outcome,
      HighBP   = "High blood pressure",
      HighChol = "High cholesterol",
      Stroke   = "Stroke history"
    ),
    outcome_event = case_when(
      Outcome == "High blood pressure" ~ Status == "High",
      Outcome == "High cholesterol"    ~ Status == "High",
      Outcome == "Stroke history"      ~ Status == "Yes",
      TRUE                             ~ NA
    )
  ) |>
  group_by(Outcome, risk_group) |>
  summarise(
    prop = mean(outcome_event, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: cvd-obesity-physact
#| fig.height: 6
#| fig.width: 14


ggplot(cvd_long, aes(x = risk_group, y = prop, fill = risk_group)) +
  geom_col(width = 0.7) +
  facet_wrap(~ Outcome, nrow = 1) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  scale_fill_brewer(palette = "Purples", name = "Risk\ncount") +
  labs(
    title = "Prevalence of cardiovascular outcomes by cardiometabolic risk factor count",
    x     = "Number of risk factors present\n(obesity, inactivity, smoking, diabetes)",
    y     = "Proportion with outcome"
  )
```

> Cardiovascular outcomes show a clear **dose–response relationship with
> the accumulation of cardiometabolic risk factors**. For both **high
> blood pressure** and **high cholesterol**, prevalence increases
> steadily as additional risk factors are present. Individuals with no
> risk factors show relatively low prevalence (around one quarter to one
> third), while those with **four or more risk factors exhibit very high
> prevalence**, exceeding roughly **70–80%** for both conditions. This
> pattern indicates that these outcomes are strongly associated with the
> combined burden of obesity, physical inactivity, smoking, and
> diabetes.
>
> **Stroke history follows the same upward trend**, but at substantially
> lower levels overall. Even among individuals with multiple risk
> factors, stroke prevalence remains below roughly **10–12%**,
> reflecting the relative rarity of stroke compared with hypertension
> and hypercholesterolaemia in the population.
>
> Overall, the figure demonstrates that **cardiometabolic risk factors
> act cumulatively rather than independently**. As risk exposures
> accumulate, the likelihood of cardiovascular conditions, particularly
> hypertension and high cholesterol, increases markedly, highlighting
> the importance of addressing multiple lifestyle and metabolic risk
> factors simultaneously.

### 3.1.8 Joint blood pressure and cholesterol profiles across age groups

```{r}
#| label: highchol-highbp-age-prep
#| include: false

bp_chol_age <- diab |>
  count(Age, HighBP, HighChol) |>
  group_by(Age) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

```

```{r}
#| label: highchol-highbp-age
#| fig-width: 14
#| fig-height: 6

ggplot(bp_chol_age,
       aes(x = Age, y = prop, fill = interaction(HighBP, HighChol,
                                                  sep = " / Chol: "))) +
  geom_col(position = "fill", width = 0.85) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(
    title = "Joint blood pressure and cholesterol status by age group",
    x     = "Age group",
    y     = "Proportion within age group",
    fill  = "BP / Cholesterol"
  ) 
```

> The distribution of cardiovascular profiles shifts systematically with
> age. Younger groups are overwhelmingly characterised by normal blood
> pressure and cholesterol, whereas the proportion with dual elevation
> (high/high) increases sharply from midlife onward and becomes dominant
> in older age bands. This pattern reflects progressive cardiometabolic
> accumulation rather than isolated risk emergence.
>
> Importantly, the high/high category expands more rapidly than either
> single-risk category, indicating clustering rather than independence
> of these conditions. By ages 65+, dual dysregulation accounts for a
> substantial share of the population within each age group, suggesting
> convergence of vascular risk factors with ageing.
>
> The pattern implies that hypertension and hypercholesterolaemia are
> not parallel but interacting processes within cardiovascular ageing,
> reinforcing their role as intermediate markers in the pathway toward
> more severe vascular outcomes.

### 3.1.9 Co-occurrence of Hypertension and High Cholesterol

```{r}
#| label: highbp-highchol-joint
#| fig.height: 8
#| fig.width: 12

diab |>
  count(HighBP, HighChol) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = HighBP, y = HighChol, fill = prop)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(
    aes(label = percent(prop,accuracy = 0.1)),
    size     = 6,
    fontface = "bold",
    colour   = "white"
  ) +
  labs(
    title = "Joint prevalence of high blood pressure and high cholesterol",
    x     = "High blood pressure",
    y     = "High cholesterol"
  ) +
  theme(
    legend.position = 'none'
  )
```

> Approximately **40.2%** of the sample has neither high blood pressure
> nor high cholesterol. The remaining **59.8%** have at least one of the
> two conditions, indicating that cardiometabolic risk is common in this
> population.
>
> The largest single elevated profile is **concurrent high blood
> pressure and high cholesterol (25.5%)**, which exceeds either isolated
> condition (17.4% BP only; 16.9% cholesterol only). This suggests
> meaningful clustering rather than independent occurrence of the two
> risk factors.
>
> Overall, hypertension and hypercholesterolaemia appear strongly
> co-patterned, with dual elevation more common than either
> single-condition state, reinforcing their joint role in cardiovascular
> risk stratification.

### 3.1.10 Stroke prevalence by blood pressure and cholesterol status

```{r}
#| label: stroke-bp-chol-prep
#| include: false
stroke_bpchol <- diab |>
  count(HighBP, HighChol, Stroke) |>
  group_by(HighBP, HighChol) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  filter(Stroke == "Yes") |>
  mutate(
    BP_Chol = paste0("BP: ", HighBP, "\nChol: ", HighChol)
  )

```

```{r}
#| label: stroke-bp-chol
#| fig-width: 14
#| fig-height: 7

ggplot(stroke_bpchol,
       aes(x = BP_Chol, y = prop, fill = prop)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = percent(prop, accuracy = 0.1)),
            vjust = -0.4, size = 5.5, fontface = "bold") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     limits = c(0, 0.12)) +
  scale_fill_viridis_c(option = "C", direction = -1, guide = "none") +
  labs(
    title = "Stroke prevalence by blood pressure and cholesterol status",
    x     = "Blood pressure / cholesterol profile",
    y     = "Proportion with stroke history"
  )
```

> Stroke prevalence increases monotonically across cardiovascular
> profiles. The highest prevalence is observed among individuals with
> both high blood pressure and high cholesterol, followed by isolated
> high blood pressure, isolated high cholesterol, and neither condition.
>
> The gradient suggests cumulative vascular burden, with blood pressure
> appearing more strongly associated with stroke than cholesterol alone.
> Stroke thus reflects downstream risk concentration rather than
> isolated cardiometabolic dysfunction.

### 3.1.11 Joint co-occurrence profiles of high blood pressure, high cholesterol, and stroke

```{r}
#| label: joint-prep
#| include: false

joint <- diab |>
  summarise(
    `HighBP & HighChol` = mean(HighBP == "High" & HighChol == "High"),
    `HighBP & Stroke`   = mean(HighBP == "High" & Stroke   == "Yes"),
    `HighChol & Stroke` = mean(HighChol == "High" & Stroke == "Yes"),
    `HighBP only` = mean(HighBP == "High" & HighChol != "High" & Stroke != "Yes"),
    `HighChol only`= mean(HighChol == "High" & HighBP != "High" & Stroke != "Yes"),
    `Stroke only` = mean(Stroke == "Yes" & HighBP != "High" & HighChol != "High"),
    `All three`= mean(HighBP == "High" & HighChol == "High" & Stroke == "Yes")
  ) |>
  pivot_longer(everything(), names_to = "Profile", values_to = "Proportion")

max_prop <- max(joint$Proportion)
```

```{r}
#| fig-height: 7
#| fig-width: 14

ggplot(joint, aes(x = reorder(Profile, Proportion), y = Proportion, fill = Proportion)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = percent(Proportion, accuracy = 0.1)),
    hjust = -0.15,
    size  = 4.5
  ) +
  coord_flip() +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, max_prop * 1.15)  
  ) +
  scale_fill_viridis_c(option = "D", direction = -1, guide = "none") +
  labs(
    title = "Co-occurrence profiles for HighBP, HighChol, and Stroke",
    x     = NULL,
    y     = "Proportion of full sample"
  )
```

### 3.1.12 Association Strength - Cramér's V

```{r}
#| label: cvd-binary-indicators-prep
#| include: false
nondm_outcomes <- c("HighBP", "HighChol", "Stroke")
pred_vars <- c("Diabetes","Age", "Sex", "Income", "Education",
               "PhysActivity", "Smoker", "BMI_cat",
               "DiffWalk", "HeartDiseaseorAttack",
               "AnyHealthcare", "NoDocbcCost",
               "GenHlth", "HvyAlcoholConsump")

diab_cv <- diab |>
  mutate(BMI_cat = cut(BMI,
                       breaks = c(0, 18.5, 25, 30, Inf),
                       labels = c("Underweight", "Normal",
                                  "Overweight", "Obese"),
                       right  = FALSE))

cv_nondm <- map_dfr(nondm_outcomes, function(outcome) {
  map_dfr(pred_vars, function(pred) {
    v <- cramerV(table(diab_cv[[outcome]], diab_cv[[pred]]))
    tibble(Outcome = outcome, Predictor = pred, CramersV = as.numeric(v))
  })
}) |>
  mutate(
    Outcome = recode(Outcome,
                     HighBP   = "High Blood Pressure",
                     HighChol = "High Cholesterol",
                     Stroke   = "Stroke")
  )
```

```{r}
#| label: cvd-binary-indicators
#| fig-height: 9
#| fig-width: 14
ggplot(cv_nondm,
       aes(x = reorder(Predictor, CramersV),
           y = CramersV,
           fill = Outcome)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = round(CramersV, 2)),
            position = position_dodge(width = 0.75),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  expand_limits(x = max(cv_nondm$CramersV) + 0.03) +
  scale_fill_manual(values = c(
    "High Blood Pressure" = "#2c7bb6",
    "High Cholesterol"    = "#f46d43",
    "Stroke"              = "#1a9641"
  )) +
  scale_y_continuous(limits = c(0, 0.45)) +
  labs(
    title = "Association strength with non-diabetes outcomes (Cramér's V)",
    x     = NULL,
    y     = "Cramér's V",
    fill  = "Outcome"
  ) +
  theme(legend.position = "bottom")
```

> This chart highlights how different risk factors weigh against
> cardiovascular outcomes using **Cramér’s V** (a measure of association
> where higher values indicate a stronger relationship).
>
> The data reveals two distinct "profiles" of disease:
>
> 1.  **Metabolic & Age-Driven (High BP & Cholesterol)**
>
> Hypertension and High Cholesterol are primarily driven by **systemic,
> long-term factors**.
>
> -   **Primary Drivers:** **Age** ($V \approx 0.35$) and **General
>     Health** ($V \approx 0.30$) are the strongest predictors.
>
> -   **Secondary Drivers:** Strong links to **Diabetes** and **BMI**,
>     suggesting these are markers of general metabolic decline.
>
> -   **Takeaway:** These conditions follow a broad, population-wide
>     risk gradient shaped by getting older and overall physical state.
>
> 2.  **Clinical & Pathological (Stroke)**
>
> Stroke follows a different logic, focusing more on **existing damage**
> than just age.
>
> -   **Primary Drivers:** **Prior Heart Disease** ($V \approx  0.20$),
>     **General Health** ($V \approx 0.19$), and **Walking Difficulty**
>     ($V \approx 0.18$).
>
> -   **Key Difference:** Age is significantly less dominant here than
>     it is for Blood Pressure. BMI has almost no direct association
>     with Stroke in this specific view.
>
> -   **Takeaway:** Stroke appears as a "clinical endpoint", it is most
>     strongly associated with patients who already show signs of
>     cardiovascular failure or physical impairment.

|                         |                         |                       |
|-------------------------|-------------------------|-----------------------|
| **Outcome**             | **Strongest Predictor** | **Second Predictor**  |
| **High Blood Pressure** | Age (0.35)              | General Health (0.30) |
| **High Cholesterol**    | Age (0.29)              | General Health (0.21) |
| **Stroke**              | Heart Disease (0.20)    | General Health (0.19) |

### 3.1.13 Summary of Key Findings

Section 3.1 establishes that cardiovascular risk in this dataset is
structured along three primary axes: **age, socioeconomic position, and
cardiometabolic clustering**.

First, **age is the dominant organising factor**. High blood pressure
and high cholesterol rise steadily from midlife onward, while stroke is
far more sharply concentrated in older age groups. Stroke appears less
as a gradual condition and more as a late-stage vascular outcome.

Second, a **clear and consistent socioeconomic gradient** is evident
across all three outcomes. Lower education and lower income strata show
substantially higher crude prevalence, with monotonic declines across
both dimensions. Although partly influenced by age composition, the
two-dimensional gradient suggests an independent structural role for
socioeconomic position.

Third, **cardiometabolic conditions cluster rather than occur
independently**. Dual high blood pressure and high cholesterol (25.5%)
exceed either isolated condition, and stroke prevalence increases
monotonically across BP/cholesterol profiles, peaking in the dual-high
group. This supports a cumulative vascular burden framework.

Fourth, **obesity strongly structures hypertension and cholesterol but
not stroke**, reinforcing the interpretation of BP and cholesterol as
intermediate metabolic states. Stroke instead aligns more strongly with
prior heart disease, general health, and walking difficulty.

Finally, association analyses (Cramér’s V) reveal two distinct disease
profiles:

-   **Hypertension and cholesterol** are predominantly age and
    general-health driven.

-   **Stroke** is more closely associated with established
    cardiovascular pathology and functional impairment.

Overall, Section 3.1 demonstrates that cardiovascular outcomes in this
dataset are not randomly distributed but follow coherent demographic,
socioeconomic, and pathological structures. These findings provide the
empirical foundation for the modelling strategy developed in Section 4.

### 3.1.14 Implications for Modelling

The exploratory analysis of cardiometabolic outcomes suggests several
considerations for subsequent modelling. First, the strong accumulation
pattern observed across cardiometabolic risk factors indicates that
lifestyle and metabolic variables such as obesity, smoking, physical
inactivity, and diabetes are likely to act as key predictors of
cardiovascular outcomes. Second, the association strength analysis shows
that demographic and health-status variables—particularly age and
self-rated general health—have the strongest relationships with high
blood pressure and high cholesterol, indicating that these variables
should be prioritised in predictive models.

For stroke, the strongest associations appear with prior cardiovascular
disease history, poorer general health, and functional limitation
(difficulty walking), suggesting that stroke risk may be more closely
linked to underlying health status and comorbidity than to behavioural
factors alone. Finally, the relatively low prevalence of stroke compared
with hypertension and hypercholesterolaemia suggests that models
predicting stroke may face class imbalance and may require approaches
suitable for rare outcomes.

Overall, the EDA indicates that **demographic factors, general health
status, and cardiometabolic risk profiles** are likely to be central
components of the modelling framework used to investigate cardiovascular
outcomes.

## 3.2 Patterns and determinants of mental health burden

This section examines the distribution and key correlates of **mental
health burden**, measured as the number of mentally unhealthy days
reported in the past 30 days. The analysis explores how mental health
burden varies across demographic, socioeconomic, and health-related
factors, including age, physical health, and lifestyle behaviours. By
identifying systematic patterns in mental health outcomes, this section
aims to highlight the factors most strongly associated with elevated
psychological distress in the population.

### 3.2.1 Distribution of mentally unhealthy days in the past 30 days

```{r}
#| label: mentlhlth-distribution
#| fig-height: 6
#| fig-width: 14
ggplot(diab, aes(x = MentHlth)) +
  geom_histogram(
    binwidth = 1,
    boundary = 0,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = seq(0, 30, by = 5),
    limits = c(0, 30)
  ) +
  scale_y_continuous(
    labels = label_number(scale = 1e-3, suffix = "K"),
    breaks = seq(0,200000,by = 25000)
  ) +
  labs(
    title = "Distribution of mentally unhealthy days in the past 30 days",
    x = "Number of mentally unhealthy days",
    y = "Frequency") 
```

> The distribution of mentally unhealthy days is **highly
> right-skewed**, with a very large proportion of respondents reporting
> **zero mentally unhealthy days in the past month**. Frequencies
> decline sharply as the number of unhealthy days increases, producing a
> long right tail that extends toward the upper bound of 30 days. This
> pattern indicates that while most individuals report little or no
> recent mental health burden, a smaller subset experiences **frequent
> or persistent distress**.
>
> The distribution also shows evidence of **excess zeros and strong
> dispersion**, with the spike at zero far exceeding the frequencies of
> neighbouring values. Such a structure suggests that standard linear
> modelling would be inappropriate, and motivates the use of
> **count-based models**, potentially including **zero-inflated or
> overdispersed models (e.g., Poisson or zero-inflated negative
> binomial)** when analysing mental health burden as an outcome in
> subsequent modelling.

### 3.2.3 Mental Health Burden by Income and Sex

```{r}
#| label: mental-health-income
#| fig.height: 7
#| fig.width: 14
diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(Income, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  
  ggplot(aes(x = Income, y = prop_high, fill = Sex)) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.7,
    colour = "black",
    linewidth = 0.2
  ) +
  geom_text(
    aes(label = percent(prop_high, accuracy = 1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  )
```

> A clear socioeconomic gradient is evident in the prevalence of high
> mental health burden. The proportion of respondents reporting ≥14
> mentally unhealthy days declines steadily as household income
> increases, indicating that individuals in lower-income groups
> experience substantially higher levels of psychological distress. In
> the lowest income category (\<\$10k), roughly 28% of females and 26%
> of males report high mental health burden, compared with only 6% and
> 4% respectively in the highest income group (\$75k+).
>
> Across all income categories, females consistently report higher
> prevalence than males, although the difference between sexes is modest
> relative to the overall income gradient. These results suggest that
> socioeconomic position is a strong determinant of mental health
> burden, with financial disadvantage associated with markedly higher
> levels of frequent psychological distress. (Money doesn't buy
> happiness :) )

### 3.2.3 Mental Health Burden by Educatioand Sex

```{r}
#| label: mental-health-education
#| fig.height: 7
#| fig.width: 14
diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(Education, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  
  ggplot(aes(x = Education, y = prop_high, fill = Sex)) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.7,
    colour = "black",
    linewidth = 0.2
  ) +
  geom_text(
    aes(label = percent(prop_high, accuracy = 1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

> A clear educational gradient is visible in the prevalence of high
> mental health burden. Individuals with lower levels of educational
> attainment report substantially higher proportions of ≥14 mentally
> unhealthy days compared with those with higher education. For example,
> among respondents with kindergarten or less education, approximately
> 22% report high mental health burden, whereas this proportion declines
> to around 8% for females and 5% for males among college graduates.
>
> Across all education categories, females consistently report higher
> levels of mental health burden than males, although the difference
> between sexes is relatively modest compared with the overall education
> gradient. The steady decline in prevalence with increasing educational
> attainment suggests that socioeconomic resources and opportunities
> associated with education may play an important role in mental health
> outcomes.
>
> Overall, the figure indicates that lower educational attainment is
> associated with substantially higher levels of frequent psychological
> distress, reinforcing the broader pattern of socioeconomic
> inequalities in mental health observed in the income analysis.

### 3.2.4 Mental Health Burden by Age Group and Sex

```{r}
#| label: mental-health-age-lollipop
#| fig.height: 7
#| fig.width: 14


mh_age_sex <- diab |>
  group_by(Age, Sex) |>
  summarise(mean_mh = mean(MentHlth, na.rm = TRUE), .groups = "drop")

pos_dodge <- position_dodge(width = 0.6)

ggplot(mh_age_sex, aes(x = Age, y = mean_mh, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = mean_mh),
    position = pos_dodge,
    linewidth = 1
  ) +
  geom_point(
    position = pos_dodge,
    size = 3.8
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_manual(
    values = pal_sex
  ) +
  labs(
    title  = "Mean mentally unhealthy days by age group and sex",
    x      = "Age group",
    y      = "Mean mentally unhealthy days (past 30 days)",
    colour = "Sex"
  ) 
```

> Mean mentally unhealthy days are higher among women than men in every
> age group. The difference is most visible in younger and middle
> adulthood, where women report roughly 1–2 more days on average than
> men.
>
> Across age, the pattern is not monotonic. Mean days are relatively
> high in early adulthood, remain similar through midlife, and then
> decline progressively after around age 60 for both sexes. The lowest
> averages are observed in the oldest age groups.

### 3.2.5 Joint Distribution of Mental and Physical Health Burden by Sex

```{r}
#| label: menth-physh-hexbin
#| fig.height: 7
#| fig.width: 14
ggplot(diab, aes(x = MentHlth, y = PhysHlth)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c(option = "magma", name = "Count", trans = "log10") +
  facet_wrap(~ Sex, nrow = 1) +
  labs(
    title = "Joint distribution of mentally and physically unhealthy days by sex",
    x     = "Mentally unhealthy days (past 30)",
    y     = "Physically unhealthy days (past 30)"
  )
```

> In both panels, the highest concentration of observations occurs near
> zero mentally and physically unhealthy days. Density decreases as
> values increase along either axis, and combinations of simultaneously
> high mental and physical burden are relatively sparse.
>
> There is visible clustering along diagonal regions, indicating that
> higher numbers of mentally unhealthy days tend to coincide with higher
> numbers of physically unhealthy days rather than being concentrated
> along a single axis.
>
> The overall distributional shape is similar for women and men, with
> broadly comparable density patterns across low, moderate, and higher
> values.

### 3.2.6 Prevalence of High Mental Health Burden by Physical and Cardiometabolic indicators

```{r}
#| include: false
diab_mh_phys <- diab |>
  mutate(
    HighMH = factor(if_else(MentHlth >= 14, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    Obese = factor(if_else(BMI >= 30, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    BadPhysHlth = factor(if_else(PhysHlth >= 14, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    Diabetes = factor(if_else(Diabetes == "Diabetes", "Yes", "No"),
                        levels = c("No", "Yes")),
    
    HighBP = factor(if_else(HighBP == "High", "Yes", "No"),
                        levels = c("No", "Yes")),
    
    HighChol = factor(if_else(HighChol == "High", "Yes", "No"),
                        levels = c("No", "Yes"))
  ) |>
  select(
    HighMH,
    Obese,
    PhysActivity,
    DiffWalk,
    BadPhysHlth,
    Diabetes,
    HeartDiseaseorAttack,
    HighBP,
    HighChol
  ) |>
  pivot_longer(
    cols = -HighMH,
    names_to = "Indicator",
    values_to = "Status"
  ) |> 
  count(Indicator, Status, HighMH) |>
  group_by(Indicator, Status) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  filter(HighMH == "Yes")
```

```{r}
#| fig.height: 7
#| fig.width: 14

diab_mh_phys |>
  ggplot(aes(x = Status, y = prop, fill = Status)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = percent(prop, accuracy = 1)),
    vjust = -0.3,
    size  = 3.3
  ) +
  facet_wrap(~ Indicator, nrow = 2, ncol = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.10))
  ) +
  scale_fill_manual(values = pal_binary) +
  labs(
    title = "Prevalence of high Mental Health Burden by Physical and Cardiometabolic Indicators",
    x     = "Indicator status",
    y     = "Proportion with ≥14 mentally unhealthy days"
  )
```

> High mental health burden (≥14 mentally unhealthy days) varies
> substantially across physical and cardiometabolic indicators. The
> strongest associations are observed for indicators of functional and
> physical health limitations. Individuals reporting high physical
> health burden (≥14 physically unhealthy days) show a markedly elevated
> prevalence of high mental health burden ( ≈ 33%), compared with only
> about 6% among those without high physical burden. A similarly large
> difference appears for difficulty walking, where prevalence rises from
> roughly 7% to 24%.
>
> Cardiometabolic conditions such as diabetes, heart disease or
> myocardial infarction, high blood pressure, and high cholesterol are
> also associated with higher levels of mental health burden, although
> the magnitude of these differences is smaller. For example,
> respondents with diabetes or heart disease exhibit prevalence around
> 15%, compared with approximately 9% among those without these
> conditions.
>
> Lifestyle factors show a mixed pattern. Obesity is associated with
> moderately higher mental health burden (≈12% vs ≈8%), while physical
> activity appears protective, with prevalence roughly halved among
> individuals reporting recent physical activity ( ≈ 8%) compared with
> those reporting none ( ≈16%).
>
> Overall, the results suggest that mental health burden is most
> strongly linked with physical health limitations and functional
> impairment, while cardiometabolic disease and lifestyle factors
> exhibit weaker but still noticeable associations.

### 3.2.7 Summary of Key Findings

Several clear structural patterns emerge in the analysis of mental
health burden. First, the distribution of mentally unhealthy days is
highly right-skewed, with a large concentration at zero and a long tail
toward higher values. This indicates substantial heterogeneity in
psychological distress across the population and motivates the use of
count-based models with potential zero inflation in subsequent
modelling.

Second, strong **socioeconomic gradients** are visible. The prevalence
of high mental health burden declines consistently with increasing
income and education, suggesting that socioeconomic disadvantage is
associated with substantially higher levels of frequent psychological
distress.

Third, **demographic differences** are evident. Women report higher
average numbers of mentally unhealthy days than men across all age
groups, while mental health burden tends to decline gradually in older
age groups after midlife.

Fourth, mental and physical health burdens show **moderate
co-occurrence**, with higher levels of mentally unhealthy days tending
to coincide with higher levels of physically unhealthy days.

Finally, mental health burden is **most strongly associated with
indicators of physical limitation**, particularly poor physical health
and difficulty walking. Cardiometabolic conditions such as diabetes,
heart disease, hypertension, and high cholesterol also show positive
associations with mental health burden, although these relationships are
weaker than those linked to functional impairment.

Overall, the results suggest that mental health burden in this
population is structured primarily by **socioeconomic position, physical
health status, and functional limitation**, providing a clear empirical
basis for the modelling analyses developed in later sections.

### 3.2.8 Implications for Modelling

The exploratory analysis provides several insights relevant for model
development. The distribution of mentally unhealthy days is highly
right-skewed with a substantial concentration at zero, suggesting that
standard linear modelling would be inappropriate and motivating the use
of count-based models such as Poisson or negative binomial regression,
potentially with zero-inflation.

Socioeconomic gradients observed across income and education indicate
that demographic and socioeconomic variables should be included as key
predictors.

Finally, the strong co-occurrence between mental and physical health
indicators suggests that physical health variables such as poor physical
health days, difficulty walking, and chronic disease indicators may play
an important explanatory role in modelling mental health burden.

### 3.x.x Spearman Correlation Heatmap: Continuous and Ordinal Variables

```{r}
#| label: spearman-prep
#| include: false
corr_vars <- diab |>
  transmute(
    BMI,
    MentHlth,
    PhysHlth,
    GenHlth   = as.integer(GenHlth),
    Age       = as.integer(Age),
    Education = as.integer(Education),
    Income    = as.integer(Income)
  )

spear_mat <- cor(
  corr_vars,
  method = "spearman",
  use    = "pairwise.complete.obs"
)

spear_tidy <- spear_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(
    cols      = -Var1,
    names_to  = "Var2",
    values_to = "rho"
  ) |>
  mutate(
    Var1 = factor(Var1, levels = colnames(spear_mat)),
    Var2 = factor(Var2, levels = colnames(spear_mat))
  )
```

```{r}
#| label: spearman-heatmap
#| fig.height: 7
#| fig.width: 10

ggplot(spear_tidy, aes(x = Var1, y = Var2, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", rho)),
    size   = 3.5,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#D6604D",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman's \u03c1"
  ) +
  labs(
    title = "Spearman rank correlations among continuous and ordinal variables",
    x     = NULL,
    y     = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    axis.text.y = element_text(hjust = 1)
  )
```

> The Spearman rank correlations indicate mostly **weak to moderate**
> monotonic associations among the continuous and ordinal variables.
>
> -   The strongest positive correlations are between **GenHlth and
>     PhysHlth** (ρ ≈ 0.45) and between **Education and Income** (ρ ≈
>     0.45), suggesting moderate monotonic relationships in both cases.
>
> -   **GenHlth** is moderately **negatively** correlated with
>     **Income** (ρ ≈ −0.35) and **Education** (ρ ≈ −0.28). Because
>     higher GenHlth scores denote poorer health, these negative
>     correlations imply that higher income and education are associated
>     with **better** self-rated health.
>
> -   **MentHlth** and **PhysHlth** show a modest positive association
>     (ρ ≈ 0.31), consistent with some co-occurrence of mental and
>     physical health burden.
>
> -   Most other pairwise correlations are small (\|ρ\| \< 0.3). In
>     particular, **BMI** and **Age** have only weak associations with
>     the other variables in this set.
>
> Overall, the correlation matrix does **not** indicate problematic
> collinearity: apart from a few moderate associations
> (GenHlth–PhysHlth, Education–Income), most variables are only weakly
> related. This supports their concurrent use as predictors in
> multivariable models, subject to routine multicollinearity checks.

### 3.x.x Cramér's V Association Heatmap: Binary Clinical and Lifestyle Variables

To complement the Spearman correlation heatmap for continuous/ordinal
variables, the strength of pairwise associations among all binary
clinical and lifestyle indicators is quantified using Cramér's V. This
symmetric measure summarises the association in each 2×2 contingency
table on a 0–1 scale, without assuming any directionality. Diabetes
status is deliberately excluded.

Together, Sections 3.14 and 3.15 establish the correlation structure
among all key variables, identifying which predictors are inter-related
and justifying the need for multivariable adjustment in all subsequent
models.

```{r}
#| label: cramer-prep
#| include: false
bin_vars <- c("HighBP", "HighChol", "Smoker", "Stroke",
              "HeartDiseaseorAttack", "PhysActivity",
              "HvyAlcoholConsump", "DiffWalk",
              "AnyHealthcare", "NoDocbcCost","Diabetes")

cv_mat <- outer(
  bin_vars, bin_vars,
  Vectorize(function(a, b) {
    cramerV(table(diab[[a]], diab[[b]]))
  })
)
dimnames(cv_mat) <- list(bin_vars, bin_vars)

cv_tidy <- cv_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(-Var1, names_to = "Var2", values_to = "V") |>
  mutate(
    Var1 = factor(Var1, levels = bin_vars),
    Var2 = factor(Var2, levels = bin_vars)
  )
```

```{r}
#| label: cramerv-heatmap
#| fig.height: 7
#| fig.width: 10

ggplot(cv_tidy, aes(x = Var1, y = Var2, fill = V)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", V)),
    size   = 3.8,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low       = "white",
    high     = "#D6604D",
    midpoint = 0,
    limits   = c(0, 1),
    name     = "Cramér's V"
  ) +
  labs(
    title = "Cramér's V among binary clinical and lifestyle variables",
    x = NULL,
    y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

> The Cramér’s V matrix indicates that most pairwise associations among
> the binary clinical and lifestyle variables are **weak**, with the
> majority of values below 0.20.
>
> The strongest association is between **HighBP and HighChol** (V ≈
> 0.30), representing a moderate relationship relative to the rest of
> the matrix. This aligns with known clustering of cardiovascular risk
> factors.
>
> Other comparatively larger associations include:
>
> -   **PhysActivity and DiffWalk** (V ≈ 0.25)
>
> -   **HighBP and DiffWalk** (V ≈ 0.22)
>
> -   **HighBP and HeartDiseaseorAttack** (V ≈ 0.21)
>
> -   **Stroke and HeartDiseaseorAttack** (V ≈ 0.20)
>
> These moderate associations are clinically plausible and reflect
> shared cardiovascular or functional impairment pathways.
>
> Lifestyle indicators such as **Smoker** and **HvyAlcoholConsump**
> exhibit generally weak associations with other variables (typically V
> \< 0.15). Healthcare access indicators (**AnyHealthcare**,
> **NoDocbcCost**) also show minimal association with most clinical
> variables.
>
> **Analytical Implication**
>
> The matrix does not suggest substantial redundancy among the binary
> predictors. Aside from a small number of moderate associations
> (primarily within cardiovascular-related variables), most
> relationships are weak.
>
> This supports inclusion of these variables simultaneously in
> multivariable models, with routine collinearity diagnostics (e.g., VIF
> or condition indices) as a precaution rather than an expected issue.

# 4. Refined Research Questions

## RQ1: Multidomain Determinants and Prediction of Diabetes Status

**Outcome Variable:** *Diabetes — Multiclass Outcome*

**Research Question**

> ***To what extent can diabetes status be explained and predicted using
> demographic, socioeconomic, behavioural, and cardiometabolic factors,
> and which predictors contribute most strongly to distinguishing
> between no diabetes, prediabetes, and diabetes?***

**Rationale**

Exploratory analysis of the CDC Diabetes Health Indicators dataset
indicates that diabetes status is structured by multiple interacting
determinants. Section 3.2.1 demonstrated a strong **age gradient**, with
prediabetes and diabetes concentrated in older age groups, suggesting
that demographic factors must be accounted for when evaluating other
risk indicators.

Cardiometabolic clustering was evident across several visualisations.
The joint distribution of **high blood pressure and high cholesterol**
increased substantially across diabetes categories, while the
risk-factor accumulation plot showed a clear gradient in which diabetes
prevalence rises with the number of cardiometabolic risk indicators.
This pattern aligns with established evidence linking obesity,
hypertension, and dyslipidaemia to type 2 diabetes risk (Ismail et al.,
2021).

Behavioural factors also exhibited variation across diabetes categories.
Physical activity levels tended to decline with worsening diabetes
status, while smoking and alcohol consumption showed weaker separation.
Because behavioural variables are potentially modifiable, evaluating
their contribution to diabetes risk is relevant for prevention
strategies (Aune et al., 2015).

Finally, socioeconomic gradients were visible in the education–income
distribution, where diabetes prevalence was higher among lower
socioeconomic groups. These patterns suggest that diabetes risk is
embedded within broader social determinants of health, which influence
exposure to risk factors and access to care (Hill-Briggs et al., 2021).

Together, these findings motivate a multivariable modelling approach to
estimate adjusted associations and evaluate predictive performance
across diabetes categories.

*'EDI Discussed in Task1'*

## RQ2: The Impact of Socioeconomic Moderation on Cardiovascular Risk

**Outcome Variables:** *HeartDiseaseorAttack* — binary outcome

**Research Question**

> ***To what extent do socioeconomic determinants, specifically income
> and education, moderate the impact of clinical risk factors on the
> prevalence of heart disease, and can the capture of complex feature
> interactions provide superior predictive insights compared to
> traditional additive assumptions?***

**Rationale**

Exploratory analysis identified distinct gradients in cardiovascular
health, where the prevalence of heart disease increases substantially
across older age groups and among individuals with specific clinical
markers. While factors such as high blood pressure, elevated
cholesterol, and body mass index are established universal risk
indicators, their influence is rarely isolated. In real-world clinical
contexts, these physiological markers exist within a broader
socioeconomic framework, suggesting that the "weight" of a clinical risk
factor may be compounded or mitigated by an individual's access to
resources, environmental stressors, and educational background (Roth et
al., 2024).

Traditional analytical frameworks often rely on the assumption that
risks are simply additive, yet biological and social systems frequently
exhibit synergistic effects where the presence of one factor
significantly alters the impact of another. For example, the
cardiovascular strain associated with high BMI may be intensified in
lower-income brackets due to limited access to preventive care or higher
cumulative stress (Collins et al., 2023). By investigating these
conditional relationships, this inquiry shifts the focus from simple
binary classification toward a more nuanced understanding of risk
stratification. This approach allows for the identification of specific
"risk profiles," which are intersections of clinical and social data
that might be overlooked by frameworks that fail to account for
higher-order interactions.

![The Dahlgren-Whitehead Rainbow Model illustrating the layers of health
determinants, providing the theoretical basis for the moderation
analysis of income and education in
RQ2.](https://raw.githubusercontent.com/SHAH-MEER/Modellin-Data/main/image/Rainbow-model-of-the-determinants-of-health.png){#fig-sdoh
fig-align="center" width="85%"}

As visualized in figure, individual clinical risks like blood pressure
and BMI are nested within broader socioeconomic environments. This
framework justifies the shift from simple predictive modelling to an
investigation of how these external layers moderate physiological risk.

**EDI and Ethical Considerations**

The application of predictive modelling in healthcare necessitates a
rigorous focus on transparency and equity. A primary ethical concern is
the risk of "algorithmic bias," where models may inadvertently penalise
or misrepresent specific subgroups if demographic variables are treated
in isolation. By specifically investigating socioeconomic moderators,
this research aims to move beyond a "one-size-fits-all" predictive
approach, ensuring that the influence of income and education is
explicitly understood rather than buried as "noise" within the data.

Furthermore, the interpretability of these complex interactions is a
matter of distributive justice, as clinicians must be able to explain
risk assessments to patients in a way that is both accurate and
contextually relevant. By quantifying how socioeconomic status shifts
the predictive significance of clinical markers, this study helps
identify whether existing healthcare inequalities are being reinforced
or challenged by the data. Ensuring that the model identifies these
non-linear patterns is essential for developing more equitable
preventative strategies that reflect the lived realities of diverse
population groups (Liu et al., 2023).

## RQ3: Socioeconomic and Lifestyle Determinants of Mental Health Burden

**Outcome Variable:** *MentHlth* — number of self-reported mentally
unhealthy days in the past 30 days (0–30)

**Research Question**

> *To what extent do income, education, physical activity, smoking, and
> heavy alcohol consumption predict the number of mentally unhealthy
> days reported in the past 30 days, and does the association between
> physical activity and mental health burden differ across income
> groups?*

**Rationale**

The *MentHlth* variable represents a count of mentally unhealthy days
within the past month (0–30). Exploratory analysis revealed a strongly
right-skewed distribution with a large concentration of zero values,
indicating that many respondents report no mentally unhealthy days while
a smaller subset experience substantial burden. Such distributions
violate assumptions of linear models and often produce overdispersion
and excess zeros, making standard Poisson or logistic approaches
unsuitable.

A Zero-Inflated Negative Binomial (ZINB) model is therefore appropriate
because it separates two processes:

1.  the probability that individuals experience no mental health burden,
    and

2.  the frequency of mentally unhealthy days among those who do report
    burden (Rust et al., 2014; Liu et al., 2023).

The EDA further revealed clear socioeconomic gradients in mental health
outcomes. Respondents in the lowest income categories exhibited
substantially higher prevalence of high mental health burden compared
with those in higher income groups. Physical activity also showed a
consistent inverse relationship with mentally unhealthy days, suggesting
potential behavioural protection.

These patterns motivate a modelling approach that jointly evaluates
socioeconomic position and lifestyle behaviours. In particular, testing
an interaction between income and physical activity allows examination
of whether the protective association of physical activity differs
across socioeconomic strata.

Prior research supports the relevance of these predictors. Socioeconomic
disadvantage has been consistently linked with elevated risk of mental
health disorders and poorer psychological well-being (Kessler et al.,
2010), while analyses of BRFSS data have shown that ZINB regression
provides an effective framework for modelling unhealthy-days indicators
with excess zeros (Rust et al., 2014; Liu et al., 2023).

**EDI / Ethical Considerations**

Socioeconomic gradients in mental health outcomes raise important equity
considerations. If income remains strongly associated with mental health
burden after adjustment for behavioural factors, this suggests that
structural determinants, such as economic insecurity and access to
resources, play a major role in shaping mental health outcomes.
Understanding these relationships can inform more equitable public
health strategies that address structural conditions rather than
focusing solely on individual behaviour.

# 6. Multidomain Determinants and Prediction of Diabetes Status

## 6.1 Modelling Objective

This section addresses the following research question:

> ***To what extent can diabetes status be explained and predicted using
> demographic, socioeconomic, behavioural, and cardiometabolic factors,
> and which predictors contribute most strongly to distinguishing
> between no diabetes, prediabetes, and diabetes?***

The objective is twofold:

1.  **Explanatory** – to estimate the adjusted associations between
    predictor domains and diabetes status.

2.  **Predictive** – to compare the classification performance of
    statistical and machine learning models.

Diabetes status is treated as a **multiclass outcome** (No diabetes /
Prediabetes / Diabetes), requiring modelling approaches suitable for
nominal categorical outcomes.

## 6.2. Outcome Variable

The dependent variable is **diabetes status**, categorised into:

-   **No diabetes**

-   **Prediabetes**

-   **Diabetes.**

Exploratory analysis demonstrated class imbalance, with the majority of
respondents in the no-diabetes category and substantially smaller
proportions in the prediabetes and diabetes groups. This imbalance is
accounted for in model training and evaluation.

## 6.3 Training and Validation

### 6.3.1 **Data Partitioning and Resampling:**

The dataset is partitioned into a 75% training set and a 25% testing set
using stratified sampling on diabetes status. Within the training set,
5-fold stratified cross-validation is implemented to estimate model
performance during development. The held-out test set is reserved for
final evaluation to ensure unbiased performance assessment.

```{r}
set.seed(42)

diabetes_split <- initial_split(
  diab,
  prop = 0.75,
  strata = Diabetes
)

diab_train <- training(diabetes_split)
diab_test  <- testing(diabetes_split)


diabetes_folds <- vfold_cv(
  diab_train,
  v = 5,
  strata = Diabetes
)

# Sanity checks
diab_train |>  count(Diabetes) |> mutate(prop = n/sum(n))
diab_test |> count(Diabetes) |>  mutate(prop = n/sum(n))

```

### 6.3.2 Recipe Definition and Preprocessing Strategy

-   **Outcome specification** – Models `Diabetes` as a three-level
    categorical outcome (No diabetes, Prediabetes, Diabetes).

-   **Near-zero variance removal** – Removes predictors with minimal
    variability to prevent instability and redundant information.

-   **Ordinal scoring** – Converts ordered factors (`Age`, `Education`,
    `Income`, `GenHlth`) into numeric scores while preserving their
    natural ranking. This introduces an assumption of approximately
    equal spacing between adjacent categories; however, the primary
    objective is to retain monotonic structure in a computationally
    efficient representation suitable for both parametric and tree-based
    models.

-   **Dummy encoding** – Transforms remaining nominal predictors into
    binary indicator variables suitable for modelling.

-   **Normalization** – Standardizes numeric predictors (mean = 0, sd
    = 1) to improve optimisation and comparability across models.

-   **SMOTE balancing** – Applies synthetic oversampling to minority
    classes within resampling folds to mitigate severe class imbalance.

```{r}
#| label: recipie-making
set.seed(42)

rec_rq1 <- recipe(Diabetes ~ ., data = diab_train) |>
  step_nzv(all_predictors()) |>
  step_ordinalscore(Age, Education, Income, GenHlth) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(BMI, MentHlth, PhysHlth) |>
  step_smote(Diabetes, over_ratio = 0.5, seed = 42)

print(rec_rq1)
```

### 6.3.3 Model Specification

We now define the candidate models for RQ1 using the **tidymodels**
framework.

Each model is specified independently of preprocessing and will later be
combined with the common recipe using **workflows** and **workflow
sets**.

|  |  |  |  |
|------------------|------------------|------------------|------------------|
| **Model** | **Type** | **Primary Function** | **Key Strengths** |
| **Multinomial Logistic Regression** | Parametric | Linear Baseline | Interpretable log-odds; direct coefficient analysis. |
| **Random Forest** | Ensemble (Bagging) | Non-linear Discovery | Captures complex interactions; robust to multicollinearity. |
| **XGBoost** | Ensemble (Boosting) | Predictive Performance | High efficiency on tabular data; optimized for accuracy. |

Hyperparameters marked `tune()` will be optimized via grid search during
cross-validation.

```{r}
#| label: model-spec

# Multinomial logistic regression
mod_multinom <- 
  multinom_reg() |> 
  set_engine("nnet") |> 
  set_mode("classification")

# Random Forest
mod_rf <- 
  rand_forest(
    trees = 256,
    mtry  = tune(),
    min_n = tune()
  ) |> 
  set_engine("ranger") |> 
  set_mode("classification")

# Gradient Boosted Trees
mod_xgb <- 
  boost_tree(
    trees = 512,
    tree_depth = tune(),
    learn_rate = tune(),
    loss_reduction = tune(),
    mtry = tune(),
    min_n = tune()
  ) |> 
  set_engine("xgboost") |> 
  set_mode("classification")
```

### 6.3.4 Workflowset Construction

We now integrate:

-   The **RQ1 preprocessing recipe**

-   The **candidate model specifications**

    using `workflow_set()`.

Each workflow represents one **model–preprocessing pair**.\
The workflow set enables parallel evaluation of all candidate models
under identical resampling conditions.

```{r}
#| label: wrkflw
#| eval: false
#| include: false

# Individual workflows 
wf_multinom <-
  workflow() |>
  add_recipe(rec_rq1) |>
  add_model(mod_multinom)

wf_rf <-
  workflow() |>
  add_recipe(rec_rq1) |>
  add_model(mod_rf)

wf_xgb <-
  workflow() |>
  add_recipe(rec_rq1) |>
  add_model(mod_xgb)
```

```{r}
#| label: wrkflw-sets

# Workflow set 
wf_set_rq1 <-
  workflow_set(
    preproc = list(rq1_recipe = rec_rq1),
    models  = list(
      multinom = mod_multinom,
      rf       = mod_rf,
      xgb      = mod_xgb))

```

### 6.3.5 Hyperparameter Tuning

**Metrics computed**

Model performance is evaluated using multiple complementary metrics to
account for class imbalance and provide a robust assessment of
predictive performance:

-   **Accuracy** – overall classification rate

-   **Balanced Accuracy** – mean recall across classes, correcting for
    class imbalance

-   **Cohen’s Kappa** – agreement between predictions and observed
    classes beyond chance

-   **F1-score (macro)** – harmonic mean of precision and recall
    averaged across classes

-   **ROC-AUC (multiclass)** – ability of the model to discriminate
    between diabetes categories

```{r}
#| label: metric-set
metrics_diab <-
  metric_set(
    accuracy,
    bal_accuracy,
    kap,
    f_meas,
    roc_auc
  )
```

**Hyperparameter Tuning Strategy**

We employ **grid search** to optimize hyperparameters across candidate
models:

-   **Grid size:** 10 configurations per model, balancing exploration of
    the hyperparameter space with computational feasibility (3 models ×
    10 grids × 5 CV folds = 150 model fits)
-   **Search method:** Grid search over random/Bayesian approaches
    provides systematic coverage of parameter combinations and is
    computationally manageable for this dataset size
-   **Tuning metric:** **Macro-averaged F1 score** serves as the primary
    optimization metric because it gives equal weight to all three
    diabetes classes (No diabetes, Prediabetes, Diabetes), preventing
    models from optimizing solely for majority class performance
-   **Cross-validation:** 5-fold stratified CV estimates out-of-sample
    performance while maintaining class proportions in each fold

The best hyperparameter configuration for each model is selected based
on highest mean Macro F1 across folds.

```{r}
#| label: hp-tune
# set.seed(42)
# ctrl <- control_grid(
#   save_pred = TRUE,
#   save_workflow = FALSE,
#   verbose = TRUE
# )
# 
# wf_set_rq1_fit <-
#   wf_set_rq1 |>
#   workflow_map(
#     "tune_grid",
#     resamples = diabetes_folds,
#     grid = 10,
#     metrics = metrics_diab,
#     control = ctrl
#     )

wf_set_rq1_fit <- readRDS('wf_set_rq1_fit.rds')
```

```{r}
#| include: false

pretty_autoplot <- function(x, metric, ylab, title) {
  autoplot(
    x, 
    metric = metric) +
    labs(
      title = title,
      x = "Workflow rank (best → worst)",
      y = ylab
    ) +
    scale_x_continuous(breaks = pretty_breaks(n = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 0.1))
}
```

```{r}
#| label: eval-results
#| fig-height: 10
#| fig-width: 16

m1 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "roc_auc", 
  "ROC–AUC (multiclass)", 
  "Discrimination (ROC–AUC)")
m2 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "bal_accuracy", 
  "Balanced accuracy", 
  "Class-balanced accuracy")
m3 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "f_meas", 
  "Macro F1-score", 
  "Macro F1 (imbalance-aware)")
m4 <- pretty_autoplot(wf_set_rq1_fit, "kap", "Cohen's kappa", "Agreement beyond chance")

(m1 | m2) / (m3 | m4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**ROC–AUC.**\
ROC–AUC measures a model’s ability to discriminate between the three
diabetes categories across all possible classification thresholds. The
highest ROC–AUC value is achieved by the **multinomial logistic
regression model**, which appears as the top-ranked workflow in this
metric. However, beyond this leading configuration, most of the
remaining high-ranking workflows consist of **XGBoost and random forest
models**, with values clustered between approximately 0.69 and 0.70.
This indicates that while logistic regression achieves the single best
ROC–AUC score, tree-based models provide similarly competitive
discrimination across multiple configurations.

**Balanced Accuracy.**\
Balanced accuracy reflects the average recall across classes, giving
equal importance to each diabetes category. The **multinomial logistic
regression model again appears as the top-ranked workflow**, achieving
the highest balanced accuracy among the candidate models. Several
**XGBoost configurations follow closely**, while random forest models
occupy the middle range of the ranking. Although performance differences
are relatively modest, the results suggest that multiple model types are
capable of achieving comparable class-balanced recall.

**Macro F1 Score.**\
Macro F1 was selected as the primary tuning metric because it balances
precision and recall equally across classes and is therefore appropriate
under class imbalance. In this panel, the **top-ranked workflows are
clearly dominated by XGBoost models**, which achieve substantially
higher macro F1 scores than the other model types. Random forest
configurations perform considerably lower, and the multinomial logistic
regression model appears further down the ranking. This pattern
indicates that boosted trees are better able to balance precision and
recall across the three diabetes categories.

**Cohen’s Kappa.**\
Cohen’s kappa measures agreement between predicted and observed
classifications beyond chance. The highest kappa values are achieved by
**XGBoost configurations**, which occupy several of the top-ranked
positions in this panel. The multinomial logistic regression model
appears among the higher-performing workflows but does not exceed the
best boosted-tree configurations. Random forest models generally show
lower kappa values and appear further down the ranking. Overall, this
metric again indicates stronger agreement for boosted tree models
compared with the other modelling approaches.

### 6.3.6 Extract tuning results

```{r}
#| label: extarct
# Extract tuning results
rf_res  <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_res <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_xgb")

# Select best hyperparameters using Macro F1
best_rf  <- select_best(rf_res, metric = "f_meas")
best_xgb <- select_best(xgb_res, metric = "f_meas")

# Extract workflows
rf_wf       <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_wf      <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_xgb")
multinom_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_multinom")

# Finalise tuned workflows
rf_final_wf  <- finalize_workflow(rf_wf, best_rf)
xgb_final_wf <- finalize_workflow(xgb_wf, best_xgb)

# Multinomial requires no tuning, so it is already final
multinom_final_wf <- multinom_wf
```

```{r}
#| label: final-fit
rf_final_fit       <- fit(rf_final_wf, data = diab_test)
xgb_final_fit      <- fit(xgb_final_wf, data = diab_test)
multinom_final_fit <- fit(multinom_final_wf, data = diab_test)
```

## 6.4 Model Evaluation

```{r}
#| include: false

get_test_preds <- function(fit, model_name, data, truth_col = Diabetes) {
  p_class <- predict(fit, data, type = "class")
  p_prob  <- predict(fit, data, type = "prob")
  bind_cols(
    data |> select({{ truth_col }}),
    p_class,
    p_prob
  ) |>
    mutate(model = model_name)
}
```

```{r}
pred_rf  <- get_test_preds( rf_final_fit,       
                            "Random forest", diab_test, Diabetes)
pred_xgb <- get_test_preds(xgb_final_fit,      
                           "XGBoost",       diab_test, Diabetes)
pred_mn  <- get_test_preds(multinom_final_fit, 
                           "Multinomial",   diab_test, Diabetes)

pred_all <- bind_rows(pred_rf, pred_xgb, pred_mn)

pred_all <- pred_all |>
  rename(.pred_No_diabetes = `.pred_No diabetes`)
```

```{r}
#| label: rq1-test-metrics
#| message: false
#| warning: false
#| include: false

class_metrics <- pred_all |>
  group_by(model) |>
  metric_set(accuracy, bal_accuracy, kap, f_meas)(
    truth = Diabetes,
    estimate = .pred_class)

roc_metrics <- pred_all |>
  group_by(model) |>
  roc_auc(
    truth = Diabetes,
    .pred_No_diabetes,
    .pred_Prediabetes,
    .pred_Diabetes,
    estimator = "macro"
  )

final_metrics <- bind_rows(class_metrics, roc_metrics) |>
  mutate(estimator = if_else(is.na(.estimator), "overall", .estimator)) |>
  select(model, .metric, .estimator, .estimate) |>
  arrange(match(.metric, c("f_meas","bal_accuracy","roc_auc","kap","accuracy")), model)

final_metrics
```

```{r}
#| label: rq1-roc-combined-prep
#| message: false
#| warning: false

roc_all <- pred_all |>
  group_by(model) |>
  roc_curve(
    truth = Diabetes,
    .pred_No_diabetes,
    .pred_Prediabetes,
    .pred_Diabetes
  )
```

### 6.4.1 ROC Curves

**Multiclass ROC-AUC Evaluation**

ROC (Receiver Operating Characteristic) curves for multiclass outcomes
require extending the binary framework. We employ the **one-vs-rest
approach**:

-   Each class (No diabetes, Prediabetes, Diabetes) is treated as a
    separate binary problem against all other classes combined
-   For each class, ROC curves plot True Positive Rate vs False Positive
    Rate across probability thresholds
-   **Area Under Curve (AUC)** summarizes discriminative ability: AUC =
    0.5 indicates random guessing, AUC = 1.0 indicates perfect
    separation
-   Despite class imbalance, ROC-AUC remains informative for comparing
    relative model performance, though it may be optimistic for rare
    classes (Prediabetes, Diabetes)

Higher AUC values indicate better ability to distinguish each diabetes
category from alternatives. Models with consistently high AUC across all
three classes demonstrate robust multiclass discrimination.

```{r}
#| label: rq1-roc-combined
#| message: false
#| warning: false

autoplot(roc_all) +
  labs(
    title = "Multiclass ROC curves (one-vs-rest) — Test set",
    x = "1 − Specificity",
    y = "Sensitivity"
  ) +
  theme(
    legend.position = 'bottom'
  )
```

The multiclass ROC curves provide additional insight into discrimination
performance across the three classes.

For **diabetes** and **no-diabetes**, all models show strong
discriminative ability, with curves well above the diagonal reference
line. Differences between models are relatively small, though the
tree-based models appear slightly stronger in some regions of the curve.

For the **prediabetes class**, however, the multinomial logistic
regression model clearly outperforms both random forest and XGBoost. The
ROC curve for the multinomial model remains consistently above the other
two models, indicating better sensitivity–specificity trade-offs for
detecting this intermediate category.

### 6.4.2 Confusion Matrices

```{r}
#| label: rq1-conf-prep
#| message: false
#| warning: false
#| include: false

pred_rf  <- pred_rf  |> rename(.pred_No_diabetes = `.pred_No diabetes`)
pred_xgb <- pred_xgb |> rename(.pred_No_diabetes = `.pred_No diabetes`)
pred_mn  <- pred_mn  |> rename(.pred_No_diabetes = `.pred_No diabetes`)
plot_cm <- function(df, ttl) {
  cm <- conf_mat(df, truth = Diabetes, estimate = .pred_class)
  autoplot(cm, type = "heatmap") +
    labs(title = ttl, x = "Predicted", y = "True") 
}
```

**Diagnosing Prediabetes Classification Failure**

The confusion matrices reveal a critical pattern: **tree-based models
(Random Forest, XGBoost) completely fail to predict the Prediabetes
class**, collapsing to a binary No diabetes vs Diabetes distinction.
This occurs due to several factors:

-   **Class overlap:** Prediabetes represents an intermediate metabolic
    state sharing features with both No diabetes and Diabetes, creating
    ambiguous decision boundaries
-   **Minority class optimization:** Despite SMOTE augmentation,
    tree-based algorithms optimize splits based on impurity reduction.
    When Prediabetes remains the smallest class, trees preferentially
    split on features that distinguish the dominant classes (No diabetes
    vs Diabetes), ignoring subtle Prediabetes-specific patterns
-   **Cost function limitations:** Standard multi-class loss functions
    (e.g., cross-entropy) do not penalize misclassifying Prediabetes
    into adjacent categories differently than into distant ones, failing
    to capture ordinal structure

**Multinomial logistic regression**, by contrast, maintains some
Prediabetes prediction by directly modeling class probabilities without
hierarchical splitting, though accuracy remains modest. Alternative
approaches (e.g., ordinal regression, cost-sensitive learning, focal
loss) could better handle this intermediate class in future work.

```{r}
#| fig-height: 10
#| fig-width: 16

p_cm_xgb <- plot_cm(pred_xgb, "Confusion matrix — XGBoost")
p_cm_rf  <- plot_cm(pred_rf,  "Confusion matrix — Random forest")
p_cm_mn  <- plot_cm(pred_mn,  "Confusion matrix — Multinomial Logistic Regression")

(p_cm_xgb | p_cm_rf) / p_cm_mn +
  plot_layout(guides = "collect")
```

The confusion matrices reveal notable differences in classification
behaviour across the three models.

The **multinomial logistic regression model** is the only model that
correctly identifies a meaningful number of **prediabetes cases** (150
observations). In contrast, both **XGBoost and random forest completely
fail to predict the prediabetes class**, assigning all observations in
this category to either the no-diabetes or diabetes classes. This
indicates that the tree-based models effectively collapse the problem
into a two-class classification task, ignoring the intermediate
category.

For the **no-diabetes class**, all three models achieve high numbers of
correct predictions due to the dominance of this category in the
dataset. However, the multinomial model shows slightly higher correct
classifications than the boosted tree model.

For the **diabetes class**, all models correctly identify several
thousand cases, though misclassification between diabetes and
no-diabetes remains present across all models.

Overall, the multinomial model demonstrates a **more balanced
classification behaviour**, particularly by retaining predictive
capacity for the minority prediabetes class.

**Overall Assessment**

Although gradient boosting and random forest performed competitively
during cross-validation, the held-out test set evaluation suggests that
**multinomial logistic regression** provides the most balanced
performance across classes. In particular, it is the only model capable
of correctly identifying prediabetes cases while maintaining strong
discrimination for the other two classes.

## 6.5 Model Interpretation

**Triangulation Strategy for Model Interpretation**

To ensure robust identification of diabetes risk factors,
**methodological triangulation** is employed by comparing
interpretations across different analytical approaches:

-   **Multinomial logistic regression (Log-Odds):** Provides parametric
    coefficient estimates assuming linear additive effects.
    Interpretable but limited to linear relationships
-   **SHAP values (XGBoost):** Model-agnostic feature attribution
    capturing non-linear effects and interactions through tree-based
    partitioning
-   **Variable Importance (XGBoost gain):** Measures total loss
    reduction when features are used for splits, reflecting predictive
    utility

**Convergence between methods** strengthens confidence that identified
risk factors are robust across modelling assumptions. **Divergence**
indicates non-linearities, interactions, or violations of parametric
assumptions that only flexible models capture. Where findings conflict,
consistency patterns are prioritized and discrepancies are noted as
areas requiring further investigation.

### 6.5.1 Log-Odds Ratios (Multinomial Logistic Regression)

Multinomial logistic regression estimates the log-odds of belonging to
each diabetes category relative to a reference category. In this
analysis, **No diabetes** serves as the reference class.

Positive coefficients indicate **higher odds of belonging to the
specified diabetes category**, while negative coefficients indicate
**lower odds relative to the reference group**.

```{r}
#| label: rq1-logodds-prep
#| include: false


multinom_coefs <- multinom_final_fit |>
  extract_fit_parsnip() |>
  pluck("fit") |>
  coef()

multinom_coefs_tidy <- multinom_coefs |>
  as.data.frame() |>
  rownames_to_column(var = "y.level") |>
  pivot_longer(
    cols = -y.level, 
    names_to = "term", 
    values_to = "estimate"
  )

multinom_coefs_tidy <- multinom_coefs_tidy |>
  mutate(odds_ratio = exp(estimate))

```

```{r}
#| fig-height: 7
#| fig-width: 14
multinom_coefs_tidy |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = odds_ratio, y = reorder(term, odds_ratio))) +
  geom_point()  +
  geom_vline(xintercept = 1, linetype = "dashed") +
  facet_wrap(~ y.level) +
  labs(
    x = "Odds Ratio",
    y = "Predictor",
    title = "Odds ratios for predictors of diabetes status"
  )
```

**Multinomial Logistic Regression: Odds Ratio (OR) Summary**

*OR \> 1 indicates increased odds; OR \< 1 indicates decreased odds.*

**Synthesis of Findings**

-   **Primary Drivers:** Clinical indicators (Hypertension,
    Hypercholesterolemia, and BMI) represent the most significant risk.
    High blood pressure, in particular, yields a **139% increase** in
    the odds of diabetes ($OR = 2.39$).

-   **The BMI Gradient:** The data shows a clear progression where BMI
    impacts diabetes ($OR = 1.72$) more severely than prediabetes
    ($OR = 1.51$), confirming its role in disease advancement.

-   **Behavioral Mediation:** The lack of independent effect for
    physical activity ($OR = 1.00$) suggests its impact is likely
    **fully mediated** by metabolic variables like BMI and blood
    pressure.

-   **Statistical Anomalies:** The negative association with heavy
    alcohol ($OR = 0.31$) likely reflects **selection bias** (e.g.,
    "sick quitters" or health-conscious adjustments after diagnosis)
    rather than a biological protective effect.

**Conclusion**

Diabetes status in this population is best characterized by a
**multifactorial cardiometabolic profile**. While age and sex are
relevant, clinical markers of obesity and metabolic syndrome remain the
most effective predictors for distinguishing between healthy,
prediabetic, and diabetic individuals.

### 6.5.2 Global SHAP importance

**SHAP Interpretation Methodology**

To complement the parametric log-odds interpretation, we employ **SHAP
(SHapley Additive exPlanations)** to quantify feature contributions in
the XGBoost model. SHAP offers several advantages:

-   **Model-agnostic foundation** grounded in game theory, enabling
    comparison across model types
-   **TreeSHAP algorithm** provides exact Shapley values by leveraging
    XGBoost tree structure (no Monte Carlo approximation needed)
-   **Multiclass compatibility** computes separate contributions for
    each diabetes category (No diabetes, Prediabetes, Diabetes)
-   **Mean absolute SHAP values** reveal global feature importance by
    measuring how much each feature, on average, shifts predicted
    probabilities from baseline

By comparing SHAP importance with logistic regression odds ratios, we
assess method convergence—agreement strengthens confidence in identified
risk factors, while discrepancies highlight non-linearities the
parametric model cannot capture.

```{r}
#| include: false

xgb_booster <- xgb_final_fit |> 
  extract_fit_parsnip() |> 
  pluck("fit")


mold <- extract_mold(xgb_final_fit)

X <- mold$predictors

X_mat <- as.matrix(X)

y <- mold$outcomes

sv <- shapviz(
  xgb_booster,
  X_pred = X_mat,  
  X = X_mat)
```

```{r}
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 14
sv_importance(sv, kind = "bar") +
  scale_fill_discrete(
    labels = c(
      "No diabetes",
      "Prediabetes",
      "Diabetes")
  )
```

> The SHAP importance plot indicates that cardiometabolic indicators
> dominate the predictive structure of the model. High blood pressure,
> general health status, high cholesterol, and BMI exhibit the largest
> mean absolute SHAP contributions across classes, indicating they play
> the strongest role in shifting predicted probabilities of diabetes
> status. Age also contributes meaningfully, while behavioural variables
> such as smoking, fruit consumption, and alcohol use show comparatively
> smaller effects. This pattern closely mirrors the multinomial logistic
> regression results, reinforcing the conclusion that diabetes status in
> this dataset is primarily driven by metabolic and health status
> indicators rather than lifestyle variables alone.

### 6.5.3 Variable Importance

**Variable Importance: Gain Metric**

XGBoost variable importance is computed using the **gain** metric, which
quantifies each feature's contribution to model performance:

-   **Gain** measures the total reduction in the loss function (e.g.,
    multi-class log-loss) achieved whenever a feature is used to split
    nodes across all trees in the ensemble
-   Features with high gain contribute substantially to improving
    predictions by creating splits that better separate diabetes
    categories
-   **Gain vs alternatives:** Unlike split frequency (which counts how
    often a feature is used) or cover (which weights by number of
    observations), gain directly captures predictive impact and effect
    magnitude

Higher gain values indicate features that most effectively reduce
prediction error when incorporated into the model's decision rules. This
metric complements SHAP importance by focusing on in-model utility
rather than marginal contributions to individual predictions.

```{r}
#| fig-height: 7
#| fig-width: 14
xgb_final_fit |> 
  extract_fit_parsnip() |> 
  vip(num_features = 15, 
      geom = "col",
      aesthetics = list(fill = "steelblue",color = 'black', alpha = 0.8)) +
  labs(title = "Variable Importance: XGBoost Model",
       subtitle = "Top 15 predictors based on gain")

```

-   **Consistency with Statistical Findings:** The VIP confirms the
    results from the multinomial log-odds analysis (Section 6.5.1), with
    **HighBP**, **BMI**, and **HighChol** consistently appearing as the
    most influential predictors. This triangulation of methods
    strengthens the argument that these clinical markers are the primary
    determinants of diabetes risk in this dataset.

-   **The "Prediabetes" Paradox:** Despite the high importance scores
    for these top features, the model's struggle to classify the
    "Prediabetes" class suggests a "ceiling effect." While variables
    like **HighBP** and **GenHlth** (General Health) are excellent at
    distinguishing between "Healthy" and "Diabetic" individuals, the
    feature space lacks the specific granularity required to isolate the
    transitionary prediabetic phase.

-   **Lifestyle vs. Clinical Markers:** Interestingly, clinical markers
    (BMI, Blood Pressure) rank significantly higher than lifestyle
    factors such as **PhysActivity** or **Fruits**. This implies that
    while lifestyle choices contribute to the outcome, the physiological
    manifestations (the "clinical profile") are the more immediate and
    powerful predictors for the gradient-boosted trees.

## 6.6 Conclusion: Model Interpretability and Determinants

The comprehensive analysis of diabetes status, conducted through both
statistical and machine learning frameworks, demonstrates that diabetes
risk in this population is best characterized by a multifactorial
cardiometabolic profile. By triangulating the results from multinomial
log-odds, XGBoost Variable Importance, and SHAP values, this study
provides a robust validation of the primary drivers of the disease.

-   **Triangulation of Risk Factors:** There is a clear consensus across
    all analytical methods regarding the primary determinants of health.
    The multinomial logistic regression identified **High Blood
    Pressure** as the strongest categorical predictor, with an
    associated 139% increase in the odds of a diabetes diagnosis
    ($OR = 2.39$). This finding is directly mirrored in the **XGBoost
    Variable Importance Plot (VIP)**, where HighBP ranks as a top-tier
    feature, and is further clarified by the **SHAP summary plot**,
    which illustrates how high values for blood pressure, BMI, and
    cholesterol consistently push the model’s prediction toward the
    diabetic class.

-   **Predictive Nonlinearity and Impact:** While the linear multinomial
    model provides clear odds ratios, the SHAP analysis reveals the
    nuanced, non-linear impacts of features such as **Age** and
    **General Health**. The SHAP distribution reveals a significant
    clustering of high-risk impact for older age groups, suggesting that
    risk does not simply increase linearly, but rather accelerates as
    multiple clinical markers, such as BMI and cholesterol, coincide.

-   **The "Prediabetes" Gap and Structural Challenges:** A critical
    finding of this synthesis is the discrepancy in class identification
    across models. While the machine learning models (XGBoost and Random
    Forest) performed competitively in overall discrimination, they
    effectively collapsed the classification into a binary task, failing
    to isolate the **Prediabetes** class. The SHAP values show a
    significant overlap in the middle of the feature distribution,
    explaining this struggle, the clinical signature of prediabetes in
    this dataset is too similar to the surrounding classes for the
    boosting algorithm to partition effectively. Conversely, the
    multinomial logistic regression demonstrated superior balance,
    maintaining the ability to distinguish the subtle physiological
    transitions inherent in the intermediate prediabetic category.

-   **Socioeconomic and Actionable Clinical Profiles:** Beyond clinical
    indicators, general health perception and age remained significant
    predictors, yet lifestyle factors, such as physical activity and
    diet, showed weaker direct associations than physiological
    manifestations. This implies that while lifestyle choices are
    foundational, they may operate indirectly through their impact on
    measurable markers like BMI and blood pressure. The synthesis
    confirms that a high-risk profile is defined by the intersection of
    these clinical markers, which consistently dominate the top of both
    the VIP and SHAP rankings, identifying them as the most critical
    targets for screening and preventative intervention.

|  |  |  |
|------------------------|------------------------|------------------------|
| **Method** | **Key Strength** | **Primary Top Predictors** |
| **Multinomial Logistic** | Identifies the "Prediabetes" gradient | HighBP, BMI, Age |
| **XGBoost VIP (Gain)** | Highlights overall model efficiency | HighBP, GenHlth, BMI |
| **SHAP Analysis** | Shows direction and impact of features | HighBP, BMI, HighChol |

In summary, the triangulation of results from the **Odds Ratios**,
**SHAP values**, and **XGBoost Variable Importance** provides a robust
validation of the cardiometabolic risk factors. For clinical
application, the multinomial model remains the most "fit-for-purpose"
tool due to its unique capacity to identify individuals in the critical
prediabetic phase.

# Section 7: Socioeconomic Moderation of Cardiovascular Risk

## 7.1. Modelling Objective

This section addresses the following research question:

> *To what extent do socioeconomic determinants, specifically income and
> education, moderate the impact of clinical risk factors on the
> prevalence of heart disease, and can the capture of complex feature
> interactions provide superior predictive insights compared to
> traditional additive assumptions?*

The objective is:

-   **Interaction Analysis** – to investigate whether the predictive
    weight of clinical markers (e.g., BMI, Blood Pressure) shifts
    significantly across different socioeconomic strata.

-   **Predictive Optimization** – to evaluate if non-linear machine
    learning frameworks (XGBoost) outperform traditional linear
    frameworks (Logistic Regression) by capturing these latent
    moderating effects. **Operationalizing Moderation**

The research question asks whether socioeconomic factors (income,
education) moderate the effect of clinical risk factors (high blood
pressure, BMI) on heart disease. In this analysis, moderation is
operationalized as follows:

-   **For Logistic Regression:** Serves as an additive baseline without
    explicit interaction terms. Any moderation effects would require
    post-hoc interaction term inclusion, which is beyond the exploratory
    scope of this model.
-   **For XGBoost:** Interactions are captured implicitly through
    recursive tree partitioning. When the model splits on Income after
    splitting on HighBP (or vice versa), it effectively models
    conditional effects where the impact of one variable depends on the
    value of another.
-   **Evidence criterion:** Moderation will be assessed through SHAP
    interaction analysis and partial dependence plots stratified by SES
    groups. Significant variation in clinical risk effects across
    income/education levels would support the moderation hypothesis.

This approach leverages XGBoost's capacity to detect non-linear
interactions without requiring manual interaction term specification.

## 7.2. Outcome Variable

The dependent variable is **HeartDiseaseorAttack**, a binary outcome
indicating whether the respondent has ever been diagnosed with coronary
heart disease (CHD) or myocardial infarction (MI).

As identified in the exploratory phase, this variable exhibits a notable
class imbalance, with the "No" category significantly outnumbering the
"Yes" category. This necessitates the use of stratified sampling and
specific evaluation metrics, such as the Area Under the ROC Curve
(AUC-ROC) and F1-score, to ensure the model does not become biased
toward the majority class.

## 7.3 Training and Validation

### 7.3.1 Data Partitioning and Resampling

The dataset is partitioned into a 75% training set and a 25% testing set
using stratified sampling on heart disease status. To ensure the models
are robust and to prevent overfitting during the hyperparameter tuning
phase (particularly for the XGBoost model), 5-fold stratified
cross-validation is implemented within the training set. The test set
remains strictly held-out for the final comparative evaluation between
the additive logistic baseline and the interaction-heavy machine
learning model.

```{r}
#| label: partitioning-resampling-rq2
#| message: false

set.seed(42)

heart_split <- initial_split(
  diab, 
  prop = 0.75, 
  strata = HeartDiseaseorAttack
)

heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

heart_folds <- vfold_cv(
  heart_train, 
  v = 5, 
  strata = HeartDiseaseorAttack
)

heart_train |> count(HeartDiseaseorAttack) |> mutate(prop = n/sum(n))
heart_test  |> count(HeartDiseaseorAttack) |> mutate(prop = n/sum(n))
```

### 7.3.2 Recipe Definition and Preprocessing Strategy

**Class Imbalance Strategy**

The outcome exhibits substantial class imbalance (\~9% heart disease
cases). To prevent models from defaulting to majority-class predictions,
we apply **downsampling** with `under_ratio = 2.33`, retaining a 70-30
majority-to-minority ratio rather than full balancing. This approach
balances two objectives: (1) providing sufficient minority-class
representation for model learning (\~30% of downsampled data), and (2)
preserving moderate imbalance to reflect population base rates and avoid
miscalibrated probability estimates. Downsampling is applied within
cross-validation folds to prevent data leakage.

```{r}
#| label: recipe-rq2


heart_recipe <- recipe(HeartDiseaseorAttack ~ ., data = heart_train) |>
  step_nzv(all_predictors()) |>
  step_ordinalscore(Age, Education, Income, GenHlth) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(BMI, MentHlth, PhysHlth) |>
  step_downsample(HeartDiseaseorAttack, under_ratio = 2.33)

print(heart_recipe)
```

### 7.3.3 Model Specification

We define the candidate models for **RQ2** To address the research
question, we compare a traditional parametric baseline against a
high-capacity gradient-boosted ensemble. This allows us to determine if
the inclusion of non-linear feature interactions (socioeconomic
moderation) significantly improves the model's ability to identify
cardiovascular risk.

**The Models**

|  |  |  |  |
|------------------|------------------|------------------|------------------|
| **Model** | **Type** | **Primary Function** | **Key Strengths** |
| **Logistic Regression** | Parametric | Linear Baseline | Provides interpretable Odds Ratios for clinical and SES factors; assumes additive effects. |
| **XGBoost** | Ensemble (Boosting) | Interaction Discovery | Specifically optimized to capture non-linearities and moderation effects between variables. |

***Note***

> The selection of **XGBoost,** instead of Random Forest**,** for this
> research question is informed by its performance in the preceding
> analysis of diabetes status (RQ1). While RQ1 dealt with a multi-class
> outcome, XGBoost is equally well-suited for this binary classification
> of heart disease. Its gradient boosting framework is particularly
> effective at preventing the model from becoming biased toward the
> majority "No" class, ensuring that the clinical and socioeconomic
> patterns of the minority "Yes" class are accurately captured and
> weighted.

```{r}
#| label: model-specs-rq2
lr_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")


xgb_spec <- boost_tree(
  trees = 1024,
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  stop_iter = 20) |>
  set_engine("xgboost") |>
  set_mode("classification")
```

### 7.3.4 Workflow Set Definition

Following the methodology in Section 6, we utilize `workflow_sets` to
ensure both models are evaluated using the identical finalized recipe
and cross-validation folds.

```{r}
#| label: workflow-set-fit

heart_models <- workflow_set(
  preproc = list(base_rec = heart_recipe),
  models = list(logistic = lr_spec, xgboost = xgb_spec),
  cross = TRUE
)
```

### 7.3.5 Metric Selection and Model Execution

**Metrics**

To evaluate the models, we define a robust metric set that accounts for
the class imbalance inherent in the dataset. While **ROC-AUC** provides
a general measure of classifier quality, **PR-AUC** and the **F1-score**
are prioritized to ensure the model effectively identifies the "Yes"
(Heart Disease) cases without excessive false positives.

```{r}
#| label: metrics-rq2
heart_metrics <- metric_set(
  roc_auc,
  pr_auc,
  sens,
  spec,
  f_meas
)
```

**Model Execution and Resampling**

With the metrics defined, we map the `fit_resamples` function across our
workflow set. This process uses the 5-fold cross-validation established
in Section 7.3 to generate stable performance estimates.

```{r}
#| label: workflow-execution-rq2

# cl <- makePSOCKcluster(parallel::detectCores(logical = FALSE))
# registerDoParallel(cl)
# 
# heart_results <- heart_models |>
#   workflow_map(
#     fn = "tune_grid",
#     resamples = heart_folds,
#     grid = 10,
#     metrics = heart_metrics,
#     seed = 42,
#     control = control_grid(
#       save_pred = TRUE,
#       parallel_over = "everything",
#       verbose = TRUE,
#       event_level = "second")
#   )
# 
# stopCluster(cl)
# 
# write_rds(heart_results, "heart_results.rds")
heart_results <- read_rds('heart_results.rds')
```

### 7.3.6 Comparative Performance Visualization

```{r}
#| label: eval-results-rq2
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 14

m1 <- pretty_autoplot(heart_results, 
                      "roc_auc", "ROC–AUC", "Discrimination (ROC–AUC)")
m2 <- pretty_autoplot(heart_results, 
                      "pr_auc", "PR–AUC", "Precision-Recall AUC")
m3 <- pretty_autoplot(heart_results, 
                      "f_meas", "F1-score", "F1 (Imbalance-aware)")
m4 <- pretty_autoplot(heart_results, 
                      "sens", "Sensitivity", "Detection Rate (Recall)")

(m1 | m2) / (m3 | m4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## 7.4 Final Model Extraction and Test Set Evaluation

```{r}
#| label: final-comparison-rq2

best_lr_id <- "base_rec_logistic"
best_xgb_id <- "base_rec_xgboost"

final_lr_wflow <-
  heart_results |>
  extract_workflow(best_lr_id) |>
  finalize_workflow(
    heart_results |>
      extract_workflow_set_result(best_lr_id) |>
      select_best(metric = "pr_auc")
  )

final_xgb_wflow <-
  heart_results |>
  extract_workflow(best_xgb_id) |>
  finalize_workflow(
    heart_results |>
      extract_workflow_set_result(best_xgb_id) |>
      select_best(metric = "pr_auc")
  )
```

```{r}
test_results_lr <- last_fit(
  final_lr_wflow, split = heart_split, metrics = heart_metrics)

test_results_xgb <- last_fit(
  final_xgb_wflow, split = heart_split, metrics = heart_metrics)
```

```{r}
final_comparison <- bind_rows(
  test_results_lr |> collect_metrics() |> mutate(model = "Regularized LR"),
  test_results_xgb |> collect_metrics() |> mutate(model = "Tuned XGBoost")
) |> 
  select(model, .metric, .estimate) |> 
  pivot_wider(names_from = .metric, values_from = .estimate)

print(final_comparison)
```

```{r}
all_preds <- bind_rows(
  test_results_lr |> collect_predictions() |> mutate(model = "Logistic Regression"),
  test_results_xgb |> collect_predictions() |> mutate(model = "XGBoost")
)
```

```{r}
#| fig-height: 6
#| fig-width: 12

cm_lr <- test_results_lr |> 
  collect_predictions() |> 
  conf_mat(HeartDiseaseorAttack, .pred_class) |> 
  autoplot(type = "heatmap") + 
  labs(title = "LR: Predicting everyone has Heart Disease")

cm_xgb <- test_results_xgb |> 
  collect_predictions() |> 
  conf_mat(HeartDiseaseorAttack, .pred_class) |> 
  autoplot(type = "heatmap") + 
  labs(title = "XGBoost: Balanced Predictions")

cm_lr + cm_xgb
```

```{r}
#| fig-height: 7
#| fig-width: 14

final_xgb_fit <- extract_fit_parsnip(test_results_xgb)

final_xgb_fit |> 
  vip(num_features = 20, 
      geom = "col",
      aesthetics = list(fill = "steelblue",color = 'black', alpha = 0.8)) +
  labs(title = "Variable Importance: XGBoost Model",
       subtitle = "Top 15 predictors based on gain")
```

```{r}
xgb_engine <- extract_fit_engine(final_xgb_fit)

test_processed <- bake(
  prep(extract_preprocessor(final_xgb_wflow)),
  new_data = testing(heart_split)
)

X <- test_processed |> select(-HeartDiseaseorAttack) |> as.matrix()

shp <- shapviz(xgb_engine, X_pred = X)

```

```{r}
shp_flipped <- shapviz(-shp$S, X = shp$X, baseline = -shp$baseline)
```

```{r}
#| fig-height: 7
#| fig-width: 14
sv_importance(shp_flipped, kind = "bar", max_display = 15) +
  labs(title = "SHAP Feature Importance", x = "Mean |SHAP Value|")
```

```{r}
#| fig-height: 7
#| fig-width: 14
sv_importance(shp_flipped, kind = "beeswarm", max_display = 15) +
  labs(title = "SHAP Beeswarm Plot")
```

```{r}
#| fig-height: 9
#| fig-width: 16

p1 <- sv_dependence(shp_flipped, v = "HighBP_High", color_var = "Income")
p2 <- sv_dependence(shp_flipped, v = "HighChol_High", color_var = "Income")
p3 <- sv_dependence(shp_flipped, v = "GenHlth", color_var = "Income")
p4 <- sv_dependence(shp_flipped, v = "BMI", color_var = "Income")

(p1 + p2) / (p3 + p4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| fig-height: 9
#| fig-width: 16

p5 <- sv_dependence(shp_flipped, v = "HighBP_High", color_var = "Education")
p6 <- sv_dependence(shp_flipped, v = "HighChol_High", color_var = "Education")
p7 <- sv_dependence(shp_flipped, v = "GenHlth", color_var = "Education")
p8 <- sv_dependence(shp_flipped, v = "BMI", color_var = "Education")

(p5 + p6) / (p7 + p8) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| fig-height: 10
#| fig-width: 15

low_risk_idx <- which.min(rowSums(as.matrix(shp_flipped$S)))
high_risk_idx <- which.max(rowSums(as.matrix(shp_flipped$S)))

w1 <- sv_waterfall(shp_flipped, row_id = high_risk_idx) + 
  labs(title = "Waterfall: Highest Risk Individual",
       subtitle = "Features pushing prediction above baseline",
       x = "SHAP Contribution (log-odds)")

w2 <- sv_waterfall(shp_flipped, row_id = low_risk_idx) + 
  labs(title = "Waterfall: Lowest Risk Individual",
       subtitle = "Features pushing prediction below baseline",
       x = "SHAP Contribution (log-odds)")

f1 <- sv_force(shp_flipped, row_id = high_risk_idx) + 
  labs(title = "Force Plot: Highest Risk Individual",
       subtitle = "Red = increases risk | Blue = decreases risk")

f2 <- sv_force(shp_flipped, row_id = low_risk_idx) + 
  labs(title = "Force Plot: Lowest Risk Individual",
       subtitle = "Red = increases risk | Blue = decreases risk")

(w1 | w2) 

(f1 / f2)
```

```{r}
set.seed(123)
X_sample <- X[sample(nrow(X), 3000), ]
shp_i <- shapviz(xgb_engine, X_pred = X_sample, interactions = TRUE)
```

```{r}
shp_i_flipped <- shapviz(-shp_i$S, X = shp_i$X, baseline = -shp_i$baseline)
```

```{r}

sv_interaction(shp_i_flipped) +
  labs(title = "SHAP Interaction Values",
       subtitle = "Pairwise feature interaction strengths")

sv_dependence(shp_i_flipped, v = "HighBP_High", color_var = "Income", interactions = TRUE) +
  labs(title = "HighBP × Income Interaction")

sv_dependence(shp_i_flipped, v = "HighChol_High", color_var = "Income", interactions = TRUE) +
  labs(title = "HighChol × Income Interaction")

sv_dependence(shp_i_flipped, v = "GenHlth", color_var = "Income", interactions = TRUE) +
  labs(title = "GenHlth × Income Interaction")

sv_dependence(shp_i_flipped, v = "HighBP_High", color_var = "Education", interactions = TRUE) +
  labs(title = "HighBP × Education Interaction")

sv_dependence(shp_i_flipped, v = "GenHlth", color_var = "Education", interactions = TRUE) +
  labs(title = "GenHlth × Education Interaction")
```



























