---
title: "Modelling Diabetes Health Indicators"
subtitle: "MA7442 Modelling Data — Assessment 1 (Task 2)"
author: "Shahmeer Shahzad"
date: "2026-03-02"
format:
  pdf:
    geometry:
      - top=2cm
      - bottom=2cm
      - left=1.5cm
      - right=1.5cm
    documentclass: article
    df-print: kable
    mainfont: "Arial"
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r}
#| label: Loading_Packages
#| message: false
#| warning: false
#| include: false

# Packages
library(tidymodels)
library(tidyverse)
library(skimr)
library(scales)
library(patchwork)
library(viridis)
library(workflowsets)
library(rcompanion)
library(themis)
library(ranger)
library(vip)
library(shapviz)
library(knitr)
library(probably)
library(conflicted)
library(doParallel)
library(pscl)
library(lmtest)
library(marginaleffects)   
library(ggeffects)
library(MASS)

tidymodels_prefer(quiet = TRUE)
conflicts_prefer(yardstick::accuracy)

```

```{r}
#| label: plot-customise
#| include: false

# Plot format and dimension
knitr::opts_chunk$set(dev = "cairo_pdf")
knitr::opts_chunk$set(fig.width = 14, fig.height = 7)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Set global ggplot theme
theme_set(
  theme_classic(base_size = 16) +
    theme(
      plot.title    = element_text(size = 20, face = "bold"),
      axis.title    = element_text(size = 18),
      axis.text     = element_text(size = 14),
      legend.title  = element_text(size = 13),
      legend.text   = element_text(size = 12),
      strip.text    = element_text(face = "bold", size = 14),
      legend.background = element_rect(
        colour    = "black",
        fill      = "white",
        linewidth = 0.5
      )
    )
)
```

```{r}
#| label: palettes-used
#| include: false
pal_sex <- c(
  "Female" = "orange", 
  "Male"   = "darkgreen"   
)

pal_binary <- c(
  "No"  = "#BBBBBB",  
  "Yes" = "#882255"  
)

pal_genhlth <- c(
  "Excellent" = "#117733",  
  "Very good" = "#44AA99",  
  "Good"      = "#DDCC77",  
  "Fair"      = "#EE8866",  
  "Poor"      = "#CC6677"   
)
```

# 1. Introduction

This report investigates the multidomain determinants of chronic disease
outcomes using the Centers for Disease Control and Prevention (CDC)
Diabetes Health Indicators dataset, which comprises 253,680 respondents
and captures demographic, socioeconomic, behavioural, and
cardiometabolic risk factors across the United States population. Three
distinct but interconnected health outcomes are examined: diabetes
status (no diabetes, prediabetes, diabetes), cardiovascular disease
burden, and mental health burden. The dataset's
population-representative structure and inclusion of both modifiable
risk factors and social determinants make it well-suited for
investigating how clinical risk accumulates within socioeconomic
contexts. The analysis applies diverse modelling approaches tailored to
each outcome type—multinomial classification, binary logistic modelling
with interaction terms, and count regression—to demonstrate that chronic
disease risk is shaped not only by physiological indicators but also by
the structural conditions within which these risks emerge and cluster.

## 1.1 Context and Motivation

Chronic disease outcomes rarely emerge from isolated risk factors.
Rather, they reflect the accumulation and interaction of physiological
vulnerabilities, behavioural choices, and structural determinants across
the lifecourse. Understanding how these domains combine to shape health
outcomes requires analytic frameworks that can accommodate different
outcome types and capture both main effects and moderation.

This report addresses three interconnected questions that together
illustrate the multidimensional patterning of chronic disease risk. The
first examines diabetes status as a multiclass outcome, evaluating which
factors distinguish prediabetes from diabetes and healthy status. The
second investigates whether socioeconomic position moderates the
relationship between clinical risk factors and cardiovascular disease,
testing whether physiological risks operate uniformly across social
strata or are amplified or buffered by income and education. The third
shifts focus to mental health burden as a count outcome, examining how
socioeconomic and lifestyle factors shape the number of mentally
unhealthy days reported in the past month.

Together, these questions demonstrate that health outcomes are embedded
within broader social and behavioural contexts. The analytic
contribution of this report lies in its application of methodologically
diverse approaches—each tailored to the structure of the outcome—while
maintaining a coherent focus on the interconnected nature of chronic
disease determinants. Rather than treating each outcome in isolation,
the report illustrates how different facets of health risk are
structured by common underlying gradients in age, socioeconomic
position, and behavioural clustering.

## 1.2 Reflection on Task 1 Feedback

Three areas of feedback informed the revised analytical approach.

**Scope of the EDA.**\
Task 1 examined associations solely in relation to diabetes status,
constraining the diversity of subsequent research questions. Task 2
introduces comprehensive EDA examining cardiovascular indicators,
general health measures, and behavioural variables as standalone
outcomes.

**Research question overlap.**\
Because the initial EDA centred exclusively on diabetes, the research
questions were conceptually similar. In response, RQ1 has been refined,
the previous RQ2 and RQ3 consolidated, and two new non-diabetes research
questions introduced based on the expanded EDA.

**Figure line type.**\
Task 1 used `geom_line` to connect mean values across ordered
categorical age groups, implying unjustified continuity. Age-structured
visualisations in Task 2 use `geom_col` or point-based representations
that preserve ordinal structure without assuming linear progression.

## 1.3 Overview of Research Questions

The report addresses three research questions that span different
outcome types and analytical objectives:

**RQ1: Multidomain Determinants and Prediction of Diabetes Status**\
Examines diabetes status (no diabetes, prediabetes, diabetes) as a
multiclass outcome. Evaluates which demographic, socioeconomic,
behavioural, and cardiometabolic factors distinguish between diabetes
categories and which modelling approaches (multinomial logistic
regression, random forest, XGBoost) best capture these distinctions.

**RQ2: The Impact of Socioeconomic Moderation on Cardiovascular Risk**\
Investigates whether socioeconomic position (income and education)
moderates the relationship between clinical risk factors (high blood
pressure, high cholesterol, BMI) and cardiovascular disease. Tests
whether physiological risks operate uniformly across social strata or
are amplified or buffered by structural determinants.

**RQ3: Socioeconomic and Lifestyle Determinants of Mental Health
Burden**\
Models the number of self-reported mentally unhealthy days in the past
month as a count outcome. Examines how socioeconomic position, physical
activity, smoking, and physical health burden shape mental health
outcomes, with particular attention to whether protective associations
differ across socioeconomic strata.

Together, these questions demonstrate that chronic disease risk is
multidimensional, requiring analytical approaches tailored to outcome
structure while maintaining focus on the interconnected roles of
clinical indicators, behavioural factors, and social determinants.

## 1.4 Analytical Approach

The diversity of outcome types across the three research questions
necessitates distinct modelling strategies, each tailored to the
structure and distributional properties of the dependent variable.

For **RQ1** (diabetes status), the multiclass categorical outcome
requires methods capable of distinguishing between three unordered
categories. Multinomial logistic regression provides interpretable
baseline estimates of adjusted associations, while tree-based ensemble
methods (random forest and XGBoost) allow flexible capture of non-linear
relationships and variable interactions. Model performance is evaluated
using macro F1-score to account for class imbalance, and interpretation
draws on both regression coefficients and SHAP values to identify
dominant predictors.

For **RQ2** (cardiovascular disease), the binary outcome is modelled
using logistic regression with explicit interaction terms between
socioeconomic variables (income, education) and clinical risk factors
(high blood pressure, high cholesterol, BMI). This approach tests
whether the effect of physiological risk varies systematically across
socioeconomic strata. Predicted probabilities and marginal effects are
computed to illustrate moderation patterns.

For **RQ3** (mental health burden), the count-based outcome (number of
mentally unhealthy days) exhibits overdispersion and excess zeros,
making negative binomial regression or zero-inflated models appropriate.
The analysis evaluates whether socioeconomic and lifestyle factors
influence both the likelihood of reporting any mentally unhealthy days
and the intensity of burden among those affected.

This methodological diversity reflects the principle that analytic
strategy should align with outcome structure and research objectives,
rather than applying a single approach uniformly across all questions.

------------------------------------------------------------------------

# 2. Data Preparation

The dataset and all variable recoding steps are identical to those
described in Task 1 and are reproduced in the setup chunk below (Hidden)
to ensure a fully self-contained, reproducible document. No additional
transformations are required for the EDA in Section 3. All 253,680
records are retained; no missing values were present.

```{r}
#| label: data-import-and-recode
#| message: false
#| warning: false
#| include: false

diab <- read_csv("dataset1.csv")

diab <- diab |>
  mutate(
    Diabetes = factor(
      Diabetes_012,
      levels = c(0, 1, 2),
      labels = c("No diabetes", "Prediabetes", "Diabetes")
    ),
    across(
      c(CholCheck, Smoker, Stroke, HeartDiseaseorAttack,
        PhysActivity, Fruits, Veggies, HvyAlcoholConsump,
        AnyHealthcare, NoDocbcCost, DiffWalk),
      function(x) factor(x, levels = c(0, 1), labels = c("No", "Yes"))
    ),
    across(
      c(HighBP, HighChol),
      function(x) factor(x, levels = c(0, 1), labels = c("Not High", "High"))
    ),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    GenHlth = factor(
      GenHlth,
      levels  = 1:5,
      labels  = c("Excellent", "Very good", "Good", "Fair", "Poor"),
      ordered = TRUE
    ),
    Age = factor(
      Age,
      levels  = 1:13,
      labels  = c("18–24", "25–29", "30–34", "35–39", "40–44",
                  "45–49", "50–54", "55–59", "60–64", "65–69",
                  "70–74", "75–79", "80+"),
      ordered = TRUE
    ),
    Education = factor(
      Education,
      levels  = 1:6,
      labels  = c("Kindergarten or less", "Grades 1–8",
                  "Grades 9–11", "High school graduate",
                  "Some college / technical", "College graduate"),
      ordered = TRUE
    ),
    Income = factor(
      Income,
      levels  = 1:8,
      labels  = c("< $10k", "$10k–$14k", "$15k–$19k", "$20k–$24k",
                  "$25k–$34k", "$35k–$49k", "$50k–$74k", "$75k+"),
      ordered = TRUE
    )
  ) |>
  select(-Diabetes_012)
```

------------------------------------------------------------------------

# 3. Additional Exploratory Data Analysis: Non-Diabetes Outcomes

The following analyses explore relationships among health, lifestyle,
and socioeconomic variables *independently of diabetes status*. The goal
is to characterise the joint structure of these variables, in particular
to understand what shapes self-reported general health, mental health
burden, and functional limitation, and to motivate the non-diabetes
research questions introduced in Section 4.

## 3.1 Cardiovascular Risk Indicators: Demographic, Socioeconomic and Behavioural Gradients

This section explores the population structure and correlates of key
cardiovascular risk indicators, namely high blood pressure, high
cholesterol, and stroke. The analysis moves beyond diabetes to examine
how these outcomes are patterned across age, sex, socioeconomic
position, and behavioural factors. Emphasis is placed on identifying
demographic concentration, socioeconomic gradients, and clustering with
related health behaviours. The objective is to establish structural
regularities in the data that may inform subsequent modelling and risk
interpretation.

```{r}
#| label: age-sex-pyramid-prep
#| include: false

bp_pyramid <- diab |>
  count(HighBP, Age, Sex) |>
  group_by(HighBP, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

chol_pyramid <- diab |>
  count(HighChol, Age, Sex) |>
  group_by(HighChol, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_))

stroke_pyramid <- diab |>
  filter(!is.na(Stroke), !is.na(Age), !is.na(Sex)) |>
  count(Stroke, Age, Sex) |>
  group_by(Stroke, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

```

### 3.1.1 Age–Sex Distribution by High Blood Pressure Status

To understand how cardiovascular risk is demographically structured, the
age and sex distribution of individuals with and without high blood
pressure is examined first. This establishes whether cardiovascular
burden is concentrated in specific demographic subgroups or distributed
broadly across the population.

```{r}
#| label: highbp-age-sex-pyramid
#| fig.height: 7
#| fig.width: 14

ggplot(bp_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ HighBP, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by high blood pressure status",
    x     = "Age group",
    y     = "Percent within blood pressure group",
    fill  = "Sex"
  )
```

### 3.1.2 Age–sex distribution by **high cholesterol** status

The age and sex distribution is examined for high cholesterol status to
assess whether demographic patterning is similar to that observed for
high blood pressure or whether distinct age-sex profiles emerge.

```{r}
#| label: highchol-age-sex-pyramid
#| fig-height: 7
#| fig-width: 14

ggplot(chol_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ HighChol, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by high cholesterol status",
    x     = "Age group",
    y     = "Percent within cholesterol group",
    fill  = "Sex")
```

### 3.1.3 Age–sex distribution by heart disease history

Heart disease or myocardial infarction represents a more severe
cardiovascular outcome than hypertension or hypercholesterolaemia. The
demographic distribution is examined to determine whether heart disease
exhibits a similar age-sex structure or shows more pronounced
concentration in specific groups.

```{r}
#| label: hrtdis-age-sex-pyramid
#| fig-height: 7
#| fig-width: 14

hrt_dis_pyramid <- diab |>
  count(HeartDiseaseorAttack, Age, Sex) |>
  group_by(HeartDiseaseorAttack, Sex) |>
  mutate(
    percent = 100 * n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    percent_signed = case_when(
      Sex == "Male"   ~ -percent,
      Sex == "Female" ~  percent,
      TRUE            ~ NA_real_
    )
  )

ggplot(hrt_dis_pyramid,
       aes(x = Age, y = percent_signed, fill = Sex)) +
  geom_col(width = 0.9, colour = "white") +
  coord_flip() +
  facet_wrap(~ HeartDiseaseorAttack, nrow = 1) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Age–sex distribution by Heart Disease/Attack history",
    x     = "Age group",
    y     = "Percent within Heart Disease history group",
    fill  = "Sex"
  )
```

> Across all three outcomes, high blood pressure, high cholesterol, and
> stroke, the dominant structural feature is age concentration. Each
> “Yes” panel is heavily skewed toward the 55–69 and older age bands,
> while representation below age 40 is minimal. This confirms that
> cardiovascular burden in this dataset is fundamentally age-structured
> rather than evenly distributed across the life course.
>
> However, the degree of age concentration differs meaningfully between
> outcomes. High blood pressure and high cholesterol show similar
> mid-to-late life clustering, with substantial presence beginning in
> the 50–54 band and peaking around 60–69. Stroke, by contrast, exhibits
> a far sharper right-skew: the “Yes” group is overwhelmingly
> concentrated in older age bands, with negligible representation in
> younger adults. This suggests stroke is not merely age-associated but
> strongly age-dependent relative to the other indicators.
>
> Sex differences are comparatively modest within age strata for high
> blood pressure and high cholesterol. Once age is conditioned upon,
> male–female distributions appear broadly parallel, suggesting that sex
> operates as a secondary structuring factor relative to age. In the
> heart disease group, however, there is a slight female predominance in
> the oldest age bands, which may reflect differential survival into
> advanced age rather than a pure sex-specific risk gradient.
>
> Importantly, these pyramids depict the **composition within outcome
> groups**, not prevalence across the population. They therefore
> highlight demographic concentration rather than risk probability. The
> consistency across high blood pressure and high cholesterol suggests
> shared cardiometabolic ageing processes, whereas heart disease
> displays a more extreme demographic compression, indicating it may
> represent the culmination of accumulated vascular risk rather than an
> early-stage cardiovascular condition.
>
> Taken together, the figures indicate that age is the primary axis of
> cardiovascular risk patterning in this sample. Any socioeconomic or
> behavioural gradients observed later must therefore be interpreted
> within the context of this strong underlying demographic structure.

```{r}
#| label: ses-heatmap
#| include: false

bp_ses <- diab |>
  mutate(
    bp_binary = if_else(HighBP == "High", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(prev_bp = mean(bp_binary),
            n = n(),
    .groups = "drop")

chol_ses <- diab |>
  mutate(chol_bin = if_else(HighChol == "High", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(
    prev_chol = mean(chol_bin),
    n = n(),
    .groups = "drop")

stroke_ses <- diab |>
  mutate(
    heart_d_bin = if_else(HeartDiseaseorAttack == "Yes", 1, 0)) |>
  group_by(Education, Income) |>
  summarise(prev_heartdis = mean(heart_d_bin),n = n(),
    .groups = "drop")
```

### 3.1.4 Joint SES heatmap of High Blood Pressure prevalence

To examine whether socioeconomic position structures cardiovascular risk
independently of demographic factors, prevalence is calculated across
all combinations of education and income categories. This
two-dimensional approach reveals whether education and income operate as
distinct gradients or merely proxy for one another.

```{r}
#| label: highbp-ses-heatmap
#| fig.height: 8
#| fig.width: 15

bp_ses |> 
  ggplot(aes(x = Income, y = Education, fill = prev_bp)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_bp, accuracy = 1)),
    colour   = "white",
    fontface = "bold",
    size     = 4.5) +
  scale_fill_gradient(
    low  = "#F1EEF6",
    high = "#882255",
    name = "Prevalence\nHigh BP",
    labels = percent_format(accuracy = 1)) +
  labs(title = "Prevalence of high blood pressure by education and income",
    x     = "Income band",
    y     = "Education level")
```

### 3.1.5 Joint SES heatmap of High Cholesterol prevalence

The same socioeconomic analysis is applied to high cholesterol to assess
whether the patterning differs from hypertension or follows a similar
structure.

```{r}
#| label: highchol-ses-heatmap
#| fig.height: 8
#| fig.width: 15

chol_ses |>
  ggplot(aes(x = Income, y = Education, fill = prev_chol)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_chol, accuracy = 1)),
    colour   = "white",
    fontface = "bold",
    size     = 3.5
  ) +
  scale_fill_gradient(
    low    = "#F1EEF6",      
    high   = "#882255",      
    name   = "Prevalence\nHigh cholesterol",
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Prevalence of high cholesterol by education and income",
    x     = "Income band",
    y     = "Education level"
  )
```

### 3.1.6 Joint SES heatmap of Heart Disease prevalence

Finally, heart disease prevalence is examined across the same
socioeconomic grid. Given the lower absolute prevalence of heart
disease, this analysis assesses whether socioeconomic gradients remain
detectable or become obscured by smaller cell counts.

```{r}
#| label: heartdis-ses-heatmap
#| fig.height: 8
#| fig.width: 15

stroke_ses |>
  ggplot(aes(x = Income, y = Education, fill = prev_heartdis)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(
    aes(label = percent(prev_heartdis, accuracy = 1)),
    colour = "white",
    fontface = "bold",
    size = 3.5) +
  scale_fill_gradient(
    low = "#F1EEF6",      
    high = "#882255",
    name = "Prevalence\nHeart Disease history",
    labels = percent_format(accuracy = 1)) +
  labs(title = "Prevalence of Heart Attack history by education and income",
        x = "Income band",
        y = "Education level")
```

> Across all three outcomes, a clear socioeconomic gradient is evident.
> Prevalence is consistently highest in the lowest education and lowest
> income strata and declines with increasing educational attainment and
> income. The lower-left cells (least educated, lowest income) exhibit
> the greatest burden, while college graduates in the highest income
> bands show the lowest prevalence. This pattern indicates cumulative
> socioeconomic disadvantage rather than a single-dimension effect.
>
> High blood pressure and high cholesterol display particularly strong
> and stable two-dimensional gradients. Within nearly every education
> category, prevalence decreases progressively as income rises.
> Likewise, within each income band, higher education corresponds to
> lower prevalence. The monotonic pattern across both axes suggests that
> both income and education independently structure cardiometabolic
> risk, rather than one merely proxying for the other.
>
> Heart disease exhibits the same directional gradient but at
> substantially lower absolute prevalence and with more variability
> across cells. Some low-education, low-income cells show markedly
> elevated heart disease prevalence (e.g., 20-27%), while higher SES
> cells approach very low levels. The noisier pattern likely reflects
> smaller counts and the relative rarity of heart disease compared with
> hypertension and hypercholesterolaemia, but the overall concentration
> in socioeconomically disadvantaged groups remains clear.
>
> Importantly, these are crude prevalence estimates. As shown in the
> age–sex pyramids, lower socioeconomic strata contain
> disproportionately older individuals, and age is a dominant
> determinant of all three outcomes. Therefore, part of the observed SES
> gradient likely reflects demographic composition. However, the
> persistence of a consistent gradient across both income and education
> dimensions suggests that socioeconomic position is not merely a proxy
> for age but a meaningful structural correlate of cardiovascular
> burden.

### 3.1.7 Prevalence of cardiovascular outcomes by accumulated risk factors

To assess whether cardiometabolic risk factors operate independently or
cumulatively, cardiovascular prevalence is examined across groups
defined by the number of concurrent risk factors present (obesity,
physical inactivity, smoking, and diabetes). This establishes whether
the joint burden of multiple risk exposures amplifies cardiovascular
risk beyond what would be expected from isolated factors.

```{r}
#| label: cvd-obesity-physact-prep
#| include: false
diab_risk <- diab |>
  mutate(
    Obese  = if_else(BMI >= 30, 1L, 0L),
    Inactive = if_else(PhysActivity == "No", 1L, 0L),
    Smoker_bin = if_else(Smoker == "Yes", 1L, 0L),
    Diabetes_bin = if_else(Diabetes != "No diabetes", 1L, 0L),
    risk_count = Obese + Inactive + Smoker_bin + Diabetes_bin,
    risk_group = case_when(
      risk_count == 0 ~ "0",
      risk_count == 1 ~ "1",
      risk_count == 2 ~ "2",
      risk_count == 3 ~ "3",
      risk_count >= 4 ~ "4+"
    ),
    risk_group = factor(risk_group, levels = c("0","1","2","3","4+"))
  )

cvd_long <- diab_risk |>
  select(risk_group, HighBP, HighChol, HeartDiseaseorAttack) |>
  pivot_longer(
    cols = c(HighBP, HighChol, HeartDiseaseorAttack),
    names_to = "Outcome",
    values_to = "Status"
  ) |>
  mutate(
    outcome_event = case_when(
      Outcome == "HighBP" ~ Status == "High",
      Outcome == "HighChol" ~ Status == "High",
      Outcome == "HeartDiseaseorAttack" ~ Status == "Yes",
      TRUE ~ NA
    ),
    Outcome = recode(
      Outcome,
      HighBP = "High blood pressure",
      HighChol = "High cholesterol",
      HeartDiseaseorAttack = "Heart Disease or Attack history")) |>
  group_by(Outcome, risk_group) |>
  summarise(
    prop = mean(outcome_event, na.rm = TRUE),
    .groups = "drop")
```

```{r}
#| label: cvd-obesity-physact
#| fig.height: 6
#| fig.width: 14


ggplot(cvd_long, aes(x = risk_group, y = prop, fill = risk_group)) +
  geom_col(width = 0.7) +
  facet_wrap(~ Outcome, nrow = 1) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  scale_fill_brewer(palette = "Purples", name = "Risk\ncount") +
  labs(
    title = "Prevalence of cardiovascular outcomes by cardiometabolic risk factor count",
    x     = "Number of risk factors present\n(obesity, inactivity, smoking, diabetes)",
    y     = "Proportion with outcome"
  )
```

> Cardiovascular outcomes show a clear **dose–response relationship with
> the accumulation of cardiometabolic risk factors**. For both **high
> blood pressure** and **high cholesterol**, prevalence increases
> steadily as additional risk factors are present. Individuals with no
> risk factors show relatively low prevalence (around one quarter to one
> third), while those with **four or more risk factors exhibit very high
> prevalence**, exceeding roughly **70–80%** for both conditions. This
> pattern indicates that these outcomes are strongly associated with the
> combined burden of obesity, physical inactivity, smoking, and
> diabetes.
>
> **Heart disease or attack history follows the same upward trend**, but
> at substantially lower levels overall. Even among individuals with
> multiple risk factors, heart disease prevalence remains below roughly
> **25–30%**, reflecting the relative rarity of this severe outcome
> compared with hypertension and hypercholesterolaemia in the
> population.
>
> Overall, the figure demonstrates that **cardiometabolic risk factors
> act cumulatively rather than independently**. As risk exposures
> accumulate, the likelihood of cardiovascular conditions, particularly
> hypertension and high cholesterol, increases markedly. This cumulative
> patterning suggests that multivariable models examining cardiovascular
> outcomes should account for the joint burden of metabolic and
> behavioural risk factors rather than treating them as isolated
> predictors.

### 3.1.8 Joint blood pressure and cholesterol profiles across age groups

Having established that both high blood pressure and high cholesterol
exhibit strong age gradients and socioeconomic patterning, this analysis
examines whether these two conditions occur independently or cluster
within individuals as they age. Understanding whether dual elevation
emerges as a distinct profile or merely reflects independent
accumulation of separate risks informs whether cardiovascular modelling
should treat these as additive effects or interacting pathways.

```{r}
#| label: highchol-highbp-age-prep
#| include: false

bp_chol_age <- diab |>
  count(Age, HighBP, HighChol) |>
  group_by(Age) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

```

```{r}
#| label: highchol-highbp-age
#| fig-width: 14
#| fig-height: 6

ggplot(bp_chol_age,
       aes(x = Age, y = prop, fill = interaction(HighBP, HighChol,
                                                  sep = " / Chol: "))) +
  geom_col(position = "fill", width = 0.85) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(
    title = "Joint blood pressure and cholesterol status by age group",
    x     = "Age group",
    y     = "Proportion within age group",
    fill  = "BP / Cholesterol"
  ) 
```

> The distribution of cardiovascular profiles shifts systematically with
> age. Younger groups are overwhelmingly characterised by normal blood
> pressure and cholesterol, whereas the proportion with dual elevation
> (high/high) increases steadily from midlife onward and becomes the
> largest single cardiovascular profile in older age bands, representing
> approximately 40% of individuals aged 70+. This pattern reflects
> progressive cardiometabolic accumulation rather than isolated risk
> emergence.
>
> Importantly, the high/high category expands more rapidly than either
> single-risk category, indicating clustering rather than independence
> of these conditions. By ages 65+, dual dysregulation accounts for a
> substantial share of the population within each age group, suggesting
> convergence of vascular risk factors with ageing.
>
> The pattern implies that hypertension and hypercholesterolaemia are
> not parallel but interacting processes within cardiovascular ageing,
> reinforcing their role as intermediate markers in the pathway toward
> more severe vascular outcomes. For cardiovascular modelling, this
> clustering suggests that interaction terms between blood pressure and
> cholesterol status may better capture risk than treating them as
> independent additive effects.

### 3.1.9 Co-occurrence of Hypertension and High Cholesterol

The previous analysis demonstrated that high blood pressure and high
cholesterol cluster within age groups more than would be expected if
they occurred independently. To quantify the overall population-level
co-occurrence, this section examines the joint prevalence of both
conditions across all respondents. This establishes the baseline
distribution of cardiovascular risk profiles and determines whether dual
elevation represents a distinct subgroup or simply reflects overlapping
but independent conditions.

```{r}
#| label: highbp-highchol-joint
#| fig.height: 8
#| fig.width: 12

diab |>
  count(HighBP, HighChol) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = HighBP, y = HighChol, fill = prop)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(
    aes(label = percent(prop,accuracy = 0.1)),
    size     = 6,
    fontface = "bold",
    colour   = "white"
  ) +
  labs(
    title = "Joint prevalence of high blood pressure and high cholesterol",
    x     = "High blood pressure",
    y     = "High cholesterol"
  ) +
  theme(
    legend.position = 'none'
  )
```

> Approximately **40.2%** of the sample has neither high blood pressure
> nor high cholesterol. The remaining **59.8%** have at least one of the
> two conditions, indicating that cardiometabolic risk is common in this
> population.
>
> The largest single elevated profile is **concurrent high blood
> pressure and high cholesterol (25.5%)**, which exceeds either isolated
> condition (17.4% BP only; 16.9% cholesterol only). This suggests
> meaningful clustering rather than independent occurrence of the two
> risk factors.
>
> Overall, hypertension and hypercholesterolaemia appear strongly
> co-patterned, with dual elevation more common than either
> single-condition state, reinforcing their joint role in cardiovascular
> risk stratification. This co-occurrence pattern supports modelling
> strategies that account for interaction effects between blood pressure
> and cholesterol when predicting more severe cardiovascular outcomes,
> rather than assuming their effects are purely additive.

### 3.1.10 Heart disease prevalence by blood pressure and cholesterol status

Having established that high blood pressure and high cholesterol
co-occur more frequently than would be expected under independence, this
section examines whether their joint presence is associated with
elevated risk of more severe cardiovascular outcomes. Specifically,
heart disease or myocardial infarction prevalence is compared across the
four BP/cholesterol profiles to test whether dual elevation represents a
meaningful risk amplification.

```{r}
#| label: stroke-bp-chol-prep
#| include: false
heart_bpchol <- diab |>
  count(HighBP, HighChol, HeartDiseaseorAttack) |>
  group_by(HighBP, HighChol) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  filter(HeartDiseaseorAttack == "Yes") |>
  mutate(
    BP_Chol = paste0("BP: ", HighBP, "\nChol: ", HighChol)
  )

```

```{r}
#| label: hrt-bp-chol
#| fig-width: 14
#| fig-height: 7

ggplot(heart_bpchol,
       aes(x = BP_Chol, y = prop, fill = prop)) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = percent(prop, accuracy = 0.1)),
    vjust = -0.4,
    size = 5.5,
    fontface = "bold"
  ) +
  scale_y_continuous(
  labels = percent_format(accuracy = 0.1)) +
  scale_fill_viridis_c(option = "C", direction = -1, guide = "none") +
  labs(
    title = "Heart disease or attack prevalence by blood pressure and cholesterol status",
    x     = "Blood pressure / cholesterol profile",
    y     = "Proportion with heart disease or attack history"
  )
```

> Heart disease prevalence increases monotonically across cardiovascular
> profiles. Among individuals with neither high blood pressure nor high
> cholesterol, heart disease prevalence is approximately **2.8%**. This
> rises to **7.2%** with isolated high cholesterol, **9.7%** with
> isolated high blood pressure, and **21.1%** among those with both
> conditions elevated. The prevalence in the dual-high group is more
> than double that observed with isolated high blood pressure,
> suggesting amplification rather than simple addition of risks.
>
> The gradient indicates cumulative vascular burden, with blood pressure
> appearing more strongly associated with heart disease than cholesterol
> alone. The sharp elevation in the dual-high group suggests that
> interaction effects between blood pressure and cholesterol may better
> capture cardiovascular risk than treating them as independent additive
> predictors. This pattern motivates the inclusion of BP×cholesterol
> interaction terms in subsequent cardiovascular disease modelling
> (RQ2).

### 3.1.11 Joint co-occurrence profiles of high blood pressure, high cholesterol, and stroke

```{r}
#| label: joint-prep
#| include: false

joint <- diab |>
  summarise(
    `HighBP & HighChol` = mean(HighBP == "High" & HighChol == "High"),
    `HighBP & HeartDisease` = mean(HighBP == "High" & HeartDiseaseorAttack == "Yes"),
    `HighChol & HeartDisease` = mean(HighChol == "High" & HeartDiseaseorAttack == "Yes"),
    `HighBP only` = mean(
      HighBP == "High" &
      HighChol != "High" &
      HeartDiseaseorAttack != "Yes"
    ),
    `HighChol only` = mean(
      HighChol == "High" &
      HighBP != "High" &
      HeartDiseaseorAttack != "Yes"
    ),
    `HeartDisease only` = mean(
      HeartDiseaseorAttack == "Yes" &
      HighBP != "High" &
      HighChol != "High"
    ),
    `All three` = mean(
      HighBP == "High" &
      HighChol == "High" &
      HeartDiseaseorAttack == "Yes"
    )
  ) |>
  pivot_longer(everything(), names_to = "Profile", values_to = "Proportion")

max_prop <- max(joint$Proportion)
```

```{r}
#| fig-height: 7
#| fig-width: 14

ggplot(joint, aes(x = reorder(Profile, Proportion), y = Proportion, fill = Proportion)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = percent(Proportion, accuracy = 0.1)),
    hjust = -0.15,
    size = 4.5
  ) +
  coord_flip() +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, max_prop * 1.15)
  ) +
  scale_fill_viridis_c(option = "D", direction = -1, guide = "none") +
  labs(
    title = "Co-occurrence profiles for HighBP, HighChol, and heart disease/attack history",
    x = NULL,
    y = "Proportion of full sample"
  )
```

### 3.1.12 Association Strength - Cramér's V

To identify which demographic, socioeconomic, behavioural, and health
factors show the strongest associations with cardiovascular outcomes,
Cramér's V is calculated for each predictor-outcome pair. This
effect-size measure (ranging 0-1) quantifies association strength for
categorical variables, enabling direct comparison across predictors with
different scales and distributions. The analysis establishes which
factors warrant prioritisation in subsequent modelling and whether
different cardiovascular outcomes share common predictive structures or
require distinct variable selection strategies.

```{r}
#| label: cvd-binary-indicators-prep
#| include: false
nondm_outcomes <- c("HighBP", "HighChol", "HeartDiseaseorAttack")

pred_vars <- c("Diabetes", "Age", "Sex", "Income", "Education",
               "PhysActivity", "Smoker", "BMI_cat",
               "DiffWalk", "Stroke",
               "AnyHealthcare", "NoDocbcCost",
               "GenHlth", "HvyAlcoholConsump")

diab_cv <- diab |>
  mutate(
    BMI_cat = cut(
      BMI,
      breaks = c(0, 18.5, 25, 30, Inf),
      labels = c("Underweight", "Normal", "Overweight", "Obese"),
      right = FALSE
    )
  )

cv_nondm <- map_dfr(nondm_outcomes, function(outcome) {
  map_dfr(pred_vars, function(pred) {
    v <- cramerV(table(diab_cv[[outcome]], diab_cv[[pred]], useNA = "no"))
    tibble(
      Outcome = outcome,
      Predictor = pred,
      CramersV = as.numeric(v)
    )
  })
}) |>
  mutate(
    Outcome = recode(
      Outcome,
      HighBP = "High Blood Pressure",
      HighChol = "High Cholesterol",
      HeartDiseaseorAttack = "Heart Disease/Attack"
    )
  )
```

```{r}
#| label: cvd-binary-indicators
#| fig-height: 9
#| fig-width: 14
ggplot(cv_nondm,aes(x = reorder(Predictor, CramersV),
                    y = CramersV,
                    fill = Outcome)) +
  geom_col(position = position_dodge(width = 0.75),
           width = 0.7) +
  geom_text(aes(label = round(CramersV, 2)),
            position = position_dodge(width = 0.75),
            hjust = -0.1,
            size = 3.2) +
  coord_flip() +
  expand_limits(y = max(cv_nondm$CramersV) + 0.03) +
  scale_fill_manual(
    values = c(
      "High Blood Pressure" = "#2c7bb6",
      "High Cholesterol" = "#f46d43",
      "Heart Disease/Attack" = "#1a9641")) +
  scale_y_continuous(limits = c(0, 0.45)) +
  labs(
    title = "Association strength with Binary outcomes (Cramér's V)",
    x = NULL,
    y = "Cramér's V",
    fill = "Outcome") +
  theme(legend.position = "bottom")
```

> This chart highlights how different risk factors weigh against
> cardiovascular outcomes using **Cramér’s V** (a measure of association
> where higher values indicate a stronger relationship).
>
> The data reveals two distinct "profiles" of disease:
>
> 1.  **Metabolic & Age-Driven (High BP & Cholesterol)**
>
> Hypertension and High Cholesterol show strongest associations with
> **systemic, long-term factors**.
>
> -   **Primary Associations:** **Age** ($V \approx 0.35$) and **General
>     Health** ($V \approx 0.30$) exhibit the strongest relationships.
>
> -   **Secondary Associations:** Notable links to **Diabetes** and
>     **BMI**, consistent with their roles as markers of general
>     metabolic dysregulation.
>
> -   **Takeaway:** These conditions follow a broad, population-wide
>     risk gradient structured by age and overall physical state,
>     indicating that age will likely dominate as a predictor in
>     multivariable models.
>
> 2.  **Clinical & Pathological (Heart Disease/Attack)**
>
> Heart disease or attack history follows a different logic, focusing
> more on **existing cardiovascular damage** than gradual metabolic
> ageing alone.
>
> -   **Primary Drivers:** **General Health** ($V \approx 0.27$) and
>     **Diabetes** ($V \approx 0.27$) are strongest, followed by **Age**
>     ($V \approx 0.23$), **Walking Difficulty** ($V \approx 0.21$), and
>     **Stroke history** ($V \approx 0.20$).
>
> -   **Key Difference:** Age is significantly less dominant for heart
>     disease than for blood pressure or cholesterol. BMI shows almost
>     no association with heart disease ($V \approx 0.06$) despite
>     moderate associations with the other outcomes.
>
> -   **Takeaway:** Heart disease appears as a **clinically acute
>     endpoint**, most strongly associated with poor general health,
>     diabetes, and functional impairment rather than age-gradient
>     metabolic factors alone. This suggests interaction effects warrant
>     investigation in subsequent cardiovascular modelling (RQ2).

|                          |                         |                       |
|--------------------------|-------------------------|-----------------------|
| **Outcome**              | **Strongest Predictor** | **Second Predictor**  |
| **High Blood Pressure**  | Age (0.35)              | General Health (0.30) |
| **High Cholesterol**     | Age (0.29)              | General Health (0.21) |
| **Heart Disease/Attack** | GenHlth/Diabetes (0.27) | Age (0.23)            |

### 3.1.13 Summary of Key Findings

Section 3.1 establishes that cardiovascular risk in this dataset is
structured along three primary axes: **age, socioeconomic position, and
cardiometabolic clustering**.

First, **age is the dominant organising factor**. High blood pressure
and high cholesterol rise steadily from midlife onward, while heart
disease is far more sharply concentrated in older age groups. Heart
disease appears less as a gradual condition and more as a late-stage
vascular outcome.

Second, a **clear and consistent socioeconomic gradient** is evident
across all three outcomes. Lower education and lower income strata show
substantially higher crude prevalence, with monotonic declines across
both dimensions. Although partly influenced by age composition, the
two-dimensional gradient suggests an independent structural role for
socioeconomic position.

Third, **cardiometabolic conditions cluster rather than occur
independently**. Dual high blood pressure and high cholesterol (25.5%)
exceed either isolated condition, and heart disease prevalence increases
monotonically across BP/cholesterol profiles, peaking in the dual-high
group. This supports a cumulative vascular burden framework.

Fourth, **obesity strongly structures hypertension and cholesterol but
not heart disease**, reinforcing the interpretation of BP and
cholesterol as intermediate metabolic states. Heart disease instead
aligns more strongly with diabetes, general health, and walking
difficulty.

Finally, association analyses (Cramér’s V) reveal two distinct disease
profiles:

-   **Hypertension and cholesterol** are predominantly age and
    general-health driven.

-   **Heart disease or myocardial infarction** is more closely
    associated with established poor general health, diabetes, and
    functional impairment.

Overall, Section 3.1 demonstrates that cardiovascular outcomes in this
dataset are not randomly distributed but follow coherent demographic,
socioeconomic, and pathological structures. These findings provide the
empirical foundation for the modelling strategy developed in Section 4.

### 3.1.14 Implications for Modelling

The exploratory analysis of cardiometabolic outcomes suggests several
considerations for subsequent modelling. First, the strong accumulation
pattern observed across cardiometabolic risk factors indicates that
lifestyle and metabolic variables such as obesity, smoking, physical
inactivity, and diabetes are likely to act as key predictors of
cardiovascular outcomes. Second, the association strength analysis shows
that demographic and health-status variables—particularly age and
self-rated general health—have the strongest relationships with high
blood pressure and high cholesterol, indicating that these variables
should be prioritised in predictive models.

For heart disease or myocardial infarction, the strongest associations
appear with general health status, diabetes, and functional limitation
(difficulty walking), suggesting that heart disease risk may be more
closely linked to underlying health status and comorbidity than to
age-gradient metabolic factors alone. The relatively low prevalence of
heart disease compared with hypertension and hypercholesterolaemia
suggests that models predicting heart disease may face class imbalance
and may require approaches suitable for rare outcomes, or alternatively,
that interaction effects between clinical risk factors and socioeconomic
position (as explored in RQ2) may reveal important moderation patterns.

Overall, the EDA indicates that **demographic factors, general health
status, and cardiometabolic risk profiles** are likely to be central
components of the modelling framework used to investigate cardiovascular
outcomes.

## 3.2 Patterns and determinants of mental health burden

This section examines the distribution and key correlates of **mental
health burden**, measured as the number of mentally unhealthy days
reported in the past 30 days. The analysis explores how mental health
burden varies across demographic, socioeconomic, and health-related
factors, including age, physical health, and lifestyle behaviours. By
identifying systematic patterns in mental health outcomes, this section
aims to highlight the factors most strongly associated with elevated
psychological distress in the population.

### 3.2.1 Distribution of mentally unhealthy days in the past 30 days

```{r}
#| label: mentlhlth-distribution
#| fig-height: 6
#| fig-width: 14
ggplot(diab, aes(x = MentHlth)) +
  geom_histogram(
    binwidth = 1,
    boundary = 0,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = seq(0, 30, by = 5),
    limits = c(0, 30)
  ) +
  scale_y_continuous(
    labels = label_number(scale = 1e-3, suffix = "K"),
    breaks = seq(0,200000,by = 25000)
  ) +
  labs(
    title = "Distribution of mentally unhealthy days in the past 30 days",
    x = "Number of mentally unhealthy days",
    y = "Frequency") 
```

> The distribution of mentally unhealthy days is **highly
> right-skewed**, with a very large proportion of respondents reporting
> **zero mentally unhealthy days in the past month**. Frequencies
> decline sharply as the number of unhealthy days increases, producing a
> long right tail that extends toward the upper bound of 30 days. This
> pattern indicates that while most individuals report little or no
> recent mental health burden, a smaller subset experiences **frequent
> or persistent distress**.
>
> The distribution also shows evidence of **excess zeros and strong
> dispersion**, with the spike at zero far exceeding the frequencies of
> neighbouring values. Such a structure suggests that standard linear
> modelling would be inappropriate, and motivates the use of
> **count-based models**, potentially including **zero-inflated or
> overdispersed models (e.g., Poisson or zero-inflated negative
> binomial)** when analysing mental health burden as an outcome in
> subsequent modelling.

### 3.2.3 Mental Health Burden by Income and Sex

```{r}
#| label: mental-health-income
#| fig.height: 7
#| fig.width: 14
diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(Income, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  
  ggplot(aes(x = Income, y = prop_high, fill = Sex)) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.7,
    colour = "black",
    linewidth = 0.2
  ) +
  geom_text(
    aes(label = percent(prop_high, accuracy = 1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  )
```

> A clear socioeconomic gradient is evident in the prevalence of high
> mental health burden. The proportion of respondents reporting ≥14
> mentally unhealthy days declines steadily as household income
> increases, indicating that individuals in lower-income groups
> experience substantially higher levels of psychological distress. In
> the lowest income category (\<\$10k), roughly 28% of females and 26%
> of males report high mental health burden, compared with only 6% and
> 4% respectively in the highest income group (\$75k+).
>
> Across all income categories, females consistently report higher
> prevalence than males, although the difference between sexes is modest
> relative to the overall income gradient. These results suggest that
> socioeconomic position is a strong determinant of mental health
> burden, with financial disadvantage associated with markedly higher
> levels of frequent psychological distress. (Money doesn't buy
> happiness :) )

### 3.2.3 Mental Health Burden by Educatioand Sex

```{r}
#| label: mental-health-education
#| fig.height: 7
#| fig.width: 14
diab |>
  mutate(HighMH = MentHlth >= 14) |>
  group_by(Education, Sex) |>
  summarise(prop_high = mean(HighMH), .groups = "drop") |>
  
  ggplot(aes(x = Education, y = prop_high, fill = Sex)) +
  geom_col(
    position = position_dodge(width = 0.75),
    width = 0.7,
    colour = "black",
    linewidth = 0.2
  ) +
  geom_text(
    aes(label = percent(prop_high, accuracy = 1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  scale_fill_manual(
    values = pal_sex
  ) +
  labs(
    title = "Prevalence of high mental health burden (≥14 days) by income and sex",
    x     = "Household income",
    y     = "Proportion with ≥14 mentally unhealthy days",
    fill  = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

> A clear educational gradient is visible in the prevalence of high
> mental health burden. Individuals with lower levels of educational
> attainment report substantially higher proportions of ≥14 mentally
> unhealthy days compared with those with higher education. For example,
> among respondents with kindergarten or less education, approximately
> 22% report high mental health burden, whereas this proportion declines
> to around 8% for females and 5% for males among college graduates.
>
> Across all education categories, females consistently report higher
> levels of mental health burden than males, although the difference
> between sexes is relatively modest compared with the overall education
> gradient. The steady decline in prevalence with increasing educational
> attainment suggests that socioeconomic resources and opportunities
> associated with education may play an important role in mental health
> outcomes.
>
> Overall, the figure indicates that lower educational attainment is
> associated with substantially higher levels of frequent psychological
> distress, reinforcing the broader pattern of socioeconomic
> inequalities in mental health observed in the income analysis.

### 3.2.4 Mental Health Burden by Age Group and Sex

```{r}
#| label: mental-health-age-lollipop
#| fig.height: 7
#| fig.width: 14


mh_age_sex <- diab |>
  group_by(Age, Sex) |>
  summarise(mean_mh = mean(MentHlth, na.rm = TRUE), .groups = "drop")

pos_dodge <- position_dodge(width = 0.6)

ggplot(mh_age_sex, aes(x = Age, y = mean_mh, colour = Sex)) +
  geom_linerange(
    aes(ymin = 0, ymax = mean_mh),
    position = pos_dodge,
    linewidth = 1
  ) +
  geom_point(
    position = pos_dodge,
    size = 3.8
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_manual(
    values = pal_sex
  ) +
  labs(
    title  = "Mean mentally unhealthy days by age group and sex",
    x      = "Age group",
    y      = "Mean mentally unhealthy days (past 30 days)",
    colour = "Sex"
  ) 
```

> Mean mentally unhealthy days are higher among women than men in every
> age group. The difference is most visible in younger and middle
> adulthood, where women report roughly 1–2 more days on average than
> men.
>
> Across age, the pattern is not monotonic. Mean days are relatively
> high in early adulthood, remain similar through midlife, and then
> decline progressively after around age 60 for both sexes. The lowest
> averages are observed in the oldest age groups.

### 3.2.5 Joint Distribution of Mental and Physical Health Burden by Sex

```{r}
#| label: menth-physh-hexbin
#| fig.height: 7
#| fig.width: 14
ggplot(diab, aes(x = MentHlth, y = PhysHlth)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c(option = "magma", name = "Count", trans = "log10") +
  facet_wrap(~ Sex, nrow = 1) +
  labs(
    title = "Joint distribution of mentally and physically unhealthy days by sex",
    x     = "Mentally unhealthy days (past 30)",
    y     = "Physically unhealthy days (past 30)"
  )
```

> In both panels, the highest concentration of observations occurs near
> zero mentally and physically unhealthy days. Density decreases as
> values increase along either axis, and combinations of simultaneously
> high mental and physical burden are relatively sparse.
>
> There is visible clustering along diagonal regions, indicating that
> higher numbers of mentally unhealthy days tend to coincide with higher
> numbers of physically unhealthy days rather than being concentrated
> along a single axis.
>
> The overall distributional shape is similar for women and men, with
> broadly comparable density patterns across low, moderate, and higher
> values.

### 3.2.6 Prevalence of High Mental Health Burden by Physical and Cardiometabolic indicators

```{r}
#| include: false
diab_mh_phys <- diab |>
  mutate(
    HighMH = factor(if_else(MentHlth >= 14, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    Obese = factor(if_else(BMI >= 30, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    BadPhysHlth = factor(if_else(PhysHlth >= 14, "Yes", "No"),
                    levels = c("No", "Yes")),
    
    Diabetes = factor(if_else(Diabetes == "Diabetes", "Yes", "No"),
                        levels = c("No", "Yes")),
    
    HighBP = factor(if_else(HighBP == "High", "Yes", "No"),
                        levels = c("No", "Yes")),
    
    HighChol = factor(if_else(HighChol == "High", "Yes", "No"),
                        levels = c("No", "Yes"))
  ) |>
  select(
    HighMH,
    Obese,
    PhysActivity,
    DiffWalk,
    BadPhysHlth,
    Diabetes,
    HeartDiseaseorAttack,
    HighBP,
    HighChol
  ) |>
  pivot_longer(
    cols = -HighMH,
    names_to = "Indicator",
    values_to = "Status"
  ) |> 
  count(Indicator, Status, HighMH) |>
  group_by(Indicator, Status) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  filter(HighMH == "Yes")
```

```{r}
#| fig.height: 7
#| fig.width: 14

diab_mh_phys |>
  ggplot(aes(x = Status, y = prop, fill = Status)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = percent(prop, accuracy = 1)),
    vjust = -0.3,
    size  = 3.3
  ) +
  facet_wrap(~ Indicator, nrow = 2, ncol = 4) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.10))
  ) +
  scale_fill_manual(values = pal_binary) +
  labs(
    title = "Prevalence of high Mental Health Burden by Physical and Cardiometabolic Indicators",
    x     = "Indicator status",
    y     = "Proportion with ≥14 mentally unhealthy days"
  )
```

> High mental health burden (≥14 mentally unhealthy days) varies
> substantially across physical and cardiometabolic indicators. The
> strongest associations are observed for indicators of functional and
> physical health limitations. Individuals reporting high physical
> health burden (≥14 physically unhealthy days) show a markedly elevated
> prevalence of high mental health burden ( ≈ 33%), compared with only
> about 6% among those without high physical burden. A similarly large
> difference appears for difficulty walking, where prevalence rises from
> roughly 7% to 24%.
>
> Cardiometabolic conditions such as diabetes, heart disease or
> myocardial infarction, high blood pressure, and high cholesterol are
> also associated with higher levels of mental health burden, although
> the magnitude of these differences is smaller. For example,
> respondents with diabetes or heart disease exhibit prevalence around
> 15%, compared with approximately 9% among those without these
> conditions.
>
> Lifestyle factors show a mixed pattern. Obesity is associated with
> moderately higher mental health burden (≈12% vs ≈8%), while physical
> activity appears protective, with prevalence roughly halved among
> individuals reporting recent physical activity ( ≈ 8%) compared with
> those reporting none ( ≈16%).
>
> Overall, the results suggest that mental health burden is most
> strongly linked with physical health limitations and functional
> impairment, while cardiometabolic disease and lifestyle factors
> exhibit weaker but still noticeable associations.

### 3.2.7 Summary of Key Findings

Several clear structural patterns emerge in the analysis of mental
health burden. First, the distribution of mentally unhealthy days is
highly right-skewed, with a large concentration at zero and a long tail
toward higher values. This indicates substantial heterogeneity in
psychological distress across the population and motivates the use of
count-based models with potential zero inflation in subsequent
modelling.

Second, strong **socioeconomic gradients** are visible. The prevalence
of high mental health burden declines consistently with increasing
income and education, suggesting that socioeconomic disadvantage is
associated with substantially higher levels of frequent psychological
distress.

Third, **demographic differences** are evident. Women report higher
average numbers of mentally unhealthy days than men across all age
groups, while mental health burden tends to decline gradually in older
age groups after midlife.

Fourth, mental and physical health burdens show **moderate
co-occurrence**, with higher levels of mentally unhealthy days tending
to coincide with higher levels of physically unhealthy days.

Finally, mental health burden is **most strongly associated with
indicators of physical limitation**, particularly poor physical health
and difficulty walking. Cardiometabolic conditions such as diabetes,
heart disease, hypertension, and high cholesterol also show positive
associations with mental health burden, although these relationships are
weaker than those linked to functional impairment.

Overall, the results suggest that mental health burden in this
population is structured primarily by **socioeconomic position, physical
health status, and functional limitation**, providing a clear empirical
basis for the modelling analyses developed in later sections.

### 3.2.8 Implications for Modelling

The exploratory analysis provides several insights relevant for model
development. The distribution of mentally unhealthy days is highly
right-skewed with a substantial concentration at zero, suggesting that
standard linear modelling would be inappropriate and motivating the use
of count-based models such as Poisson or negative binomial regression,
potentially with zero-inflation.

Socioeconomic gradients observed across income and education indicate
that demographic and socioeconomic variables should be included as key
predictors.

Finally, the strong co-occurrence between mental and physical health
indicators suggests that physical health variables such as poor physical
health days, difficulty walking, and chronic disease indicators may play
an important explanatory role in modelling mental health burden.

## 3.3 Predictor Association Structure and Multicollinearity Assessment

Before proceeding to multivariable modelling, it is essential to examine
the correlation structure among potential predictors. Strong pairwise
associations may indicate redundancy or multicollinearity, which can
inflate standard errors and complicate interpretation. This section
quantifies associations among continuous, ordinal, and binary variables
using Spearman's rank correlation and Cramér's V, respectively. The
analysis establishes which predictors are inter-related and informs
variable selection and multicollinearity diagnostics in subsequent
models.

### 3.3.1 Spearman Correlation Heatmap: Continuous and Ordinal Variables

```{r}
#| label: spearman-prep
#| include: false
corr_vars <- diab |>
  transmute(
    BMI,
    MentHlth,
    PhysHlth,
    GenHlth   = as.integer(GenHlth),
    Age       = as.integer(Age),
    Education = as.integer(Education),
    Income    = as.integer(Income)
  )

spear_mat <- cor(
  corr_vars,
  method = "spearman",
  use    = "pairwise.complete.obs"
)

spear_tidy <- spear_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(
    cols      = -Var1,
    names_to  = "Var2",
    values_to = "rho"
  ) |>
  mutate(
    Var1 = factor(Var1, levels = colnames(spear_mat)),
    Var2 = factor(Var2, levels = colnames(spear_mat))
  )
```

```{r}
#| label: spearman-heatmap
#| fig.height: 7
#| fig.width: 10

ggplot(spear_tidy, aes(x = Var1, y = Var2, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", rho)),
    size   = 3.5,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#D6604D",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman's \u03c1"
  ) +
  labs(
    title = "Spearman rank correlations among continuous and ordinal variables",
    x     = NULL,
    y     = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1),
    axis.text.y = element_text(hjust = 1)
  )
```

> The Spearman rank correlations indicate mostly **weak to moderate**
> monotonic associations among the continuous and ordinal variables.
>
> -   The strongest positive correlations are between **GenHlth and
>     PhysHlth** (ρ ≈ 0.45) and between **Education and Income** (ρ ≈
>     0.45), suggesting moderate monotonic relationships in both cases.
>
> -   **GenHlth** is moderately **negatively** correlated with
>     **Income** (ρ ≈ −0.35) and **Education** (ρ ≈ −0.28). Because
>     higher GenHlth scores denote poorer health, these negative
>     correlations imply that higher income and education are associated
>     with **better** self-rated health.
>
> -   **MentHlth** and **PhysHlth** show a modest positive association
>     (ρ ≈ 0.31), consistent with some co-occurrence of mental and
>     physical health burden.
>
> -   Most other pairwise correlations are small (\|ρ\| \< 0.3). In
>     particular, **BMI** and **Age** have only weak associations with
>     the other variables in this set.
>
> Overall, the correlation matrix does **not** indicate problematic
> collinearity: apart from a few moderate associations
> (GenHlth–PhysHlth, Education–Income), most variables are only weakly
> related. This supports their concurrent use as predictors in
> multivariable models, subject to routine multicollinearity checks.

### 3.3.2 Cramér's V Association Heatmap: Binary Clinical and Lifestyle Variables

To complement the Spearman correlation heatmap for continuous/ordinal
variables, the strength of pairwise associations among all binary
clinical and lifestyle indicators is quantified using Cramér's V. This
symmetric measure summarises the association in each 2×2 contingency
table on a 0–1 scale, without assuming any directionality. The analysis
establishes whether binary predictors cluster or operate independently,
informing whether multicollinearity diagnostics will be necessary in
subsequent regression models. Diabetes status is included here as a
predictor to assess its associations with other binary indicators.

```{r}
#| label: cramer-prep
#| include: false
bin_vars <- c("HighBP", "HighChol", "Smoker", "Stroke",
              "HeartDiseaseorAttack", "PhysActivity",
              "HvyAlcoholConsump", "DiffWalk",
              "AnyHealthcare", "NoDocbcCost","Diabetes")

cv_mat <- outer(
  bin_vars, bin_vars,
  Vectorize(function(a, b) {
    cramerV(table(diab[[a]], diab[[b]]))
  })
)
dimnames(cv_mat) <- list(bin_vars, bin_vars)

cv_tidy <- cv_mat |>
  as.data.frame() |>
  rownames_to_column("Var1") |>
  pivot_longer(-Var1, names_to = "Var2", values_to = "V") |>
  mutate(
    Var1 = factor(Var1, levels = bin_vars),
    Var2 = factor(Var2, levels = bin_vars)
  )
```

```{r}
#| label: cramerv-heatmap
#| fig.height: 7
#| fig.width: 10

ggplot(cv_tidy, aes(x = Var1, y = Var2, fill = V)) +
  geom_tile(colour = "white") +
  geom_text(
    aes(label = sprintf("%.2f", V)),
    size   = 3.8,
    colour = "black"
  ) +
  scale_fill_gradient2(
    low       = "white",
    high     = "#D6604D",
    midpoint = 0,
    limits   = c(0, 1),
    name     = "Cramér's V"
  ) +
  labs(
    title = "Cramér's V among binary clinical and lifestyle variables",
    x = NULL,
    y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

> The Cramér’s V matrix indicates that most pairwise associations among
> the binary clinical and lifestyle variables are **weak**, with the
> majority of values below 0.20.
>
> The strongest association is between **HighBP and HighChol** (V ≈
> 0.30), representing a moderate relationship relative to the rest of
> the matrix. This is consistent with known clustering of cardiovascular
> risk factors. **Diabetes** shows its strongest associations with
> **HighBP** (V ≈ 0.27) and **DiffWalk** (V ≈ 0.22), followed by
> **HighChol** (V ≈ 0.21), indicating that diabetes tends to co-occur
> with cardiovascular risk factors and functional limitation.
>
> Other comparatively larger associations include:
>
> -   **PhysActivity and DiffWalk** (V ≈ 0.25)
>
> -   **HighBP and DiffWalk** (V ≈ 0.22)
>
> -   **HighBP and HeartDiseaseorAttack** (V ≈ 0.21)
>
> -   **Stroke and HeartDiseaseorAttack** (V ≈ 0.20)
>
> These moderate associations are clinically plausible and consistent
> with shared cardiovascular or functional impairment pathways.
>
> Lifestyle indicators such as **Smoker** and **HvyAlcoholConsump**
> exhibit generally weak associations with other variables (typically V
> \< 0.15). Healthcare access indicators (**AnyHealthcare**,
> **NoDocbcCost**) also show minimal association with most clinical
> variables.
>
> **Analytical Implication**
>
> The matrix does not suggest substantial redundancy among the binary
> predictors. Aside from a small number of moderate associations
> (primarily within cardiovascular-related variables and diabetes), most
> relationships are weak.

# 4. Refined Research Questions

## RQ1: Multidomain Determinants and Prediction of Diabetes Status

**Outcome Variable:** *Diabetes — Multiclass Outcome*

**Research Question**

> ***To what extent can diabetes status be explained and predicted using
> demographic, socioeconomic, behavioural, and cardiometabolic factors,
> and which predictors contribute most strongly to distinguishing
> between no diabetes, prediabetes, and diabetes?***

**Rationale**

Exploratory analysis of the CDC Diabetes Health Indicators dataset
indicates that diabetes status is structured by multiple interacting
determinants. Section 3.2.1 demonstrated a strong **age gradient**, with
prediabetes and diabetes concentrated in older age groups, suggesting
that demographic factors must be accounted for when evaluating other
risk indicators.

Cardiometabolic clustering was evident across several visualisations.
The joint distribution of **high blood pressure and high cholesterol**
increased substantially across diabetes categories, while the
risk-factor accumulation plot showed a clear gradient in which diabetes
prevalence rises with the number of cardiometabolic risk indicators.
This pattern aligns with established evidence linking obesity,
hypertension, and dyslipidaemia to type 2 diabetes risk (Ismail et al.,
2021).

Behavioural factors also exhibited variation across diabetes categories.
Physical activity levels tended to decline with worsening diabetes
status, while smoking and alcohol consumption showed weaker separation.
Because behavioural variables are potentially modifiable, evaluating
their contribution to diabetes risk is relevant for prevention
strategies (Aune et al., 2015).

Finally, socioeconomic gradients were visible in the education–income
distribution, where diabetes prevalence was higher among lower
socioeconomic groups. These patterns suggest that diabetes risk is
embedded within broader social determinants of health, which influence
exposure to risk factors and access to care (Hill-Briggs et al., 2021).

Together, these findings motivate a multivariable modelling approach to
estimate adjusted associations and evaluate predictive performance
across diabetes categories.

*'EDI Discussed in Task1'*

## RQ2: The Impact of Socioeconomic Moderation on Cardiovascular Risk

**Outcome Variables:** *HeartDiseaseorAttack* — binary outcome

**Research Question**

> ***To what extent do socioeconomic determinants, specifically income
> and education, moderate the impact of clinical risk factors on the
> prevalence of heart disease, and can the capture of complex feature
> interactions provide superior predictive insights compared to
> traditional additive assumptions?***

**Rationale**

Exploratory analysis identified distinct gradients in cardiovascular
health, where the prevalence of heart disease increases substantially
across older age groups and among individuals with specific clinical
markers. While factors such as high blood pressure, elevated
cholesterol, and body mass index are established universal risk
indicators, their influence is rarely isolated. In real-world clinical
contexts, these physiological markers exist within a broader
socioeconomic framework, suggesting that the "weight" of a clinical risk
factor may be compounded or mitigated by an individual's access to
resources, environmental stressors, and educational background (Roth et
al., 2024).

Traditional analytical frameworks often rely on the assumption that
risks are simply additive, yet biological and social systems frequently
exhibit synergistic effects where the presence of one factor
significantly alters the impact of another. For example, the
cardiovascular strain associated with high BMI may be intensified in
lower-income brackets due to limited access to preventive care or higher
cumulative stress (Collins et al., 2023). By investigating these
conditional relationships, this inquiry shifts the focus from simple
binary classification toward a more nuanced understanding of risk
stratification. This approach allows for the identification of specific
"risk profiles," which are intersections of clinical and social data
that might be overlooked by frameworks that fail to account for
higher-order interactions.

![The Dahlgren-Whitehead Rainbow Model illustrating the layers of health
determinants, providing the theoretical basis for the moderation
analysis of income and education in
RQ2.](https://raw.githubusercontent.com/SHAH-MEER/Modellin-Data/main/image/Rainbow-model-of-the-determinants-of-health.png){#fig-sdoh
fig-align="center" width="85%"}

As visualized in figure, individual clinical risks like blood pressure
and BMI are nested within broader socioeconomic environments. This
framework justifies the shift from simple predictive modelling to an
investigation of how these external layers moderate physiological risk.

**EDI and Ethical Considerations**

The application of predictive modelling in healthcare necessitates a
rigorous focus on transparency and equity. A primary ethical concern is
the risk of "algorithmic bias," where models may inadvertently penalise
or misrepresent specific subgroups if demographic variables are treated
in isolation. By specifically investigating socioeconomic moderators,
this research aims to move beyond a "one-size-fits-all" predictive
approach, ensuring that the influence of income and education is
explicitly understood rather than buried as "noise" within the data.

Furthermore, the interpretability of these complex interactions is a
matter of distributive justice, as clinicians must be able to explain
risk assessments to patients in a way that is both accurate and
contextually relevant. By quantifying how socioeconomic status shifts
the predictive significance of clinical markers, this study helps
identify whether existing healthcare inequalities are being reinforced
or challenged by the data. Ensuring that the model identifies these
non-linear patterns is essential for developing more equitable
preventative strategies that reflect the lived realities of diverse
population groups (Liu et al., 2023).

## RQ3: Socioeconomic and Lifestyle Determinants of Mental Health Burden

**Outcome Variable:** *MentHlth* — number of self-reported mentally
unhealthy days in the past 30 days (0–30)

**Research Question**

> *To what extent do socioeconomic position (income, education),
> lifestyle behaviours (physical activity, smoking), and physical health
> burden predict the number of mentally unhealthy days reported in the
> past 30 days, and do the protective associations of physical activity
> differ across income and education strata?*

**Rationale**

The *MentHlth* variable represents a count of mentally unhealthy days
within the past month (0–30). Exploratory analysis revealed a strongly
right-skewed distribution with a large concentration of zero values,
indicating that many respondents report no mentally unhealthy days while
a smaller subset experience substantial burden. Such distributions
violate assumptions of linear models and often produce overdispersion
and excess zeros, making standard Poisson or logistic approaches
unsuitable.

A Zero-Inflated Negative Binomial (ZINB) model is therefore appropriate
because it separates two processes:

1.  the probability that individuals experience no mental health burden,
    and

2.  the frequency of mentally unhealthy days among those who do report
    burden (Rust et al., 2014; Liu et al., 2023).

The EDA further revealed clear socioeconomic gradients in mental health
outcomes. Respondents in the lowest income categories exhibited
substantially higher prevalence of high mental health burden compared
with those in higher income groups. Physical activity also showed a
consistent inverse relationship with mentally unhealthy days, suggesting
potential behavioural protection.

These patterns motivate a modelling approach that jointly evaluates
socioeconomic position and lifestyle behaviours. In particular, testing
an interaction between income and physical activity allows examination
of whether the protective association of physical activity differs
across socioeconomic strata.

Prior research supports the relevance of these predictors. Socioeconomic
disadvantage has been consistently linked with elevated risk of mental
health disorders and poorer psychological well-being (Kessler et al.,
2010), while analyses of BRFSS data have shown that ZINB regression
provides an effective framework for modelling unhealthy-days indicators
with excess zeros (Rust et al., 2014; Liu et al., 2023).

**EDI / Ethical Considerations**

Socioeconomic gradients in mental health outcomes raise important equity
considerations. If income remains strongly associated with mental health
burden after adjustment for behavioural factors, this suggests that
structural determinants, such as economic insecurity and access to
resources, play a major role in shaping mental health outcomes.
Understanding these relationships can inform more equitable public
health strategies that address structural conditions rather than
focusing solely on individual behaviour.

# 6. Multidomain Determinants and Prediction of Diabetes Status

## 6.1 Modelling Objective

This section addresses the following research question:

> ***To what extent can diabetes status be explained and predicted using
> demographic, socioeconomic, behavioural, and cardiometabolic factors,
> and which predictors contribute most strongly to distinguishing
> between no diabetes, prediabetes, and diabetes?***

The objective is twofold:

1.  **Explanatory** – to estimate the adjusted associations between
    predictor domains and diabetes status.

2.  **Predictive** – to compare the classification performance of
    statistical and machine learning models.

Diabetes status is treated as a **multiclass outcome** (No diabetes /
Prediabetes / Diabetes), requiring modelling approaches suitable for
nominal categorical outcomes.

## 6.2. Outcome Variable

The dependent variable is **diabetes status**, categorised into:

-   **No diabetes**

-   **Prediabetes**

-   **Diabetes.**

Exploratory analysis demonstrated class imbalance, with the majority of
respondents in the no-diabetes category and substantially smaller
proportions in the prediabetes and diabetes groups. This imbalance is
accounted for in model training and evaluation.

## 6.3 Training and Validation

### 6.3.1 **Data Partitioning and Resampling:**

The dataset is partitioned into a 75% training set and a 25% testing set
using stratified sampling on diabetes status. Within the training set,
5-fold stratified cross-validation is implemented to estimate model
performance during development. The held-out test set is reserved for
final evaluation to ensure unbiased performance assessment.

```{r}
set.seed(42)

diabetes_split <- initial_split(
  diab,
  prop = 0.75,
  strata = Diabetes
)

diab_train <- training(diabetes_split)
diab_test  <- testing(diabetes_split)


diabetes_folds <- vfold_cv(
  diab_train,
  v = 5,
  strata = Diabetes
)

# Sanity checks
diab_train |>  count(Diabetes) |> mutate(prop = n/sum(n))
diab_test |> count(Diabetes) |>  mutate(prop = n/sum(n))

```

### 6.3.2 Recipe Definition and Preprocessing Strategy

-   **Outcome specification** – Models `Diabetes` as a three-level
    categorical outcome (No diabetes, Prediabetes, Diabetes).

-   **Near-zero variance removal** – Removes predictors with minimal
    variability to prevent instability and redundant information.

-   **Ordinal scoring** – Converts ordered factors (`Age`, `Education`,
    `Income`, `GenHlth`) into numeric scores while preserving their
    natural ranking. This introduces an assumption of approximately
    equal spacing between adjacent categories; however, the primary
    objective is to retain monotonic structure in a computationally
    efficient representation suitable for both parametric and tree-based
    models.

-   **Dummy encoding** – Transforms remaining nominal predictors into
    binary indicator variables suitable for modelling.

-   **Normalization** – Standardizes numeric predictors (mean = 0, sd
    = 1) to improve optimisation and comparability across models.

-   **SMOTE balancing** – Applies synthetic oversampling to minority
    classes within resampling folds to mitigate severe class imbalance.

```{r}
#| label: recipie-making
set.seed(42)

rec_rq1 <- recipe(Diabetes ~ ., data = diab_train) |>
  step_nzv(all_predictors()) |>
  step_ordinalscore(Age, Education, Income, GenHlth) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(BMI, MentHlth, PhysHlth) |>
  step_smote(Diabetes, over_ratio = 0.5, seed = 42)

print(rec_rq1)
```

### 6.3.3 Model Specification

We now define the candidate models for RQ1 using the **tidymodels**
framework.

Each model is specified independently of preprocessing and will later be
combined with the common recipe using **workflows** and **workflow
sets**.

|  |  |  |  |
|------------------|------------------|------------------|------------------|
| **Model** | **Type** | **Primary Function** | **Key Strengths** |
| **Multinomial Logistic Regression** | Parametric | Linear Baseline | Interpretable log-odds; direct coefficient analysis. |
| **Random Forest** | Ensemble (Bagging) | Non-linear Discovery | Captures complex interactions; robust to multicollinearity. |
| **XGBoost** | Ensemble (Boosting) | Predictive Performance | High efficiency on tabular data; optimized for accuracy. |

Hyperparameters marked `tune()` will be optimized via grid search during
cross-validation.

```{r}
#| label: model-spec

# Multinomial logistic regression
mod_multinom <- 
  multinom_reg() |> 
  set_engine("nnet") |> 
  set_mode("classification")

# Random Forest
mod_rf <- 
  rand_forest(
    trees = 256,
    mtry  = tune(),
    min_n = tune()
  ) |> 
  set_engine("ranger") |> 
  set_mode("classification")

# Gradient Boosted Trees
mod_xgb <- 
  boost_tree(
    trees = 512,
    tree_depth = tune(),
    learn_rate = tune(),
    loss_reduction = tune(),
    mtry = tune(),
    min_n = tune()
  ) |> 
  set_engine("xgboost") |> 
  set_mode("classification")
```

### 6.3.4 Workflowset Construction

We now integrate:

-   The **RQ1 preprocessing recipe**

-   The **candidate model specifications**

Each workflow represents one **model–preprocessing pair**.\
The workflow set enables parallel evaluation of all candidate models
under identical resampling conditions.

```{r}
#| label: wrkflw-sets

# Workflow set 
wf_set_rq1 <-
  workflow_set(
    preproc = list(rq1_recipe = rec_rq1),
    models  = list(
      multinom = mod_multinom,
      rf       = mod_rf,
      xgb      = mod_xgb))

```

### 6.3.5 Hyperparameter Tuning

**Metrics computed**

Model performance is evaluated using multiple complementary metrics to
account for class imbalance and provide a robust assessment of
predictive performance:

-   **Accuracy** – overall classification rate

-   **Balanced Accuracy** – mean recall across classes, correcting for
    class imbalance

-   **Cohen’s Kappa** – agreement between predictions and observed
    classes beyond chance

-   **F1-score (macro)** – harmonic mean of precision and recall
    averaged across classes

-   **ROC-AUC (multiclass)** – ability of the model to discriminate
    between diabetes categories

```{r}
#| label: metric-set
metrics_diab <-
  metric_set(
    accuracy,
    bal_accuracy,
    kap,
    f_meas,
    roc_auc
  )
```

**Hyperparameter Tuning Strategy**

We employ **grid search** to optimize hyperparameters across candidate
models:

-   **Grid size:** 10 configurations per model, balancing exploration of
    the hyperparameter space with computational feasibility (3 models ×
    10 grids × 5 CV folds = 150 model fits)
-   **Search method:** Grid search over random/Bayesian approaches
    provides systematic coverage of parameter combinations and is
    computationally manageable for this dataset size
-   **Tuning metric:** **Macro-averaged F1 score** serves as the primary
    optimization metric because it gives equal weight to all three
    diabetes classes (No diabetes, Prediabetes, Diabetes), preventing
    models from optimizing solely for majority class performance
-   **Cross-validation:** 5-fold stratified CV estimates out-of-sample
    performance while maintaining class proportions in each fold

The best hyperparameter configuration for each model is selected based
on highest mean Macro F1 across folds.

```{r}
#| label: hp-tune
# set.seed(42)
# ctrl <- control_grid(
#   save_pred = TRUE,
#   save_workflow = FALSE,
#   verbose = TRUE
# )
# 
# wf_set_rq1_fit <-
#   wf_set_rq1 |>
#   workflow_map(
#     "tune_grid",
#     resamples = diabetes_folds,
#     grid = 10,
#     metrics = metrics_diab,
#     control = ctrl
#     )

wf_set_rq1_fit <- readRDS('wf_set_rq1_fit.rds')
```

```{r}
#| include: false

pretty_autoplot <- function(x, metric, ylab, title) {
  autoplot(
    x, 
    metric = metric) +
    labs(
      title = title,
      x = "Workflow rank (best → worst)",
      y = ylab
    ) +
    scale_y_continuous(labels = percent_format(accuracy = 0.1))
}
```

```{r}
#| label: eval-results
#| fig-height: 10
#| fig-width: 16

m1 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "roc_auc", 
  "ROC–AUC (multiclass)", 
  "Discrimination (ROC–AUC)")
m2 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "bal_accuracy", 
  "Balanced accuracy", 
  "Class-balanced accuracy")
m3 <- pretty_autoplot(
  wf_set_rq1_fit, 
  "f_meas", 
  "Macro F1-score", 
  "Macro F1 (imbalance-aware)")
m4 <- pretty_autoplot(wf_set_rq1_fit, "kap", "Cohen's kappa", "Agreement beyond chance")

(m1 | m2) / (m3 | m4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**ROC–AUC.**\
ROC–AUC measures a model’s ability to discriminate between the three
diabetes categories across all possible classification thresholds. The
highest ROC–AUC value is achieved by the **multinomial logistic
regression model**, which appears as the top-ranked workflow in this
metric. However, beyond this leading configuration, most of the
remaining high-ranking workflows consist of **XGBoost and random forest
models**, with values clustered between approximately 0.69 and 0.70.
This indicates that while logistic regression achieves the single best
ROC–AUC score, tree-based models provide similarly competitive
discrimination across multiple configurations.

**Balanced Accuracy.**\
Balanced accuracy reflects the average recall across classes, giving
equal importance to each diabetes category. The **multinomial logistic
regression model again appears as the top-ranked workflow**, achieving
the highest balanced accuracy among the candidate models. Several
**XGBoost configurations follow closely**, while random forest models
occupy the middle range of the ranking. Although performance differences
are relatively modest, the results suggest that multiple model types are
capable of achieving comparable class-balanced recall.

**Macro F1 Score.**\
Macro F1 was selected as the primary tuning metric because it balances
precision and recall equally across classes and is therefore appropriate
under class imbalance. In this panel, the **top-ranked workflows are
clearly dominated by XGBoost models**, which achieve substantially
higher macro F1 scores than the other model types. Random forest
configurations perform considerably lower, and the multinomial logistic
regression model appears further down the ranking. This pattern
indicates that boosted trees are better able to balance precision and
recall across the three diabetes categories.

**Cohen’s Kappa.**\
Cohen’s kappa measures agreement between predicted and observed
classifications beyond chance. The highest kappa values are achieved by
**XGBoost configurations**, which occupy several of the top-ranked
positions in this panel. The multinomial logistic regression model
appears among the higher-performing workflows but does not exceed the
best boosted-tree configurations. Random forest models generally show
lower kappa values and appear further down the ranking. Overall, this
metric again indicates stronger agreement for boosted tree models
compared with the other modelling approaches.

### 6.3.6 Extract tuning results

```{r}
#| label: extarct
# Extract tuning results
rf_res  <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_res <- extract_workflow_set_result(wf_set_rq1_fit, "rq1_recipe_xgb")


best_rf  <- select_best(rf_res, metric = "f_meas")
best_xgb <- select_best(xgb_res, metric = "f_meas")


rf_wf       <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_rf")
xgb_wf      <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_xgb")
multinom_wf <- extract_workflow(wf_set_rq1_fit, "rq1_recipe_multinom")


rf_final_wf  <- finalize_workflow(rf_wf, best_rf)
xgb_final_wf <- finalize_workflow(xgb_wf, best_xgb)

# Multinomial requires no tuning, so it is already final
multinom_final_wf <- multinom_wf
```

```{r}
#| label: final-fit
rf_final_fit <- rf_final_wf |> last_fit(diabetes_split)

xgb_final_fit <- xgb_final_wf |> last_fit(diabetes_split)

multinom_final_fit <- multinom_final_wf |> last_fit(diabetes_split)
```

## 6.4 Model Evaluation

```{r}
pred_rf <- rf_final_fit |>
  collect_predictions() |>
  mutate(model = "Random forest")

pred_xgb <- xgb_final_fit |>
  collect_predictions() |>
  mutate(model = "XGBoost")

pred_mn <- multinom_final_fit |>
  collect_predictions() |>
  mutate(model = "Multinomial")

pred_all <- bind_rows(pred_rf, pred_xgb, pred_mn) |>
  rename(.pred_No_diabetes = `.pred_No diabetes`)
```

```{r}
#| label: rq1-test-metrics
#| message: false
#| warning: false

metrics_rf <- rf_final_fit |>
  collect_metrics() |>
  mutate(model = "Random forest")

metrics_xgb <- xgb_final_fit |>
  collect_metrics() |>
  mutate(model = "XGBoost")

metrics_mn <- multinom_final_fit |>
  collect_metrics() |>
  mutate(model = "Multinomial")

metrics_all <- bind_rows(metrics_rf, metrics_xgb, metrics_mn)

metrics_wide <- metrics_all |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate)
```

### 6.4.1 ROC Curves

**Multiclass ROC-AUC Evaluation**

ROC (Receiver Operating Characteristic) curves for multiclass outcomes
require extending the binary framework. We employ the **one-vs-rest
approach**:

-   Each class (No diabetes, Prediabetes, Diabetes) is treated as a
    separate binary problem against all other classes combined
-   For each class, ROC curves plot True Positive Rate vs False Positive
    Rate across probability thresholds
-   **Area Under Curve (AUC)** summarizes discriminative ability: AUC =
    0.5 indicates random guessing, AUC = 1.0 indicates perfect
    separation
-   Despite class imbalance, ROC-AUC remains informative for comparing
    relative model performance, though it may be optimistic for rare
    classes (Prediabetes, Diabetes)

Higher AUC values indicate better ability to distinguish each diabetes
category from alternatives. Models with consistently high AUC across all
three classes demonstrate robust multiclass discrimination.

```{r}
#| label: rq1-roc-combined
#| message: false
#| warning: false

roc_all <- pred_all |>
  group_by(model) |>
  roc_curve(
    truth = Diabetes,
    .pred_No_diabetes,
    .pred_Prediabetes,
    .pred_Diabetes)

autoplot(roc_all) +
  labs(
    title = "Multiclass ROC curves (one-vs-rest) — Test set",
    x = "1 − Specificity",
    y = "Sensitivity"
  ) +
  theme(
    legend.position = 'bottom')
```

The multiclass ROC curves provide additional insight into discrimination
performance across the three classes.

For **diabetes** and **no-diabetes**, all models show strong
discriminative ability, with curves well above the diagonal reference
line. Differences between models are relatively small, though the
tree-based models appear slightly stronger in some regions of the curve.

For the **prediabetes class**, however, the multinomial logistic
regression model clearly outperforms both random forest and XGBoost. The
ROC curve for the multinomial model remains consistently above the other
two models, indicating better sensitivity–specificity trade-offs for
detecting this intermediate category.

### 6.4.2 Confusion Matrices

```{r}
#| label: rq1-conf-prep
#| message: false
#| warning: false
#| include: false

pred_rf  <- pred_rf  |> rename(.pred_No_diabetes = `.pred_No diabetes`)
pred_xgb <- pred_xgb |> rename(.pred_No_diabetes = `.pred_No diabetes`)
pred_mn  <- pred_mn  |> rename(.pred_No_diabetes = `.pred_No diabetes`)
plot_cm <- function(df, ttl) {
  cm <- conf_mat(df, truth = Diabetes, estimate = .pred_class)
  autoplot(cm, type = "heatmap") +
    labs(title = ttl, x = "Predicted", y = "True") 
}
```

**Diagnosing Prediabetes Classification Failure**

The confusion matrices reveal a critical pattern: **tree-based models
(Random Forest, XGBoost) completely fail to predict the Prediabetes
class**, collapsing to a binary No diabetes vs Diabetes distinction.
This occurs due to several factors:

-   **Class overlap:** Prediabetes represents an intermediate metabolic
    state sharing features with both No diabetes and Diabetes, creating
    ambiguous decision boundaries
-   **Minority class optimization:** Despite SMOTE augmentation,
    tree-based algorithms optimize splits based on impurity reduction.
    When Prediabetes remains the smallest class, trees preferentially
    split on features that distinguish the dominant classes (No diabetes
    vs Diabetes), ignoring subtle Prediabetes-specific patterns
-   **Cost function limitations:** Standard multi-class loss functions
    (e.g., cross-entropy) do not penalize misclassifying Prediabetes
    into adjacent categories differently than into distant ones, failing
    to capture ordinal structure

**Multinomial logistic regression**, by contrast, maintains some
Prediabetes prediction by directly modeling class probabilities without
hierarchical splitting, though accuracy remains modest. Alternative
approaches (e.g., ordinal regression, cost-sensitive learning, focal
loss) could better handle this intermediate class in future work.

```{r}
#| fig-height: 10
#| fig-width: 16

p_cm_xgb <- plot_cm(pred_xgb, "Confusion matrix — XGBoost")
p_cm_rf  <- plot_cm(pred_rf,  "Confusion matrix — Random forest")
p_cm_mn  <- plot_cm(pred_mn,  "Confusion matrix — Multinomial Logistic Regression")

(p_cm_xgb | p_cm_rf) / p_cm_mn +
  plot_layout(guides = "collect")
```

The confusion matrices reveal notable differences in classification
behaviour across the three models.

The **multinomial logistic regression model** is the only model that
correctly identifies a meaningful number of **prediabetes cases** (150
observations). In contrast, both **XGBoost and random forest completely
fail to predict the prediabetes class**, assigning all observations in
this category to either the no-diabetes or diabetes classes. This
indicates that the tree-based models effectively collapse the problem
into a two-class classification task, ignoring the intermediate
category.

For the **no-diabetes class**, all three models achieve high numbers of
correct predictions due to the dominance of this category in the
dataset. However, the multinomial model shows slightly higher correct
classifications than the boosted tree model.

For the **diabetes class**, all models correctly identify several
thousand cases, though misclassification between diabetes and
no-diabetes remains present across all models.

Overall, the multinomial model demonstrates a **more balanced
classification behaviour**, particularly by retaining predictive
capacity for the minority prediabetes class.

**Overall Assessment**

Although gradient boosting and random forest performed competitively
during cross-validation, the held-out test set evaluation suggests that
**multinomial logistic regression** provides the most balanced
performance across classes. In particular, it is the only model capable
of correctly identifying prediabetes cases while maintaining strong
discrimination for the other two classes.

## 6.5 Model Interpretation

**Triangulation Strategy for Model Interpretation**

To ensure robust identification of diabetes risk factors,
**methodological triangulation** is employed by comparing
interpretations across different analytical approaches:

-   **Multinomial logistic regression (Log-Odds):** Provides parametric
    coefficient estimates assuming linear additive effects.
    Interpretable but limited to linear relationships
-   **SHAP values (XGBoost):** Model-agnostic feature attribution
    capturing non-linear effects and interactions through tree-based
    partitioning
-   **Variable Importance (XGBoost gain):** Measures total loss
    reduction when features are used for splits, reflecting predictive
    utility

**Convergence between methods** strengthens confidence that identified
risk factors are robust across modelling assumptions. **Divergence**
indicates non-linearities, interactions, or violations of parametric
assumptions that only flexible models capture. Where findings conflict,
consistency patterns are prioritized and discrepancies are noted as
areas requiring further investigation.

### 6.5.1 Log-Odds Ratios (Multinomial Logistic Regression)

Multinomial logistic regression estimates the log-odds of belonging to
each diabetes category relative to a reference category. In this
analysis, **No diabetes** serves as the reference class.

Positive coefficients indicate **higher odds of belonging to the
specified diabetes category**, while negative coefficients indicate
**lower odds relative to the reference group**.

```{r}
#| label: rq1-logodds-prep
#| include: false


multinom_coefs <- multinom_final_fit |>
  extract_fit_parsnip() |>
  pluck("fit") |>
  coef()

multinom_coefs_tidy <- multinom_coefs |>
  as.data.frame() |>
  rownames_to_column(var = "y.level") |>
  pivot_longer(
    cols = -y.level, 
    names_to = "term", 
    values_to = "estimate"
  )

multinom_coefs_tidy <- multinom_coefs_tidy |>
  mutate(odds_ratio = exp(estimate))

```

```{r}
#| label: ods-plot-rq1
#| fig-height: 7
#| fig-width: 14
multinom_coefs_tidy |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = odds_ratio, y = reorder(term, odds_ratio))) +
  geom_point()  +
  geom_vline(xintercept = 1, linetype = "dashed") +
  facet_wrap(~ y.level) +
  labs(
    x = "Odds Ratio",
    y = "Predictor",
    title = "Odds ratios for predictors of diabetes status"
  )
```

**Multinomial Logistic Regression: Odds Ratio (OR) Summary**

*OR \> 1 indicates increased odds; OR \< 1 indicates decreased odds.*

**Synthesis of Findings**

-   **Primary Drivers:** Clinical indicators (Hypertension,
    Hypercholesterolemia, and BMI) represent the most significant risk.
    High blood pressure, in particular, yields a **139% increase** in
    the odds of diabetes ($OR = 2.39$).

-   **The BMI Gradient:** The data shows a clear progression where BMI
    impacts diabetes ($OR = 1.72$) more severely than prediabetes
    ($OR = 1.51$), confirming its role in disease advancement.

-   **Behavioral Mediation:** The lack of independent effect for
    physical activity ($OR = 1.00$) suggests its impact is likely
    **fully mediated** by metabolic variables like BMI and blood
    pressure.

-   **Statistical Anomalies:** The negative association with heavy
    alcohol ($OR = 0.31$) likely reflects **selection bias** (e.g.,
    "sick quitters" or health-conscious adjustments after diagnosis)
    rather than a biological protective effect.

**Conclusion**

Diabetes status in this population is best characterized by a
**multifactorial cardiometabolic profile**. While age and sex are
relevant, clinical markers of obesity and metabolic syndrome remain the
most effective predictors for distinguishing between healthy,
prediabetic, and diabetic individuals.

### 6.5.2 Global SHAP importance

**SHAP Interpretation Methodology**

To complement the parametric log-odds interpretation, we employ **SHAP
(SHapley Additive exPlanations)** to quantify feature contributions in
the XGBoost model. SHAP offers several advantages:

-   **Model-agnostic foundation** grounded in game theory, enabling
    comparison across model types
-   **TreeSHAP algorithm** provides exact Shapley values by leveraging
    XGBoost tree structure (no Monte Carlo approximation needed)
-   **Multiclass compatibility** computes separate contributions for
    each diabetes category (No diabetes, Prediabetes, Diabetes)
-   **Mean absolute SHAP values** reveal global feature importance by
    measuring how much each feature, on average, shifts predicted
    probabilities from baseline

By comparing SHAP importance with logistic regression odds ratios, we
assess method convergence—agreement strengthens confidence in identified
risk factors, while discrepancies highlight non-linearities the
parametric model cannot capture.

```{r}
#| label: shapviz-rq1

xgb_engine <- extract_fit_engine(xgb_final_fit)

test_processed <- bake(
  prep(extract_preprocessor(xgb_final_wf)),
  new_data = diab_test
)

X <- test_processed |>
  select(-Diabetes) |>
  as.matrix()

set.seed(123)
X_sample <- X[sample(nrow(X), 10000), ]

# shp <- shapviz(xgb_engine, X_pred = X_sample)
# write_rds(shp, "shp_rq1.rds")

# Load precomputed SHAP values to avoid recomputing expensive shapviz calculations
shp <- read_rds("shp_rq1.rds")


```

```{r}
#| label: global-shp-rq1
#| message: false
#| warning: false
#| fig-height: 7
#| fig-width: 14
sv_importance(shp, kind = "bar") +
  scale_fill_discrete(
    labels = c(
      "No diabetes",
      "Prediabetes",
      "Diabetes")
  )
```

> The SHAP importance plot indicates that cardiometabolic indicators
> dominate the predictive structure of the model. High blood pressure,
> general health status, high cholesterol, and BMI exhibit the largest
> mean absolute SHAP contributions across classes, indicating they play
> the strongest role in shifting predicted probabilities of diabetes
> status. Age also contributes meaningfully, while behavioural variables
> such as smoking, fruit consumption, and alcohol use show comparatively
> smaller effects. This pattern closely mirrors the multinomial logistic
> regression results, reinforcing the conclusion that diabetes status in
> this dataset is primarily driven by metabolic and health status
> indicators rather than lifestyle variables alone.

### 6.5.3 Variable Importance

**Variable Importance: Gain Metric**

XGBoost variable importance is computed using the **gain** metric, which
quantifies each feature's contribution to model performance:

-   **Gain** measures the total reduction in the loss function (e.g.,
    multi-class log-loss) achieved whenever a feature is used to split
    nodes across all trees in the ensemble
-   Features with high gain contribute substantially to improving
    predictions by creating splits that better separate diabetes
    categories
-   **Gain vs alternatives:** Unlike split frequency (which counts how
    often a feature is used) or cover (which weights by number of
    observations), gain directly captures predictive impact and effect
    magnitude

Higher gain values indicate features that most effectively reduce
prediction error when incorporated into the model's decision rules. This
metric complements SHAP importance by focusing on in-model utility
rather than marginal contributions to individual predictions.

```{r}
#| label: vip-rq1
#| fig-height: 7
#| fig-width: 14
xgb_final_fit |> 
  extract_fit_parsnip() |> 
  vip(num_features = 15, 
      geom = "col",
      aesthetics = list(fill = "steelblue",color = 'black', alpha = 0.8)) +
  labs(title = "Variable Importance: XGBoost Model",
       subtitle = "Top 15 predictors based on gain")

```

-   **Consistency with Statistical Findings:** The VIP confirms the
    results from the multinomial log-odds analysis (Section 6.5.1), with
    **HighBP**, **BMI**, and **HighChol** consistently appearing as the
    most influential predictors. This triangulation of methods
    strengthens the argument that these clinical markers are the primary
    determinants of diabetes risk in this dataset.

-   **The "Prediabetes" Paradox:** Despite the high importance scores
    for these top features, the model's struggle to classify the
    "Prediabetes" class suggests a "ceiling effect." While variables
    like **HighBP** and **GenHlth** (General Health) are excellent at
    distinguishing between "Healthy" and "Diabetic" individuals, the
    feature space lacks the specific granularity required to isolate the
    transitionary prediabetic phase.

-   **Lifestyle vs. Clinical Markers:** Interestingly, clinical markers
    (BMI, Blood Pressure) rank significantly higher than lifestyle
    factors such as **PhysActivity** or **Fruits**. This implies that
    while lifestyle choices contribute to the outcome, the physiological
    manifestations (the "clinical profile") are the more immediate and
    powerful predictors for the gradient-boosted trees.

## 6.6 Conclusion: Model Interpretability and Determinants

The comprehensive analysis of diabetes status, conducted through both
statistical and machine learning frameworks, demonstrates that diabetes
risk in this population is best characterized by a multifactorial
cardiometabolic profile. By triangulating the results from multinomial
log-odds, XGBoost Variable Importance, and SHAP values, this study
provides a robust validation of the primary drivers of the disease.

-   **Triangulation of Risk Factors:** There is a clear consensus across
    all analytical methods regarding the primary determinants of health.
    The multinomial logistic regression identified **High Blood
    Pressure** as the strongest categorical predictor, with an
    associated 139% increase in the odds of a diabetes diagnosis
    ($OR = 2.39$). This finding is directly mirrored in the **XGBoost
    Variable Importance Plot (VIP)**, where HighBP ranks as a top-tier
    feature, and is further clarified by the **SHAP summary plot**,
    which illustrates how high values for blood pressure, BMI, and
    cholesterol consistently push the model’s prediction toward the
    diabetic class.

-   **Predictive Nonlinearity and Impact:** While the linear multinomial
    model provides clear odds ratios, the SHAP analysis reveals the
    nuanced, non-linear impacts of features such as **Age** and
    **General Health**. The SHAP distribution reveals a significant
    clustering of high-risk impact for older age groups, suggesting that
    risk does not simply increase linearly, but rather accelerates as
    multiple clinical markers, such as BMI and cholesterol, coincide.

-   **The "Prediabetes" Gap and Structural Challenges:** A critical
    finding of this synthesis is the discrepancy in class identification
    across models. While the machine learning models (XGBoost and Random
    Forest) performed competitively in overall discrimination, they
    effectively collapsed the classification into a binary task, failing
    to isolate the **Prediabetes** class. The SHAP values show a
    significant overlap in the middle of the feature distribution,
    explaining this struggle, the clinical signature of prediabetes in
    this dataset is too similar to the surrounding classes for the
    boosting algorithm to partition effectively. Conversely, the
    multinomial logistic regression demonstrated superior balance,
    maintaining the ability to distinguish the subtle physiological
    transitions inherent in the intermediate prediabetic category.

-   **Socioeconomic and Actionable Clinical Profiles:** Beyond clinical
    indicators, general health perception and age remained significant
    predictors, yet lifestyle factors, such as physical activity and
    diet, showed weaker direct associations than physiological
    manifestations. This implies that while lifestyle choices are
    foundational, they may operate indirectly through their impact on
    measurable markers like BMI and blood pressure. The synthesis
    confirms that a high-risk profile is defined by the intersection of
    these clinical markers, which consistently dominate the top of both
    the VIP and SHAP rankings, identifying them as the most critical
    targets for screening and preventative intervention.

|  |  |  |
|------------------------|------------------------|------------------------|
| **Method** | **Key Strength** | **Primary Top Predictors** |
| **Multinomial Logistic** | Identifies the "Prediabetes" gradient | HighBP, BMI, Age |
| **XGBoost VIP (Gain)** | Highlights overall model efficiency | HighBP, GenHlth, BMI |
| **SHAP Analysis** | Shows direction and impact of features | HighBP, BMI, HighChol |

In summary, the triangulation of results from the **Odds Ratios**,
**SHAP values**, and **XGBoost Variable Importance** provides a robust
validation of the cardiometabolic risk factors. For clinical
application, the multinomial model remains the most "fit-for-purpose"
tool due to its unique capacity to identify individuals in the critical
prediabetic phase.

# 7. Socioeconomic Moderation of Cardiovascular Risk

## 7.1. Modelling Objective

This section addresses the following research question:

> *To what extent do socioeconomic determinants, specifically income and
> education, moderate the impact of clinical risk factors on the
> prevalence of heart disease, and can the capture of complex feature
> interactions provide superior predictive insights compared to
> traditional additive assumptions?*

The objective is:

-   **Interaction Analysis** – to investigate whether the predictive
    weight of clinical markers (e.g., BMI, Blood Pressure) shifts
    significantly across different socioeconomic strata.

-   **Predictive Optimization** – to evaluate if non-linear machine
    learning frameworks (XGBoost) outperform traditional linear
    frameworks (Logistic Regression) by capturing these latent
    moderating effects. **Operationalizing Moderation**

The research question asks whether socioeconomic factors (income,
education) moderate the effect of clinical risk factors (high blood
pressure, BMI) on heart disease. In this analysis, moderation is
operationalized as follows:

-   **For Logistic Regression:** Serves as an additive baseline without
    explicit interaction terms. Any moderation effects would require
    post-hoc interaction term inclusion, which is beyond the exploratory
    scope of this model.
-   **For XGBoost:** Interactions are captured implicitly through
    recursive tree partitioning. When the model splits on Income after
    splitting on HighBP (or vice versa), it effectively models
    conditional effects where the impact of one variable depends on the
    value of another.
-   **Evidence criterion:** Moderation will be assessed through SHAP
    interaction analysis and partial dependence plots stratified by SES
    groups. Significant variation in clinical risk effects across
    income/education levels would support the moderation hypothesis.

This approach leverages XGBoost's capacity to detect non-linear
interactions without requiring manual interaction term specification.

## 7.2. Outcome Variable

The dependent variable is **HeartDiseaseorAttack**, a binary outcome
indicating whether the respondent has ever been diagnosed with coronary
heart disease (CHD) or myocardial infarction (MI).

As identified in the exploratory phase, this variable exhibits a notable
class imbalance, with the "No" category significantly outnumbering the
"Yes" category. This necessitates the use of stratified sampling and
specific evaluation metrics, such as the Area Under the ROC Curve
(AUC-ROC) and F1-score, to ensure the model does not become biased
toward the majority class.

## 7.3 Training and Validation

### 7.3.1 Data Partitioning and Resampling

The dataset is partitioned into a 75% training set and a 25% testing set
using stratified sampling on heart disease status. To ensure the models
are robust and to prevent overfitting during the hyperparameter tuning
phase (particularly for the XGBoost model), 5-fold stratified
cross-validation is implemented within the training set. The test set
remains strictly held-out for the final comparative evaluation between
the additive logistic baseline and the interaction-heavy machine
learning model.

```{r}
#| label: partitioning-resampling-rq2
#| message: false

set.seed(42)

heart_split <- initial_split(
  diab, 
  prop = 0.75, 
  strata = HeartDiseaseorAttack
)

heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

heart_folds <- vfold_cv(
  heart_train, 
  v = 5, 
  strata = HeartDiseaseorAttack
)

heart_train |> count(HeartDiseaseorAttack) |> mutate(prop = n/sum(n))
heart_test  |> count(HeartDiseaseorAttack) |> mutate(prop = n/sum(n))
```

### 7.3.2 Recipe Definition and Preprocessing Strategy

**Class Imbalance Strategy**

The outcome exhibits substantial class imbalance (\~9% heart disease
cases). To prevent models from defaulting to majority-class predictions,
we apply **downsampling** with `under_ratio = 2.33`, retaining a 70-30
majority-to-minority ratio rather than full balancing. This approach
balances two objectives: (1) providing sufficient minority-class
representation for model learning (\~30% of downsampled data), and (2)
preserving moderate imbalance to reflect population base rates and avoid
miscalibrated probability estimates. Downsampling is applied within
cross-validation folds to prevent data leakage.

```{r}
#| label: recipe-rq2


heart_recipe <- recipe(HeartDiseaseorAttack ~ ., data = heart_train) |>
  step_nzv(all_predictors()) |>
  step_ordinalscore(Age, Education, Income, GenHlth) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(BMI, MentHlth, PhysHlth) |>
  step_downsample(HeartDiseaseorAttack, under_ratio = 2.33)

print(heart_recipe)
```

### 7.3.3 Model Specification

We define the candidate models for **RQ2** To address the research
question, we compare a traditional parametric baseline against a
high-capacity gradient-boosted ensemble. This allows us to determine if
the inclusion of non-linear feature interactions (socioeconomic
moderation) significantly improves the model's ability to identify
cardiovascular risk.

**The Models**

|  |  |  |  |
|------------------|------------------|------------------|------------------|
| **Model** | **Type** | **Primary Function** | **Key Strengths** |
| **Logistic Regression** | Parametric | Linear Baseline | Provides interpretable Odds Ratios for clinical and SES factors; assumes additive effects. |
| **XGBoost** | Ensemble (Boosting) | Interaction Discovery | Specifically optimized to capture non-linearities and moderation effects between variables. |

***Note***

> The selection of **XGBoost,** instead of Random Forest**,** for this
> research question is informed by its performance in the preceding
> analysis of diabetes status (RQ1). While RQ1 dealt with a multi-class
> outcome, XGBoost is equally well-suited for this binary classification
> of heart disease. Its gradient boosting framework is particularly
> effective at preventing the model from becoming biased toward the
> majority "No" class, ensuring that the clinical and socioeconomic
> patterns of the minority "Yes" class are accurately captured and
> weighted.

```{r}
#| label: model-specs-rq2
lr_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")


xgb_spec <- boost_tree(
  trees = 1024,
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  stop_iter = 20) |>
  set_engine("xgboost") |>
  set_mode("classification")
```

### 7.3.4 Workflow Set Definition

Following the methodology in Section 6, we utilize `workflow_sets` to
ensure both models are evaluated using the identical finalized recipe
and cross-validation folds.

```{r}
#| label: workflow-set-fit

heart_models <- workflow_set(
  preproc = list(base_rec = heart_recipe),
  models = list(logistic = lr_spec, xgboost = xgb_spec),
  cross = TRUE
)
```

### 7.3.5 Metric Selection and Model Execution

**Metrics**

To evaluate the models, we define a robust metric set that accounts for
the class imbalance inherent in the dataset. While **ROC-AUC** provides
a general measure of classifier quality, **PR-AUC** and the **F1-score**
are prioritized to ensure the model effectively identifies the "Yes"
(Heart Disease) cases without excessive false positives.

```{r}
#| label: metrics-rq2
heart_metrics <- metric_set(
  roc_auc,
  pr_auc,
  sens,
  spec,
  f_meas
)
```

**Model Execution and Resampling**

With the metrics defined, we map the `fit_resamples` function across our
workflow set. This process uses the 5-fold cross-validation established
in Section 7.3 to generate stable performance estimates.

```{r}
#| label: workflow-execution-rq2

# cl <- makePSOCKcluster(parallel::detectCores(logical = FALSE))
# registerDoParallel(cl)
# 
# heart_results <- heart_models |>
#   workflow_map(
#     fn = "tune_grid",
#     resamples = heart_folds,
#     grid = 10,
#     metrics = heart_metrics,
#     seed = 42,
#     control = control_grid(
#       save_pred = TRUE,
#       parallel_over = "everything",
#       verbose = TRUE,
#       event_level = "second")
#   )
# 
# stopCluster(cl)
# 
# write_rds(heart_results, "heart_results.rds")
heart_results <- read_rds('heart_results.rds')
```

### 7.3.6 Comparative Performance Visualization

```{r}
#| label: eval-results-rq2
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 14

m1 <- pretty_autoplot(heart_results, 
                      "roc_auc", "ROC–AUC", "Discrimination (ROC–AUC)")
m2 <- pretty_autoplot(heart_results, 
                      "pr_auc", "PR–AUC", "Precision-Recall AUC")
m3 <- pretty_autoplot(heart_results, 
                      "f_meas", "F1-score", "F1 (Imbalance-aware)")
m4 <- pretty_autoplot(heart_results, 
                      "sens", "Sensitivity", "Detection Rate (Recall)")

(m1 | m2) / (m3 | m4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

This figure compares the performance of all candidate workflows across
four evaluation metrics ranked from best to worst.

For ROC–AUC, most workflows achieve very similar discrimination
performance around the mid-0.80 range, indicating that both model
families are generally capable of separating the classes effectively.
Logistic regression workflows appear more stable across the ranking,
while some boosted tree configurations show a noticeable decline toward
the lower ranks, suggesting greater sensitivity to preprocessing or
tuning choices.

The Precision–Recall AUC plot shows lower absolute values, which is
expected in imbalanced classification settings. Both models perform
similarly across most workflows, although boosted tree models display
slightly greater variability, with a clearer performance drop among the
worst-ranked configurations.

The F1-score results show a narrow band among the higher-ranked
workflows, indicating that several configurations achieve a balanced
trade-off between precision and recall. Logistic regression workflows
maintain relatively consistent F1 values, whereas boosted tree models
again exhibit more spread, with some configurations performing
noticeably worse.

Finally, the Detection Rate (Recall) plot highlights differences in
sensitivity. Top-ranked workflows for both models achieve similar recall
levels, but logistic regression maintains more consistent detection
rates across ranks. Some boosted tree workflows show reduced recall in
lower ranks, indicating weaker ability to identify positive cases in
those configurations.

Overall, logistic regression workflows demonstrate more stable
performance across preprocessing variations, while boosted tree models
show slightly greater variability. Despite this, the best-performing
configurations of both models achieve broadly comparable predictive
performance across all evaluation metrics.

## 7.4 Final Model Extraction and Test Set Evaluation

This stage identifies the best-performing workflow for each modelling
approach based on cross-validated tuning results. The logistic
regression and XGBoost workflows with the highest Precision–Recall AUC
are selected, as PR–AUC provides a more informative evaluation metric in
the presence of class imbalance.

For each model type, the optimal hyperparameter configuration obtained
during tuning is applied to the corresponding workflow, producing a
finalised model specification. These final workflows therefore represent
the best-performing logistic regression and boosted tree models
identified during the resampling stage.

The purpose of this step is to prepare the final candidate models for
unbiased evaluation on the held-out test dataset, enabling a direct
comparison of their predictive performance.

```{r}
#| label: final-comparison-rq2

best_lr_id <- "base_rec_logistic"
best_xgb_id <- "base_rec_xgboost"

final_lr_wflow <-
  heart_results |>
  extract_workflow(best_lr_id) |>
  finalize_workflow(
    heart_results |>
      extract_workflow_set_result(best_lr_id) |>
      select_best(metric = "pr_auc")
  )

final_xgb_wflow <-
  heart_results |>
  extract_workflow(best_xgb_id) |>
  finalize_workflow(
    heart_results |>
      extract_workflow_set_result(best_xgb_id) |>
      select_best(metric = "pr_auc")
  )
```

The finalised logistic regression and XGBoost models are evaluated on
the held-out test set to obtain an unbiased estimate of predictive
performance.

This evaluation provides a fair comparison of the two modelling
approaches under identical conditions. The resulting performance metrics
summarise how well each model generalises to unseen data, allowing the
relative effectiveness of logistic regression and the boosted tree model
to be assessed for the prediction task.

```{r}
#| label: final-pred-rq2
test_results_lr <- last_fit(
  final_lr_wflow, split = heart_split, metrics = heart_metrics)

test_results_xgb <- last_fit(
  final_xgb_wflow, split = heart_split, metrics = heart_metrics)
```

```{r}
#| label: comaprison-results-rq2
final_comparison <- bind_rows(
  test_results_lr |> collect_metrics() |> mutate(model = "Regularized LR"),
  test_results_xgb |> collect_metrics() |> mutate(model = "Tuned XGBoost")
) |> 
  select(model, .metric, .estimate) |> 
  pivot_wider(names_from = .metric, values_from = .estimate)

print(final_comparison)
```

The final test-set evaluation compares the predictive performance of the
Regularized Logistic Regression model and the Tuned XGBoost model across
multiple metrics.

The Regularized Logistic Regression model achieves a sensitivity of
1.00, specificity of 0.00, F1-score of 0.950, ROC–AUC of 0.50, and
PR–AUC of 0.953. While the high sensitivity and F1-score may initially
appear strong, the specificity of zero indicates that the model predicts
all observations as the positive class. This behaviour results in no
ability to correctly identify negative cases, which is reflected in the
ROC–AUC of 0.50, equivalent to random classification. Consequently, the
model fails to provide meaningful class discrimination despite
apparently high values on certain metrics.

In contrast, the Tuned XGBoost model demonstrates substantially stronger
and more balanced predictive performance. It achieves a sensitivity of
0.882, specificity of 0.550, F1-score of 0.915, ROC–AUC of 0.840, and
PR–AUC of 0.979. The high ROC–AUC indicates strong discrimination
between classes, while the high PR–AUC suggests effective performance
under class imbalance. Although the specificity is moderate, the model
successfully identifies a meaningful proportion of both positive and
negative cases, resulting in a more balanced and practically useful
classifier.

Overall, the results indicate that the Tuned XGBoost model substantially
outperforms the Regularized Logistic Regression model. The boosted tree
model provides strong discrimination and maintains a reasonable balance
between sensitivity and specificity, suggesting that the underlying
relationships between predictors and heart disease are better captured
by a flexible non-linear model than by a linear logistic framework.

```{r}
#| label: result-bind
all_preds <- bind_rows(
  test_results_lr |> collect_predictions() |> mutate(model = "Logistic Regression"),
  test_results_xgb |> collect_predictions() |> mutate(model = "XGBoost")
)
```

### 7.4.3 Confusion Matrix Analysis

```{r}
#| label: conf-mat-rq2
#| fig-height: 6
#| fig-width: 12

cm_lr <- test_results_lr |> 
  collect_predictions() |> 
  conf_mat(HeartDiseaseorAttack, .pred_class) |> 
  autoplot(type = "heatmap") + 
  labs(title = "LR: Predicting everyone has Heart Disease")

cm_xgb <- test_results_xgb |> 
  collect_predictions() |> 
  conf_mat(HeartDiseaseorAttack, .pred_class) |> 
  autoplot(type = "heatmap") + 
  labs(title = "XGBoost: Balanced Predictions")

cm_lr + cm_xgb
```

The confusion matrices provide a clearer interpretation of how each
model performs at the class prediction level.

The regularized logistic regression model predicts every observation as
the positive class (Heart Disease). As a result, all true negative cases
(57,436) are misclassified as positive, while all true positive cases
(5,984) are also predicted as positive. This explains the perfect
sensitivity (1.00) and zero specificity observed earlier. Although the
model detects all positive cases, it completely fails to identify
individuals without heart disease, demonstrating that it has collapsed
into a trivial classifier that does not meaningfully distinguish between
classes.

In contrast, the tuned XGBoost model produces more balanced predictions
across both classes. It correctly identifies 50,682 individuals without
heart disease and 3,289 individuals with heart disease, while
misclassifying 6,754 non-cases as positive and 2,695 true cases as
negative. This reflects a more balanced trade-off between sensitivity
and specificity, consistent with the stronger ROC–AUC and overall
performance observed in the test-set evaluation.

Overall, the confusion matrices reinforce the earlier metric-based
comparison. Logistic regression fails to provide meaningful class
separation, whereas XGBoost demonstrates effective discrimination by
correctly identifying a substantial proportion of both positive and
negative cases.

### 7.4.4 Variable Importance Analysis

```{r}
#| label: vip-rq2
#| fig-height: 7
#| fig-width: 14

final_xgb_fit <- extract_fit_parsnip(test_results_xgb)

final_xgb_fit |> 
  vip(num_features = 20, 
      geom = "col",
      aesthetics = list(fill = "steelblue",color = 'black', alpha = 0.8)) +
  labs(title = "Variable Importance: XGBoost Model",
       subtitle = "Top 15 predictors based on gain")
```

> The variable importance plot presents the relative contribution of
> predictors to the final XGBoost model based on gain, which measures
> how much each variable improves the model’s decision splits.
>
> General health (GenHlth) emerges as the most influential predictor,
> followed closely by age. This indicates that overall health status and
> age group are the strongest contributors to distinguishing individuals
> with and without heart disease in the model. High blood pressure and
> high cholesterol also rank highly, reinforcing the importance of
> cardiometabolic risk factors in predicting heart disease outcomes.
>
> Several additional variables contribute moderate predictive value,
> including sex, difficulty walking, smoking status, and diabetes
> status. These predictors likely capture functional limitations and
> behavioural risk factors associated with cardiovascular health.
>
> In contrast, lifestyle variables such as fruit consumption, vegetable
> consumption, physical activity, and alcohol use show minimal
> importance within the model. Socioeconomic indicators such as
> education and income also appear to contribute relatively little
> compared with clinical and health-status variables.
>
> Overall, the importance ranking suggests that the XGBoost model
> primarily relies on indicators of general health and cardiometabolic
> risk, with demographic and behavioural factors playing a secondary
> role in predicting heart disease status.

## 7.5 Formal Tests of Socioeconomic Moderation

### 7.5.1 Logistic Regression Interaction Models

To formally evaluate socioeconomic moderation, logistic regression
models including interaction terms between key clinical risk factors
(HighBP, HighChol, BMI) and socioeconomic indicators (Income and
Education) were estimated. These interaction terms allow the association
between clinical risk factors and heart disease to vary across
socioeconomic strata.

**1. Baseline main-effects model**

```{r}
dim(model.matrix(
  HeartDiseaseorAttack ~ HighBP + HighChol + BMI +
    Age + Sex + Income + Education +
    Smoker + PhysActivity,
  data = heart_train
))
```

```{r}
#| label: base-rq2
mod_main <- glm(
  HeartDiseaseorAttack ~ HighBP + HighChol + BMI +
    Age + Sex + Income + Education +
    Smoker + PhysActivity,
  data = heart_train,
  family = binomial()
)
```

**2. Income interaction model**

```{r}
#| label: income-inter-rq2
mod_income <- glm(
  HeartDiseaseorAttack ~ (HighBP + HighChol + BMI) * Income +
    Age + Sex + Education +
    Smoker + PhysActivity,
  data = heart_train,
  family = binomial()
)
```

**3. Education interaction model**

```{r}
#| label: edu-inter-rq2
mod_edu <- glm(
  HeartDiseaseorAttack ~ (HighBP + HighChol + BMI) * Education +
    Age + Sex + Income +
    Smoker + PhysActivity,
  data = heart_train,
  family = binomial()
)

```

## 7.5.2 Likelihood Ratio Tests

This subsection formally tests whether adding interaction terms
**improves model fit** relative to the baseline model.

**1. Income moderation test**

```{r}
lrt_income <- anova(mod_main, mod_income, test = "Chisq")
```

**2. Education moderation test**

```{r}
lrt_edu <- anova(mod_main, mod_edu, test = "Chisq")
```

```{r}
lrt_income
lrt_edu
```

### 7.5.3 Interpretation of Moderation Effects

Likelihood ratio tests were used to compare the baseline logistic
regression model containing only main effects with models that
additionally included interaction terms between key clinical risk
factors (HighBP, HighChol, BMI) and socioeconomic indicators (Income and
Education). These tests assess whether allowing the effects of clinical
predictors to vary across socioeconomic strata improves model fit.

The model including interactions between clinical risk factors and
income significantly improved model fit relative to the baseline
specification ($Δχ² = 49.40, df = 21, p < 0.001$). Similarly, the model
including interactions between clinical predictors and education also
produced a statistically significant improvement in fit
($Δχ² = 31.18, df = 15, p = 0.008$). These results indicate that the
relationships between key cardiometabolic risk factors and heart disease
are not constant across socioeconomic groups. Instead, the strength of
these associations varies across levels of income and educational
attainment.

However, while statistically significant, these interaction effects
should be interpreted with caution. Given the large sample size of the
dataset (approximately 190,000 observations), even relatively small
deviations from the additive model can produce statistically significant
likelihood ratio tests. As such, the presence of moderation should be
interpreted as evidence that socioeconomic position contributes to
heterogeneity in the association between clinical risk factors and heart
disease risk, rather than as evidence of large effect modification.

To further explore the structure and magnitude of these interaction
effects in a more flexible modelling framework, the following section
uses SHAP-based interpretation of the XGBoost model to examine how
predicted contributions of clinical risk factors vary across
socioeconomic strata.

## 7.6 SHAP-Based Model Interpretation

The likelihood ratio tests provided formal evidence that interactions
between key clinical risk factors and socioeconomic indicators improved
model fit. To further examine the structure of these moderation effects,
SHAP-based interpretation of the final XGBoost model was used to explore
how interaction patterns varied across observations in a nonlinear
setting.

```{r}
#| label: shapviz-rq2
# xgb_engine <- extract_fit_engine(final_xgb_fit)
# 
# test_processed <- bake(
#   prep(extract_preprocessor(final_xgb_wflow)),
#   new_data = testing(heart_split)
# )
# 
# X <- test_processed |> 
#   select(-HeartDiseaseorAttack) |> 
#   as.matrix()
# 
# set.seed(123)
# X_sample <- X[sample(nrow(X), 10000), ]
# shp <- shapviz(xgb_engine, X_pred = X_sample)
# 
# write_rds(shp,"shp_rq2.rds")


# Load precomputed SHAP values to avoid recomputing expensive shapviz calculations
shp <- read_rds("shp_rq2.rds")

```

The SHAP values are subsequently sign-adjusted to align with the outcome
orientation used in this analysis. In this dataset the positive event is
encoded as the second level of the outcome factor (No = 0, Yes = 1),
whereas the SHAP implementation assumes that the first level corresponds
to the event of interest. As a result, the SHAP values are multiplied by
−1 so that positive contributions correctly represent increased
predicted risk of heart disease and negative contributions represent
decreased risk. This adjustment ensures that subsequent SHAP
visualisations are interpreted consistently with the outcome definition
used throughout the analysis.

```{r}
#| label: flip-shp
shp_flipped <- shapviz(-shp$S, X = shp$X, baseline = -shp$baseline)
```

### 7.5.1 Global Feature Importance (SHAP)

The SHAP feature importance plot ranks predictors according to their
average absolute SHAP value, which represents the overall magnitude of
each variable’s contribution to the model’s predictions across all
observations.

```{r}
#| label: global-shp-rq2
#| dev: png
#| fig-height: 7
#| fig-width: 14
sv_importance(shp_flipped, kind = "bar", max_display = 15) +
  labs(title = "SHAP Feature Importance", x = "Mean |SHAP Value|")
```

> Age emerges as the most influential predictor, indicating that changes
> in age group have the largest average effect on predicted heart
> disease risk. General health (GenHlth) is the second most important
> variable, suggesting that self-reported health status strongly
> influences the model’s assessment of cardiovascular risk.
>
> Cardiometabolic indicators such as high blood pressure (HighBP) and
> high cholesterol (HighChol) also show substantial importance,
> reinforcing their established role as key risk factors for
> cardiovascular disease. These variables contribute strongly to the
> model’s decision structure and appear consistently among the top
> predictors.
>
> Several demographic and behavioural variables provide moderate
> contributions, including sex, smoking status, difficulty walking, and
> diabetes status. These predictors likely capture functional health
> limitations and behavioural risk factors associated with
> cardiovascular conditions.
>
> In contrast, variables such as BMI, mental health, alcohol
> consumption, education, and healthcare cost barriers show relatively
> small mean SHAP values. This suggests that, within the context of the
> full model, they contribute comparatively little to the prediction of
> heart disease status.
>
> Overall, the SHAP importance ranking broadly aligns with the earlier
> model-based variable importance results, highlighting age, general
> health, and cardiometabolic risk factors as the dominant predictors
> driving the XGBoost model’s predictions.

### 7.5.2 Distribution of Feature Effects (SHAP Beeswarm Plot)

The SHAP beeswarm plot provides a detailed view of how individual
feature values influence the predicted probability of heart disease
across observations. Each point represents a single observation, where
the horizontal position indicates the SHAP value (the contribution of
the feature to the prediction), and colour represents the magnitude of
the feature value.

```{r}
#| label: shap-beesawrm-rq2
#| dev: png
#| fig-height: 7
#| fig-width: 14
sv_importance(shp_flipped, kind = "beeswarm", max_display = 15) +
  labs(title = "SHAP Beeswarm Plot")
```

> Age shows the widest spread of SHAP values, indicating that it has the
> strongest and most consistent influence on model predictions. Higher
> age categories are associated with positive SHAP values, meaning they
> increase the predicted risk of heart disease, whereas lower age values
> tend to reduce predicted risk.
>
> General health (GenHlth) also demonstrates a strong directional
> effect. Poorer health levels (higher values) shift predictions toward
> higher heart disease risk, while better self-reported health
> contributes negatively to the prediction. Similarly, clinical risk
> factors such as high blood pressure and high cholesterol consistently
> push predictions toward increased risk when present.
>
> Demographic and behavioural factors such as sex, smoking, and
> difficulty walking exhibit moderate contributions, where the presence
> of these conditions tends to increase predicted risk but with smaller
> magnitudes compared to primary clinical indicators.
>
> Socioeconomic variables, particularly income and education, display
> comparatively smaller SHAP magnitudes. Their distributions are centred
> closer to zero, indicating that while they contribute to predictions,
> their direct effects are weaker than clinical risk factors. This
> suggests that socioeconomic determinants may influence predictions
> more subtly, potentially through interactions with health-related
> variables rather than through strong independent effects.
>
> Overall, the beeswarm distribution reinforces that the model’s
> predictions are primarily driven by clinical health indicators, while
> demographic and socioeconomic factors contribute secondary adjustments
> to predicted heart disease risk.

### 7.6.3 Global Interaction Strengths

To examine whether predictors jointly influence model predictions, SHAP
interaction values were computed using the `sv_interaction()` function.
Unlike standard SHAP importance measures, which quantify the independent
contribution of each feature, SHAP interaction values capture how pairs
of predictors interact within the model to influence predicted heart
disease risk. This provides a global overview of the strength of
pairwise interactions learned by the XGBoost model.

```{r}
#| label: shap-interaction
if (!file.exists("shp_i.rds")) {
  shp_i <- shapviz(xgb_engine, X_pred = X_sample, interactions = TRUE)
  write_rds(shp_i, "shp_i.rds")
} else {
  shp_i <- read_rds("shp_i.rds")
}
```

```{r}
#| label: shp-pairwise
#| fig-height: 10
#| fig-width: 18
sv_interaction(shp_i)
```

The interaction summary plot indicates that most pairwise interactions
in the model are relatively small, with the majority of SHAP interaction
values concentrated near zero. This suggests that much of the predictive
signal in the model arises from additive effects of individual
predictors rather than strong nonlinear interactions.

Among the predictors displayed, age and general health status exhibit
the largest overall influence on predicted heart disease risk. The
distribution of SHAP values for age shows a clear gradient, where higher
age values are associated with increasingly positive SHAP contributions,
indicating a higher predicted risk of heart disease. Similarly, poorer
self-reported general health is associated with positive SHAP
contributions, reflecting its strong relationship with cardiovascular
outcomes.

Clinical risk factors such as high blood pressure and high cholesterol
also show visible contributions, although their interaction effects with
other variables appear relatively modest. The SHAP values for these
variables are clustered around small positive values when the risk
factor is present, indicating that they consistently increase predicted
risk but do not strongly interact with other predictors in a nonlinear
manner.

Behavioural and functional indicators such as smoking status and
difficulty walking display smaller interaction magnitudes overall. While
these variables contribute to predictions individually, the interaction
plot suggests that their joint effects with other predictors are limited
relative to major clinical and demographic risk factors.

Overall, the interaction summary indicates that the model is primarily
driven by strong main effects, with interactions playing a secondary
role. Nevertheless, the presence of some spread in the interaction
distributions suggests that certain predictors may still exhibit
heterogeneous effects across subgroups. The following subsection
therefore examines targeted SHAP dependence plots to explore how
clinical risk factors interact with socioeconomic indicators such as
income and education.

### 7.6.4 Targeted Dependence Analysis

To examine the moderation patterns more directly, targeted SHAP
dependence plots areproduced for the most relevant clinical risk
factor–socioeconomic variable combinations. Based on the formal
interaction tests and the substantive focus of RQ2, attention was
restricted to interactions between key cardiometabolic risk factors
(`HighBP`, `HighChol`) and the socioeconomic indicators `Income` and
`Education`.

#### **Income interactions**

```{r}
#| label: income-trgt-dependence
#| dev: png
#| fig-height: 10
#| fig-width: 14

p_inc_bp <- sv_dependence(shp_i,v = "HighBP_High",
                          color_var = "Income",
                          interactions = TRUE) +
  labs(title = "High Blood Pressure by Income",
       x = "HighBP",
       y = "Interaction SHAP Value")

p_inc_chol <- sv_dependence(shp_i,v = "HighChol_High",
                            color_var = "Income",
                            interactions = TRUE) +
  labs(title = "High Cholesterol by Income",
    x = "HighChol",
    y = "Interaction SHAP Value")

p_inc_bmi <- sv_dependence(shp_i,v = "BMI",
                           color_var = "Income",
                           interactions = TRUE) +
  labs(title = "BMI by Income",x = "BMI",y = "Interaction SHAP Value")

p_inc_gen <- sv_dependence(shp_i,v = "GenHlth",
                           color_var = "Income",interactions = TRUE) +
  labs(title = "General Health by Income",x = "General Health",
       y = "Interaction SHAP Value")

(p_inc_bp | p_inc_chol) / (p_inc_gen | p_inc_bmi)
```

> The dependence plots illustrate how the contribution of key clinical
> risk factors to heart disease prediction varies across different
> income levels.
>
> For high blood pressure, the SHAP values shift clearly from negative
> to positive when the condition is present, indicating that individuals
> with high blood pressure are predicted to have a substantially higher
> risk of heart disease. The colour distribution across the two clusters
> shows little systematic separation by income level, suggesting that
> the effect of high blood pressure on predicted risk is relatively
> consistent across socioeconomic groups.
>
> A similar pattern is observed for high cholesterol. When high
> cholesterol is present, SHAP values increase markedly, indicating an
> increased contribution to predicted heart disease risk. As with high
> blood pressure, the colour gradient representing income does not
> display strong vertical separation within the clusters, implying that
> income does not strongly moderate the impact of high cholesterol on
> model predictions.
>
> General health demonstrates a strong monotonic relationship with SHAP
> values. As self-reported health worsens (moving from excellent toward
> poor health), the SHAP contribution increases steadily, indicating a
> progressively higher predicted risk of heart disease. While
> observations from different income levels are present within each
> health category, the dominant pattern remains the effect of worsening
> health rather than a clear socioeconomic interaction.
>
> The BMI dependence plot shows a weaker and more diffuse pattern. SHAP
> values remain close to zero across most BMI values, with only modest
> variation in predicted risk. Income levels are mixed throughout the
> distribution, again indicating limited evidence that income
> substantially modifies the influence of BMI on predicted heart disease
> risk.
>
> Overall, these results suggest that the primary drivers of heart
> disease prediction in the model are the clinical risk factors
> themselves. Although the XGBoost model is capable of capturing complex
> feature interactions, the dependence plots provide limited visual
> evidence that income strongly moderates the impact of these clinical
> predictors on predicted heart disease prevalence.

#### **Clinical Risk Factors by Education**

```{r}
#| label: edu-dependence
#| dev: png
#| fig-height: 10
#| fig-width: 14

p_edu_bp <- sv_dependence(shp_i,v = "HighBP_High",
                          color_var = "Education",
                          interactions = TRUE) +
  labs(title = "High Blood Pressure by Education",x = "HighBP",
       y = "Interaction SHAP Value")

p_edu_chol <- sv_dependence(shp_i,v = "HighChol_High",
                            color_var = "Education",
                            interactions = TRUE) +
  labs(title = "High Cholesterol by Education",
       x = "HighChol",
       y = "Interaction SHAP Value")

p_edu_bmi <- sv_dependence(shp_i,v = "BMI",
                           color_var = "Education",
                           interactions = TRUE) +
  labs(title = "BMI by Education",x = "BMI",y = "Interaction SHAP Value")

p_edu_gen <- sv_dependence(shp_i,v = "GenHlth",color_var = "Education",
                           interactions = TRUE) +
  labs(title = "General Health by Education",x = "General Health",
       y = "Interaction SHAP Value")

(p_edu_bp | p_edu_chol) / (p_edu_gen | p_edu_bmi)
```

> The SHAP dependence plots examining interactions between clinical
> predictors and education levels show generally weak interaction
> effects across the variables considered. In contrast to the income
> plots, most SHAP interaction values are tightly clustered around zero,
> indicating that education has limited influence on how these clinical
> factors contribute to predicted heart disease risk within the model.
>
> For high blood pressure, the interaction values are effectively zero
> across all observations. Both the absence and presence of high blood
> pressure show no visible spread in SHAP interaction values, and the
> colour distribution representing education levels appears uniform
> across the points. This suggests that the contribution of high blood
> pressure to predicted heart disease risk is largely independent of
> educational attainment in the model.
>
> The high cholesterol plot shows slightly more variation than high
> blood pressure, but the magnitude of interaction effects remains very
> small. A few observations exhibit minor positive or negative
> interaction values, particularly among lower education levels, but
> these deviations are sparse and remain close to zero overall. As a
> result, there is little evidence that education meaningfully alters
> the predicted risk contribution of high cholesterol.
>
> The interaction between general health and education also appears
> minimal. While poorer self-reported health is associated with
> increasing main SHAP contributions elsewhere in the analysis, the
> interaction values here remain very small and centered around zero.
> Education levels are distributed across the entire range of SHAP
> values without any clear systematic gradient.
>
> Similarly, BMI shows virtually no interaction with education. The SHAP
> interaction values form a flat horizontal band at zero across the
> entire range of BMI values, indicating that the model does not detect
> any meaningful interaction between body mass index and educational
> attainment when predicting heart disease risk.
>
> Overall, these results suggest that educational attainment does not
> substantially modify the relationship between the examined clinical
> risk factors and predicted heart disease risk in the XGBoost model.
> Although the earlier likelihood ratio test indicated a statistically
> significant interaction block for education, the SHAP visualisations
> suggest that the magnitude of these interaction effects is very small
> and unlikely to be substantively important in practical terms.

### 7.6.5 Local Explanations

To complement the global and interaction-level interpretations, local
explanation plots were generated for individual predictions. These plots
decompose a single prediction into feature-level contributions,
illustrating how different variables increase or decrease predicted
heart disease risk relative to the model baseline.

```{r}
#| label: high-low-risk-shap
#| dev: png
#| fig-height: 10
#| fig-width: 15

low_risk_idx <- which.min(rowSums(as.matrix(shp_flipped$S)))
high_risk_idx <- which.max(rowSums(as.matrix(shp_flipped$S)))

w1 <- sv_waterfall(shp_flipped, row_id = high_risk_idx) + 
  labs(title = "Waterfall: Highest Risk Individual",
       subtitle = "Features pushing prediction above baseline",
       x = "SHAP Contribution (log-odds)")

w2 <- sv_waterfall(shp_flipped, row_id = low_risk_idx) + 
  labs(title = "Waterfall: Lowest Risk Individual",
       subtitle = "Features pushing prediction below baseline",
       x = "SHAP Contribution (log-odds)")

f1 <- sv_force(shp_flipped, row_id = high_risk_idx) + 
  labs(title = "Force Plot: Highest Risk Individual",
       subtitle = "Red = increases risk | Blue = decreases risk")

f2 <- sv_force(shp_flipped, row_id = low_risk_idx) + 
  labs(title = "Force Plot: Lowest Risk Individual",
       subtitle = "Red = increases risk | Blue = decreases risk")

(w1 | w2) 

(f1 / f2)
```

The local SHAP explanations illustrate how the model combines individual
features to generate predictions for specific observations. Two
contrasting cases are shown: the individual with the highest predicted
risk of heart disease and the individual with the lowest predicted risk.

For the highest-risk individual, the waterfall and force plots show that
several clinical and demographic risk factors push the prediction
substantially above the baseline. Poor general health (GenHlth = 5)
provides the largest positive contribution to the prediction, followed
by older age (Age = 13). Additional factors such as the presence of high
blood pressure, male sex, diabetes, difficulty walking, and high
cholesterol all contribute positively to the predicted risk. Together,
these factors accumulate to shift the model output well above the
baseline log-odds, indicating a strong predicted likelihood of heart
disease.

In contrast, the lowest-risk individual shows a sequence of negative
SHAP contributions that push the prediction far below the baseline.
Younger age (Age = 2) provides the strongest protective effect, followed
by excellent self-reported health (GenHlth = 1). The absence of major
clinical risk factors, including high blood pressure, high cholesterol,
smoking, and diabetes, further reduces the predicted risk. Female sex
and higher income also contribute modestly toward lowering the
prediction.

### Summary of SHAP-Based Model Interpretation

**Global Feature Importance and Main Effects.**\
The SHAP importance and distribution plots indicate that the model is
primarily driven by strong main effects from demographic and clinical
variables, particularly age and general health status. These predictors
consistently produce the largest SHAP contributions, suggesting that
worsening health status and increasing age substantially raise predicted
heart disease risk. Other clinical risk factors such as high blood
pressure and high cholesterol also contribute positively to predicted
risk, though to a lesser extent.

**Interaction Structure of the Model.**\
The SHAP interaction analysis shows that most predictor relationships
are largely additive. Pairwise interaction values are generally
concentrated near zero, indicating that the model relies mainly on
independent feature effects rather than strong nonlinear interactions.
Although earlier logistic regression tests detected statistically
significant socioeconomic interactions, the SHAP interaction magnitudes
appear relatively small in practice.

**Income-Based Heterogeneity.**\
Targeted dependence plots suggest some variation in how clinical risk
factors contribute to predicted risk across income levels. In
particular, high blood pressure, high cholesterol, and general health
show modest differences in SHAP contributions across income categories.
Lower income groups occasionally display slightly larger positive
contributions, indicating some heterogeneity in risk accumulation across
socioeconomic strata. However, BMI shows little interaction with income.

**Education-Based Effects.**\
In contrast, the education-based dependence plots show minimal
interaction effects. SHAP interaction values for high blood pressure,
BMI, and general health remain tightly clustered around zero across
education levels, suggesting that educational attainment does not
substantially alter the model’s estimated contribution of these
predictors. This indicates that while statistical interaction terms were
detected earlier, their practical magnitude appears small.

**Local Prediction Explanations.**\
Local SHAP explanations illustrate how individual predictions are formed
by combining multiple feature contributions. High-risk individuals are
typically characterised by older age, poorer general health, and the
presence of cardiometabolic risk factors such as hypertension or
diabetes. Conversely, low-risk individuals tend to be younger, report
better health, and lack major cardiovascular risk factors. These
case-level explanations confirm that the model’s predictions align with
well-established epidemiological risk patterns.

**Overall Interpretation.**\
Taken together, the SHAP analyses suggest that cardiovascular risk
predictions are driven primarily by demographic and clinical factors,
with socioeconomic variables contributing only modestly through
interaction effects. While some socioeconomic heterogeneity exists,
particularly across income groups, the overall structure of the model
remains largely additive, with clinical health indicators playing the
dominant role in predicted heart disease risk.

## 7.6 Conclusions: Discussion and Limitations

This section addressed the second research question by investigating
whether complex machine learning models can capture interactions between
socioeconomic determinants and clinical risk factors when predicting
heart disease prevalence. To do this, a modelling framework was
implemented using the tidymodels ecosystem, comparing a regularized
logistic regression model with a tuned XGBoost model. Logistic
regression was included as a baseline representing traditional additive
modelling assumptions, while XGBoost was used to capture potential
non-linear relationships and interaction effects within the data.

Models were trained using cross-validated tuning and evaluated on a
held-out test set using metrics appropriate for imbalanced
classification, including ROC–AUC, PR–AUC, F1-score, sensitivity, and
specificity. The final evaluation demonstrated that the tuned XGBoost
model substantially outperformed logistic regression in terms of
discrimination and overall predictive performance. Logistic regression
effectively collapsed to predicting the majority class, highlighting the
limitations of a simple additive framework for this prediction task.

To interpret the behaviour of the XGBoost model, SHAP-based
explainability techniques were applied. Global importance and beeswarm
plots showed that predictions were primarily driven by clinical and
health-related predictors such as age, general health status, high blood
pressure, and high cholesterol. Dependence plots confirmed clear
monotonic relationships between worsening health indicators and
increasing predicted heart disease risk.

A central aim of this section was to evaluate whether socioeconomic
determinants, specifically income and education, moderate the impact of
these clinical risk factors. SHAP interaction analyses were therefore
conducted to isolate pairwise feature interactions. The results
indicated that although the XGBoost model is capable of capturing
non-linear and interaction effects, the moderation of clinical risk
factors by income and education appears relatively limited. Interaction
SHAP values for combinations such as high blood pressure × income, high
cholesterol × income, and general health × education remained small
compared with the direct effects of the clinical variables themselves.

These findings suggest that, within this dataset, the prevalence of
heart disease is primarily associated with established clinical risk
factors rather than strong socioeconomic moderation of those factors.
While socioeconomic variables contribute to the predictive model, their
influence appears secondary compared with physiological and
health-status indicators. Consequently, the use of a flexible machine
learning model provided improved predictive performance and
interpretability, but did not reveal strong interaction structures
between socioeconomic and clinical predictors.

Several limitations should be considered when interpreting these
results. First, the analysis relies on self-reported survey data, which
may contain measurement error or reporting bias. Second, the variables
used to represent socioeconomic status are relatively coarse indicators,
which may limit the ability to detect nuanced socioeconomic
interactions. Third, the dataset is cross-sectional, meaning causal
relationships cannot be inferred from the observed associations.
Finally, although SHAP provides powerful interpretability tools, the
explanations remain dependent on the structure of the fitted model and
should be interpreted as model-based insights rather than definitive
causal effects.

Overall, this section demonstrates that machine learning models such as
XGBoost can provide improved predictive performance and detailed
interpretability through explainable AI techniques. However, in the
context of this dataset, the dominant determinants of predicted heart
disease risk remain clinical health indicators, with socioeconomic
moderation effects appearing relatively modest.

# 8. Socioeconomic and Lifestyle Determinants of Mental Health Burden

#### Data Modification

Numeric versions of ordinal predictors for count models (pscl does not
handle ordered factors cleanly)

```{r}
diab_zinb <- diab |>
  mutate(
    Income    = as.integer(Income),
    Education = as.integer(Education),
    Age = as.integer(Age),
    PhysActivity = as.integer(PhysActivity == "Yes"),
    Smoker = as.integer(Smoker == "Yes"),
    Sex = as.integer(Sex == "Male")
  )
```

## 8.1 Modelling Objective

This section addresses the following research question:

> *To what extent do socioeconomic position (income, education),
> lifestyle behaviours (physical activity, smoking), and physical health
> burden predict the number of mentally unhealthy days reported in the
> past 30 days, and do the protective associations of physical activity
> differ across income and education strata?*

The objective is threefold:

1.  **Distributional justification** — to formally establish that the
    count distribution of mentally unhealthy days violates Poisson
    assumptions, motivating a zero-inflated negative binomial framework.

2.  **Explanatory** — to estimate adjusted associations between the
    predictor set and mental health burden, distinguishing between the
    process governing who reports zero mentally unhealthy days and the
    process governing how many days are reported among those with any
    burden.

3.  **Interaction analysis** — to test whether the protective
    association of physical activity with mental health burden differs
    across income strata and education strata, assessed as two separate
    moderation tests against the main-effects baseline.

MentHlth is treated as an integer count outcome bounded between 0 and
30. Ordinal predictors (Income, Education, Age) are converted to integer
scores preserving their monotonic structure, consistent with the ordinal
scoring approach adopted in Sections 6 and 7.

## 8.2 Outcome Variable and Model Justification

The MentHlth variable records the number of days in the past 30 days
that the respondent felt their mental health was not good. Its
distributional properties, established in the EDA (Section 3.2.1), have
direct implications for model selection.

### 8.2.1 Distributional Diagnostics

```{r}
zero_prop <- mean(diab$MentHlth == 0)

# Mean and variance — overdispersion check
mh_mean <- mean(diab$MentHlth)
mh_var  <- var(diab$MentHlth)
overdispersion_ratio <- mh_var / mh_mean

tibble(
  Statistic = c("Proportion zero", "Mean", "Variance", "Variance / Mean ratio"),
  Value = round(c(zero_prop, mh_mean, mh_var, overdispersion_ratio), 3)
)
```

The proportion of respondents reporting zero mentally unhealthy days
substantially exceeds what either a Poisson or standard negative
binomial distribution would predict from the observed mean alone. The
variance-to-mean ratio confirms severe overdispersion: under a Poisson
process this ratio should equal one, whereas the observed ratio of
approximately 17 indicates that the conditional variance far exceeds the
conditional mean. This rules out standard Poisson regression as an
appropriate modelling framework.

A negative binomial model addresses overdispersion by introducing a
dispersion parameter, but does not explicitly account for the excess
concentration of zeros. A **Zero-Inflated Negative Binomial (ZINB)**
model is therefore specified, which separates two distinct processes: a
logistic component modelling the probability that an individual belongs
to the structurally zero group (those who would never report mentally
unhealthy days regardless of predictor values), and a negative binomial
count component modelling the expected number of mentally unhealthy days
among those who do experience some burden.

## 8.3 Model Fitting and Justification

### 8.3.1 Poisson Regression

As a baseline, a Poisson regression model is fitted with MentHlth as the
outcome and the full predictor set. Under a Poisson process, the
conditional variance of the outcome is assumed to equal the conditional
mean. As established in Section 8.2.1, the observed variance-to-mean
ratio of approximately 17 already indicates this assumption is
untenable. The Poisson model is nevertheless fitted explicitly so that
its inadequacy can be formally demonstrated rather than assumed.

```{r}
mod_pois <- glm(
  MentHlth ~ Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex,
  data   = diab_zinb,
  family = poisson()
)

tidy(mod_pois, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value)
```

The coefficient table confirms the overdispersion problem directly.
Confidence intervals are implausibly narrow, for example, the interval
for Income spans only 0.897 to 0.899, and all predictors return p-values
of zero, indicating that standard errors are so severely deflated that
no meaningful inference can be drawn. These results reflect the Poisson
variance constraint rather than genuine precision in the data.

### 8.3.2 Negative Binomial Regression

The negative binomial model extends the Poisson by introducing a
dispersion parameter θ, allowing the conditional variance to exceed the
conditional mean. This directly addresses the overdispersion identified
above.

```{r}
mod_nb <- glm.nb(
  MentHlth ~ Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex,
  data = diab_zinb
)

tidy(mod_nb, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(term, estimate, conf.low, conf.high, p.value)
```

The negative binomial specification produces notably wider and more
credible confidence intervals throughout.

-   **Income gradient:** remains strong *(IRR = 0.890, 95% CI:
    0.885–0.896)*, indicating that each step up the income scale is
    associated with approximately an **11% reduction in the expected
    number of mentally unhealthy days**, consistent with the
    socioeconomic gradient identified in **Section 3.2**.

-   **Physical activity:** shows a protective association *(IRR = 0.784,
    95% CI: 0.764–0.805)*.

-   **Physical health burden:** is positively associated with mental
    health burden *(IRR = 1.061, 95% CI: 1.060–1.063)*.

-   **Education:** notably loses statistical significance under the
    negative binomial specification *(IRR = 0.989, 95% CI: 0.977–1.001,
    p = 0.082)*, suggesting that its apparent association in the Poisson
    model was an artefact of **overdispersion-inflated precision**
    rather than a genuine independent effect.

-   **Male sex:** is associated with lower expected burden *(IRR =
    0.670)*, consistent with the **sex gap observed in Section 3.2.4**.

However, the negative binomial model treats all zero observations as
arising from the same count process as non-zero values, implicitly
assuming that a respondent reporting zero mentally unhealthy days simply
falls at the low end of the distribution rather than potentially
representing a structurally distinct group.

### 8.3.3 Zero-Inflated Negative Binomial

The Zero-Inflated Negative Binomial model extends the negative binomial
by explicitly modelling two distinct data-generating processes. The
**count component** is a negative binomial regression estimating the
expected number of mentally unhealthy days among individuals who
experience some mental health burden. The **zero-inflation component**
is a logistic regression estimating the log-odds that an individual
belongs to the structural zero group, those who would report zero
mentally unhealthy days regardless of their predictor values. This
distinction is substantively meaningful: a respondent who reports zero
mentally unhealthy days may either be genuinely psychologically well (a
count-process zero) or may be someone who structurally never experiences
or reports mental health burden (a structural zero). The ZINB model
separates these two processes rather than collapsing them into a single
distribution.

The same predictor set is specified in both components. This reflects
the assumption that the factors shaping whether someone belongs to the
structural zero group are not categorically different from those shaping
the frequency of mentally unhealthy days, socioeconomic position,
physical health burden, and behavioural factors plausibly operate
through both processes.

```{r}
#| include: false
zinb_tidy <- function(model) {
  s <- summary(model)
  
  bind_rows(
    as_tibble(s$coefficients$count, rownames = "term") |>
      mutate(Component = "Count"),
    as_tibble(s$coefficients$zero, rownames = "term") |>
      mutate(Component = "Zero-inflation")
  ) |>
    rename(
      estimate  = Estimate,
      std.error = `Std. Error`,
      p.value   = `Pr(>|z|)`
    ) |>
    filter(term != "(Intercept)", term != "Log(theta)") |>
    mutate(
      conf.low  = exp(estimate - 1.96 * std.error),
      conf.high = exp(estimate + 1.96 * std.error),
      estimate  = exp(estimate),
      across(where(is.numeric), \(x) round(x, 3))
    ) |>
    select(Component, term, estimate, conf.low, conf.high, p.value) |>
    arrange(Component, term)
}
```

```{r}
mod_zinb_main <- zeroinfl(
  MentHlth ~ Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex |
    Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex,
  data = diab_zinb,
  dist = "negbin")

zinb_results <- zinb_tidy(mod_zinb_main)

zinb_results |> filter(Component == "Count") |> select(-Component)

zinb_results |> filter(Component == "Zero-inflation") |> select(-Component)
```

### 8.3.4 Vuong Test: Negative Binomial vs ZINB

The Vuong test compares the negative binomial and ZINB models on an
observation-by-observation basis, evaluating whether the zero-inflation
component provides a statistically significant improvement in fit. A
significant result favouring model 2 (ZINB) indicates that the excess
zeros in the outcome distribution cannot be fully accounted for by the
negative binomial dispersion parameter alone.

```{r}
vuong_nb_zinb <- vuong(mod_nb, mod_zinb_main)

```

Across all three corrections — **raw**, **AIC-corrected**, and
**BIC-corrected**, the **Vuong statistic** is large and negative *(z ≈
−88)*, with *p \< 2.22e−16* in every case.

-   The consistent designation of **model 2 \> model 1** confirms that
    the **ZINB model** provides a dramatically better fit than the
    **standard negative binomial model**.

-   The **magnitude of the statistic**, together with its stability
    across penalty corrections, leaves little ambiguity in the result.

-   These findings indicate that a **substantial proportion of the zero
    observations in MentHlth** represent a structurally distinct group,
    which is not adequately captured by the count process alone.

Consequently, the ZINB main effects model is adopted as the core
inferential specification for all subsequent analysis.

### 8.3.5 Interpretation of ZINB Main Effects

The ZINB model produces two sets of estimates that must be interpreted
separately. **Incidence rate ratios (IRRs)** from the count component
describe multiplicative changes in the expected number of mentally
unhealthy days among individuals who experience any mental health
burden. **Odds ratios (ORs)** from the zero-inflation component describe
the odds of belonging to the structural zero group, individuals who
would report zero mentally unhealthy days regardless of their predictor
values. A predictor may operate differently across the two components,
and these divergences are substantively informative.

```{r}
#| fig-height: 7
#| fig-width: 14

p_count <- zinb_results |>
  filter(Component == "Count") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                width = 0.25, orientation = "y") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    title    = "Count component: Incidence Rate Ratios",
    subtitle = "Expected mentally unhealthy days | any burden",
    x        = "IRR (95% CI)",
    y        = NULL
  )

p_zero <- zinb_results |>
  filter(Component == "Zero-inflation") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                width = 0.25, orientation = "y") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    title    = "Zero-inflation component: Odds Ratios",
    subtitle = "Odds of structural zero (never reports burden)",
    x        = "OR (95% CI)",
    y        = NULL
  )

p_count | p_zero
```

### Socioeconomic Gradients

**Income** shows a consistent protective association across both model
components. In the **count component**, each step up the income scale
reduces expected mentally unhealthy days by **6.3%** *(IRR = 0.937, 95%
CI: 0.933–0.941)*. In the **zero-inflation component**, higher income
also increases the odds of belonging to the **structural zero group**
*(OR = 1.099, 95% CI: 1.093–1.105)*. Together, these results indicate
that socioeconomic advantage both **reduces the frequency of mental
health burden and increases the likelihood of being structurally
insulated from it**.

**Education** shows a similar but weaker pattern. Higher education
reduces expected mentally unhealthy days *(IRR = 0.942, 95% CI:
0.933–0.951)* but is associated with **lower odds of structural zero
status** *(OR = 0.905, 95% CI: 0.894–0.915)*, suggesting that more
educated individuals may be **more likely to recognise and report burden
when it occurs**, despite experiencing fewer unhealthy days overall.

### Behavioural Factors

**Physical activity** shows the strongest protective behavioural
association, reducing expected mentally unhealthy days by **17.6%**
*(IRR = 0.824, 95% CI: 0.809–0.840)*. Its effect on structural zero
status is **small** *(OR = 1.045, 95% CI: 1.020–1.070)*, indicating that
activity primarily reduces the **frequency of burden rather than
eliminating it entirely**.

**Smoking** shows the opposite pattern, increasing expected mentally
unhealthy days by **19.9%** *(IRR = 1.199, 95% CI: 1.179–1.219)* and
reducing the odds of structural zero status *(OR = 0.811, 95% CI:
0.794–0.828)*, indicating greater mental health burden through both
processes.

### Physical Health Burden

Physical health burden strongly predicts mental health burden. Each
additional physically unhealthy day increases mentally unhealthy days by
**2.8%** *(IRR = 1.028, 95% CI: 1.028–1.029)* and reduces the
probability of structural zero status *(OR = 0.948, 95% CI:
0.947–0.949)*, reinforcing the close co-occurrence of physical and
mental health burden.

### Sex Differences

Sex shows a divergence between components. In the **count model**, male
sex is **small and non-significant** *(IRR = 0.987, 95% CI:
0.970–1.004)*. However, men have **much higher odds of structural zero
status** *(OR = 1.830, 95% CI: 1.792–1.869)*, suggesting the observed
sex gap in mentally unhealthy days may partly reflect **differences in
reporting behaviour rather than underlying burden**.

## 8.4 Moderation Analysis: Physical Activity Interactions

### 8.4.1 Rationale

The main effects model established that physical activity is associated
with a substantial reduction in mentally unhealthy days among those with
any burden (IRR = 0.824). However, the EDA demonstrated that both
physical activity participation and mental health burden are
independently structured by socioeconomic position, lower income and
education groups report both higher mental health burden and lower
physical activity rates. This raises the question of whether the
protective association of physical activity is uniform across
socioeconomic strata, or whether it is stronger or weaker at particular
levels of income or educational attainment.

Two moderation hypotheses are tested separately. Model 4 adds a
PhysActivity × Income interaction term to both components of the ZINB,
and Model 5 adds a PhysActivity × Education interaction. Separate models
are preferred over a single combined interaction model given the
moderate correlation between income and education (Spearman $ρ ≈ 0.45$),
which would partially confound the two interaction terms if estimated
simultaneously. Each model is compared against the main effects baseline
via likelihood ratio test.

### 8.4.2 Model Fitting

```{r}
mod_zinb_inc <- zeroinfl(
  MentHlth ~ Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex +
    PhysActivity:Income |
    Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex +
    PhysActivity:Income,
  data = diab_zinb,
  dist = "negbin"
)

mod_zinb_edu <- zeroinfl(
  MentHlth ~ Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex +
    PhysActivity:Education |
    Income + Education + PhysActivity +
    Smoker + PhysHlth + Age + Sex +
    PhysActivity:Education,
  data = diab_zinb,
  dist = "negbin"
)
```

### 8.4.3 Likelihood Ratio Tests

```{r}
# LRT: ZINB main vs ZINB + income interaction
lrt_inc <- lrtest(mod_zinb_main, mod_zinb_inc)

# LRT: ZINB main vs ZINB + education interaction
lrt_edu <- lrtest(mod_zinb_main, mod_zinb_edu)

bind_rows(
  tibble(
    Interaction = "PhysActivity × Income",
    df    = lrt_inc$Df[2],
    Chisq = round(lrt_inc$Chisq[2], 2),
    p.value = "< 2.2e-16"
  ),
  tibble(
    Interaction = "PhysActivity × Education",
    df    = lrt_edu$Df[2],
    Chisq = round(lrt_edu$Chisq[2], 2),
    p.value = "< 2.2e-16"
  )
)
```

Both interaction models provide statistically significant improvements
in fit relative to the **main effects baseline**. The **PhysActivity ×
Income** interaction produces a likelihood ratio statistic of
$\Delta\chi^2 = 110.37$ $(df = 2,\ p < 2.2\times10^{-16})$, and the
**PhysActivity × Education** interaction yields $\Delta\chi^2 = 87.33$
$(df = 2,\ p < 2.2\times10^{-16})$.

The **two degrees of freedom** in each case arise because the
interaction term appears in **both the count and zero-inflation
components** of the ZINB model.

The significance of both tests confirms that the **protective
association of physical activity with mental health burden is not
uniform across socioeconomic strata**. Instead, its magnitude differs
meaningfully across both **income and education levels**.

### 8.4.4 Interaction Plots and Interpretation

```{r}
pred_inc <- ggpredict(
  mod_zinb_inc,
  terms = c("Income [1:8]", "PhysActivity [0,1]")
)

# PhysActivity × Education predicted marginal means
pred_edu <- ggpredict(
  mod_zinb_edu,
  terms = c("Education [1:6]", "PhysActivity [0,1]")
)
```

```{r}
#| fig-height: 8
#| fig-width: 18

p_int_inc <- pred_inc |>
  as_tibble() |>
  mutate(Activity = if_else(group == "1", "Active", "Inactive")) |>
  ggplot(aes(x = x, y = predicted,
             colour = Activity, fill = Activity)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(
    breaks = 1:8,
    labels = c("< $10k", "$10k–$14k", "$15k–$19k", "$20k–$24k",
               "$25k–$34k", "$35k–$49k", "$50k–$74k", "$75k+")
  ) +
  scale_colour_manual(values = c("Active" = "#1a9641", "Inactive" = "#d7191c")) +
  scale_fill_manual(values   = c("Active" = "#1a9641", "Inactive" = "#d7191c")) +
  labs(
    title   = "Predicted mentally unhealthy days by income and physical activity",
    x       = "Household income",
    y       = "Predicted mentally unhealthy days (past 30)",
    colour  = "Physical activity",
    fill    = "Physical activity"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_int_edu <- pred_edu |>
  as_tibble() |>
  mutate(Activity = if_else(group == "1", "Active", "Inactive")) |>
  ggplot(aes(x = x, y = predicted,
             colour = Activity, fill = Activity)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(
    breaks = 1:6,
    labels = c("Kindergarten\nor less", "Grades 1–8",
               "Grades 9–11", "High school\ngraduate",
               "Some college /\ntechnical", "College\ngraduate")
  ) +
  scale_colour_manual(values = c("Active" = "#1a9641", "Inactive" = "#d7191c")) +
  scale_fill_manual(values   = c("Active" = "#1a9641", "Inactive" = "#d7191c")) +
  labs(
    title   = "Predicted mentally unhealthy days by education and physical activity",
    x       = "Education level",
    y       = "Predicted mentally unhealthy days (past 30)",
    colour  = "Physical activity",
    fill    = "Physical activity"
  ) 

(p_int_inc | p_int_edu) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**Physical activity and income.** Predicted mentally unhealthy days
decline across the income distribution for both groups, but far more
steeply among those who are physically active. Among inactive
individuals, predicted burden falls modestly from **9.33** at the lowest
income level to **7.29** at the highest. Among active individuals, the
decline is much larger, from **9.21** to **5.34**. At the lowest income
level the two groups are nearly identical, indicating little protective
association of activity. The divergence emerges progressively with
income, suggesting that the mental health benefits of physical activity
are **strongest among higher-income individuals**, consistent with a
resource constraint interpretation.

**Physical activity and education.** The education gradient differs
markedly. Among inactive respondents, predicted mentally unhealthy days
remain essentially **flat** across education levels (7.34 to 7.22),
indicating little protective effect of education alone. Among active
respondents, however, a clear decline is observed, from **8.37** at the
lowest education level to **5.39** among college graduates. At the
lowest education levels, active individuals initially show **higher
predicted burden than inactive individuals**, before the trajectories
reverse at higher education levels. This pattern suggests that physical
activity may not confer the same psychological benefits at low education
levels, potentially reflecting **occupational or manual labour rather
than voluntary leisure activity**, with the protective association
emerging more clearly among individuals with greater educational
resources.

## 8.5 Model Comparison

```{r}
model_comparison <- tibble(
  Model       = c("Poisson", "Negative Binomial",
                  "ZINB (main effects)",
                  "ZINB + PhysActivity × Income",
                  "ZINB + PhysActivity × Education"),
  AIC         = c(AIC(mod_pois), AIC(mod_nb),
                  AIC(mod_zinb_main), AIC(mod_zinb_inc), AIC(mod_zinb_edu)),
  BIC         = c(BIC(mod_pois), BIC(mod_nb),
                  BIC(mod_zinb_main), BIC(mod_zinb_inc), BIC(mod_zinb_edu)),
  LogLik      = c(logLik(mod_pois), logLik(mod_nb),
                  logLik(mod_zinb_main), logLik(mod_zinb_inc), logLik(mod_zinb_edu))
) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))

model_comparison |>
  select(Model, AIC, BIC, LogLik)
```

The **model comparison table** confirms that each successive
specification represents a genuine improvement in fit. The transition
from **Poisson to negative binomial** produces the largest reduction in
information criteria, reflecting the substantial overdispersion present
in the outcome distribution.

Moving to the **ZINB main effects model** yields a further improvement
in fit, formally confirmed by the **Vuong test**
$(z \approx -88,\ p < 2.2\times10^{-16})$. This indicates that a
meaningful proportion of zero observations arise from a **structurally
distinct process**, rather than representing the lower tail of the count
distribution.

Both **interaction models** further improve on the main effects
baseline. The **PhysActivity × Income** interaction produces the larger
improvement $(\Delta\chi^2 = 110.37)$ compared with the **PhysActivity ×
Education** interaction $(\Delta\chi^2 = 87.33)$, although both
likelihood ratio tests are **highly significant**, confirming that the
protective effect of physical activity varies across socioeconomic
strata.

## 8.6 Discussion and Limitations

This section addressed the third research question by examining the
socioeconomic, behavioural, and physical health determinants of mental
health burden using a zero-inflated negative binomial (ZINB) framework.
The model separates two processes: the probability of structural zero
status and the expected number of mentally unhealthy days among those
with any burden, avoiding the loss of heterogeneity that would occur if
both were collapsed into a single distribution.

The main effects model identified a strong income gradient across both
components. Higher income was associated with fewer mentally unhealthy
days in the count component and greater odds of structural zero status
in the zero-inflation component, indicating that socioeconomic advantage
operates through two mechanisms consistent with Dahlgren and Whitehead's
(1991) social determinants framework. Physical activity emerged as the
strongest behavioural predictor in the count component, aligning with
evidence linking leisure-time activity with improved psychological
wellbeing *(Liu et al., 2023)*.

The moderation analyses showed that the protective association of
physical activity varies across socioeconomic strata. The PhysActivity ×
Income interaction indicates that the protective effect is minimal at
the lowest income levels and increases with affluence, consistent with a
resource constraint interpretation, where the benefits of physical
activity depend on supportive material conditions. The PhysActivity ×
Education interaction shows a crossover at the lowest education levels,
where active individuals report higher predicted burden than inactive
ones. This pattern suggests that physical activity at low educational
attainment may reflect occupational exertion rather than voluntary
leisure activity, which has a different psychological profile *(Liu et
al., 2023)*.

The sex effect diverges across model components. Male sex is
non-significant in the count component but strongly associated with
structural zero status \$(\\mathrm{OR}=1.830)\$ in the zero-inflation
component. This pattern suggests that the observed sex gap in reported
mental health burden may partly reflect differences in reporting
behaviour, rather than solely differences in underlying distress.

Several limitations apply:

-   **Self-reported data** may introduce measurement error and social
    desirability bias.

-   The **cross-sectional design** prevents causal inference and leaves
    open the possibility of reverse causality.

-   The **binary physical activity measure** does not distinguish
    leisure from occupational activity, limiting interpretation of the
    education interaction.

-   **Ordinal predictors** are treated as integer scores, implicitly
    assuming equal interval spacing between categories.

## 8.7 Conclusion

The results indicate that **mental health burden is jointly shaped by
socioeconomic position, physical health, and behavioural factors**, with
the **income gradient emerging as the dominant determinant** operating
across both model components. Physical activity appears to be a
meaningful protective factor, but its benefits are conditional on
socioeconomic context, most consistently observed among higher-income
and more educated individuals, and potentially reversed when activity
reflects occupational necessity rather than voluntary health behaviour.

Overall, the findings highlight the limitations of purely behavioural
explanations for mental health inequalities and emphasise the importance
of material and structural conditions in shaping both the experience and
reporting of mental health burden.

# 9. Discussion and Conclusion

## 9.1 Overview

This report investigated health determinants and outcomes across three
research questions using a large nationally representative health survey
dataset. Each research question was motivated by the exploratory data
analysis in Section 3 and addressed using a methodologically distinct
analytical framework: multinomial logistic regression and ensemble
machine learning for diabetes prediction (RQ1); regularised logistic
regression and XGBoost with SHAP-based explainability for cardiovascular
risk moderation (RQ2); and zero-inflated negative binomial regression
for mental health burden (RQ3). Taken together, the three analyses
converge on a consistent set of substantive findings that are most
clearly visible when examined across, rather than within, individual
sections.

## 9.2 Cross-Cutting Findings

### 9.2.1 The primacy of cardiometabolic pathways

Across all three research questions, biological and physiological
indicators consistently emerged as the dominant predictors of health
outcomes, outperforming behavioural and socioeconomic variables in
direct predictive terms. In RQ1, high blood pressure and high
cholesterol were the strongest predictors of diabetes status by both
SHAP magnitude and odds ratio, with **HighBP** alone associated with a
139% increase in the odds of diabetes relative to no diabetes. In RQ2,
the same cardiometabolic indicators, alongside age and general health
status, drove cardiovascular risk prediction, with socioeconomic
variables contributing only modest independent effects. In RQ3, physical
health burden emerged as the strongest covariate in the count component
of the ZINB model, reflecting the tight coupling between physical and
mental health burden identified in the EDA.

This pattern is consistent with a pathophysiological perspective in
which metabolic dysregulation and chronic physical illness represent
upstream determinants that mediate much of the apparent influence of
behavioural and socioeconomic factors on health outcomes. It also
carries a methodological implication: models that omit cardiometabolic
indicators risk substantial confounding, with behavioural and
socioeconomic variables absorbing variance that properly belongs to
physiological pathways.

### 9.2.2 Socioeconomic gradients are outcome-specific

While socioeconomic position structured all three outcomes, its
magnitude and mechanism varied considerably across research questions.
In RQ1, income and education were present as predictors but ranked well
below cardiometabolic variables in both SHAP importance and odds ratio
magnitude. In RQ2, socioeconomic moderation of cardiovascular risk was
statistically significant; income and education interactions both
reached *p* \< 0.01. However, the practical magnitude of these
interactions was small, with SHAP interaction values remaining
negligible relative to the direct effects of clinical predictors.

In RQ3, by contrast, income emerged as the dominant predictor of mental
health burden, operating simultaneously through the count and
zero-inflation components and moderating the protective association of
physical activity across the full income distribution. This variation
suggests that socioeconomic gradients in health are not uniform across
outcomes and that their apparent importance is sensitive to the outcome
under consideration. Conditions with strong biological determinants,
such as diabetes and cardiovascular disease, may be less amenable to
socioeconomic explanation than outcomes with greater psychosocial
mediation, such as self-reported mental health burden.

### 9.2.3 Behavioural explanations require structural context

A consistent finding across all three research questions was that
behavioural variables, including physical activity, smoking, and diet,
operate within structural constraints and cannot be interpreted
independently of socioeconomic context. In RQ1, physical activity showed
essentially no independent association with diabetes status once
cardiometabolic variables were included in the model, suggesting that
its apparent protective effect in unadjusted analyses is largely
mediated through metabolic pathways.

In RQ2, smoking contributed to cardiovascular risk prediction but ranked
consistently below clinical and demographic indicators. In RQ3, physical
activity was the strongest behavioural predictor in the count component;
however, the moderation analyses revealed that its protective
association was negligible at the lowest income levels and subject to
reversal at the lowest education levels, where it may reflect
occupational exertion rather than voluntary leisure activity.

Taken together, these findings are consistent with **Dahlgren and
Whitehead's (1991)** social determinants framework, in which individual
behaviours are embedded within, and shaped by, broader material and
structural conditions. Interventions targeting behavioural change in
isolation, without addressing the socioeconomic contexts that constrain
and enable those behaviours, are unlikely to achieve uniform benefit
across the population.

## 9.3 Methodological Reflections

The three research questions employed genuinely distinct method
families, each contributing insights that would not have been
recoverable from the others. The multinomial logistic regression in RQ1
was the only model capable of retaining predictive capacity for the
prediabetes class, which tree-based models collapsed entirely due to
class overlap and the absence of ordinal structure in standard splitting
criteria. This finding illustrates that predictive performance metrics
aggregated across classes can obscure clinically important failures in
minority class prediction.

The XGBoost framework in RQ2 demonstrated that flexible machine learning
models can substantially outperform logistic regression for imbalanced
binary outcomes. However, SHAP-based interaction analysis revealed
predominantly additive structure even in a dataset where interactions
were theoretically expected. This finding cautions against assuming that
model flexibility necessarily implies empirical interaction richness.

The ZINB framework in RQ3 demonstrated the value of distributional
modelling for outcomes with excess zeros. A single-component model would
conflate two substantively distinct processes and produce misleading
coefficient estimates for predictors that operate differently across
those processes.

## 9.4 Broader Limitations

Several limitations apply across all three research questions and should
be considered when interpreting the findings collectively.

The dataset is **cross-sectional**, meaning that causal relationships
cannot be established from the observed associations. Reverse causality
is a plausible alternative explanation for several findings; poor health
may reduce income and physical activity participation just as readily as
low income and inactivity may worsen health outcomes.

The dataset relies entirely on **self-reported measures**, introducing
the possibility of measurement error, social desirability bias, and
differential reporting across demographic groups. This concern is
particularly relevant for mental health outcomes, where the sex
divergence in the ZINB zero-inflation component suggests that reporting
behaviour itself may vary systematically.

The socioeconomic indicators available, namely income band and education
level, are relatively **coarse proxies** for the multidimensional
construct of socioeconomic position. Their categorical nature limits the
precision with which gradients can be estimated.

Finally, the dataset does not include **clinical biomarkers,
longitudinal measures of socioeconomic trajectory, or healthcare
utilisation patterns**, all of which would be necessary to adjudicate
between competing causal explanations for the observed associations.

## 9.5 Conclusion

This report demonstrated that health outcomes in this population are
jointly shaped by cardiometabolic status, socioeconomic position, and
behavioural factors. However, the relative importance of these domains
varies substantially across outcomes and cannot be reduced to a single
explanatory framework.

Physiological indicators dominate prediction of diabetes and
cardiovascular disease, while socioeconomic gradients are most
consequential for mental health burden. Behavioural variables, on their
own, are consistently insufficient explanations and must be interpreted
within the structural contexts that enable or constrain them.

Methodologically, the three analyses illustrate that no single modelling
framework is universally appropriate. The choice between interpretable
parametric models, flexible machine learning approaches, and
distributional count models should be guided by the inferential
objectives, outcome properties, and class structure of the problem.

Together, these findings contribute to a broader evidence base linking
social determinants, cardiometabolic risk, and psychological wellbeing,
and underscore the value of multi-method analytical approaches for
characterising health inequalities in large-scale survey data.
